---
title: "Topic 4: Data Visualization and a Grammar of Graphics"
format: 
  live-html:
    include-in-header:
      text: |
        <style>
        .btn-exercise-solution { display: none !important; }
        </style>
    webr:
      packages:
        - dplyr
        - gradethis
        - skimr
        - ggplot2
        - ggmosaic
        - jsonlite
        - base64enc
      cell-options: 
        persist: true
    live:
      show-solutions: true
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}
{{< include ./_extensions/r-wasm/live/_gradethis.qmd >}}

```{webr}
#| edit: false
#| output: false
#| define:
#|   - ok_response

library(htmltools)
ok_response <- function(response, n) {
  if (is.na(response)) div(HTML("You haven't answered yet."), style = "color: purple")
  else if (response == n) div(HTML("Correct ✓"), style = "color: green")
  else div(HTML("Incorrect ✗"), style = "color: red")
}
```

```{ojs}
//| echo: false

function ok_checkbox(response, n) {
  if (!response || response.length === 0) 
    return html`<span style="color:purple">You haven't answered yet.</span>`;
  if (response.toString() === n) 
    return html`<span style="color:green">Correct ✓</span>`;
  return html`<span style="color:red">Not Yet! ✗</span>`;
}
```

```{webr}
#| edit: false
#| output: false
#| define:
#|   - normalize_checkbox_response

normalize_checkbox_response <- function(x) {
  return(cat(sort(x), sep = ","))
}
```

```{webr}
#| context: setup
#| echo: false
#| output: false

library(ggplot2)
library(stringr)
library(skimr)
library(ggmosaic)
library(jsonlite)
library(base64enc)

exercise_results <- list(
  notebook             = "Topic 4: Data Visualization",
  print_mpg            = "unattempted",
  plot_displ_cty       = "unattempted",
  plot_hwy_cyl         = "unattempted",
  boxplot_hwy_cyl      = "unattempted",
  hwy_displ_class      = "unattempted",
  facet_wrapped        = "unattempted",
  barplot_class_drv    = "unattempted",
  trans2_fill_class    = "unattempted",
  trans2_drv_plot      = "unattempted"
)
```

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(stringr)
library(gradethis)
library(skimr)
library(ggmosaic)
library(kableExtra)

data <- data.frame(
  "FuelType"    = c("F1", "F1", "F1", "F2", "F2", "F2"),
  "EngineBrand" = c("E1", "E2", "E3", "E1", "E2", "E3")
)
```

:::{.callout-note}
## About

This activity provides an introduction to the grammar of graphics, plotting strategies, and using ggplot. The discussion you'll work through loosely follows Chapter 3 of Wickham and Grolemund's *R for Data Science*. Here's [a link to the updated second edition of the text](https://r4ds.hadley.nz/){target="_blank"}.

```{ojs}
//| echo: false

{
  const btn = html`<button style="
    background: #436b95;
    color: white;
    border: none;
    padding: 8px 16px;
    border-radius: 4px;
    cursor: pointer;
    font-size: 0.9em;
  ">🔄 Reset My Responses</button>`;

  btn.onclick = () => {
    if (confirm('Reset all responses for this activity? This cannot be undone.')) {
      Object.keys(localStorage)
        .filter(k => k.endsWith('_selected_t4') || k.endsWith('_correct_t4') || k.endsWith('_result_t4'))
        .forEach(k => localStorage.removeItem(k));
      location.reload();
    }
  };

  return btn;
}
```


*Note.* The button above resets multiple choice and checkbox questions. Currently, resetting code cells must be done manually via hitting the *Start Over* button on each individual interactive cell.

:::

## Data Visualization

Through this workbook you'll learn techniques for data visualization. You'll think about how the type of plot chosen impacts what information is conveyed (or lost). Choosing the wrong plot type can result in a useless plot (at best) or even a plot which is misleading. Think about the *types* (numerical, categorical) of variables you are working with and how they dictate which plot type should be utilized -- for example, a scatterplot is not appropriate in every scenario!

## Objectives

**Workbook Objectives:** After completing this workbook you should be able to:

+ Read a plot, describe any interesting features, and accurately convey the *story* being told by the plot.
+ Identify and utilize the most appropriate plot types for your scenario by considering the *types* of variables you are working with.
+ Utilize color, size, and shape to show additional dimensions of data in a plot, but also recognize that adding more information to a plot negatively impacts our ability to read the graphic. 
+ Use your new found knowledge of data visualization to generate informative plots which uncover a compelling (and truthful) story about the relationships that exist between variables.

Note, much of this notebook is inspired by [Chapter 3 of R for Data Science](https://r4ds.had.co.nz/data-visualisation.html){target="_blank"} by Hadley Wickham and Garrett Grolemund. This is a great freely-available resource. Please check it out.

## The Grammar of Graphics

**Grammar of Graphics:** Jeffrey Gitomer said "Your grammar is a reflection of your image". Here, we take the converse literally. Your image is a reflection of your grammar. Just like a well-written sentence follows the rules of grammar, so does an informative statistical graphic.

* Graphics need a subject (an underlying dataset)
* Graphics need verbs (a type of plotting structure, called a `geom`)
* Graphics need adjectives (attributes attached to the visualized data, called `aesthetics`)
* Graphics need context (a title, scales, axes, legends, etc.)

In this workbook we explore `{ggplot2}` and think about graphics in terms of their layered grammar. 

**A Note on Plots:** Choosing an appropriate plot is extremely important in data visualization. Some plots don't make sense for certain variable types. For example, a scatterplot in the case of two-categorical predictors is quite a silly choice -- the only thing this plot tells us is that every combination of fuel type and engine brand exists in our dataset. We have no idea which combinations are most or least popular. 

```{r}
#| echo: false
#| eval: true
#| fig.height: 2.25
#| fig.width: 2.25
#| fig-alt: "A scatterplot of Engine Brand (E1, E2, E3) against Fuel Type (F1, F2), with one point at every combination of the two categories. Because both variables are categorical, the plot only shows that each combination exists -- it conveys no information about how common each combination is, making it an ineffective way to visualize this data."

data |> 
  ggplot() + 
  geom_point(aes(x = FuelType, y = EngineBrand), size = 2) + 
  labs(y = "Engine Brand", x = "Fuel Type")
```

The following are some recommended basic plot types under certain scenarios:

+ A single numerical variable: a histogram, boxplot, or density plot
+ A single categorical variable: a bar plot
+ Numerical versus numerical: a scatterplot or heatmap
+ Numerical versus categorical: side-by-side boxplots, stacked or overlaid densities, stacked or overlaid histograms
+ Categorical versus categorical: bar plot with fill color, mosaic plot, heatmap

:::{.callout-caution}
## Check Your Understanding: A Better Plot

Which of the following plot types would have been a better choice to display the engine brand and fuel type data? Select all that apply.

```{ojs}
//| echo: false

mutable ok_response = (response, n) => { return html`Loading...` };
viewof q1 = Inputs.checkbox(
  new Map([
    ["A histogram", 1],
    ["A barplot with fill color", 2],
    ["A heatmap", 3],
    ["A dotplot", 4],
    ["Side-by-side boxplots", 5]
  ]),
  {value: JSON.parse(localStorage.getItem("q1_selected_t4") ?? "[]") ?? []}
);

{
  localStorage.setItem("q1_selected_t4", JSON.stringify(q1));
  localStorage.setItem("q1_correct_t4", "2,3");
  localStorage.setItem("q1_result_t4", (!q1 || q1.length === 0) ? "unattempted" : (q1.toString() === "2,3" ? "correct" : "incorrect"));
}
ok_checkbox(q1, "2,3");
```

:::

## Getting Started: Exploring Data

Throughout this workbook, we will explore fuel efficiencies of multiple classes of vehicles using the `mpg` data frame.

1. Remember that a *data frame* is like an Excel spreadsheet. Explore the `mpg` data frame by typing `mpg` in the following code block and running it.

```{webr}
#| exercise: print-mpg-dataframe
______
```

```{webr}
#| exercise: print-mpg-dataframe
#| solution: true

mpg
```

```{webr}
#| exercise: print-mpg-dataframe
#| check: true

exercise_results$print_mpg <<- "fail"
gradethis::grade_this({
  feedback <- gradethis::code_feedback()
  if (is.null(feedback)) {
    exercise_results$print_mpg <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

Notice that when you execute a line of code which just calls the name of a data frame, a snippet of that data frame is printed out. This is true for other objects (variables, vectors, functions, etc) in R as well.

**Some Basic Exploratory Functions:** It is useful to know more about your dataset when you first start working with data. R has a few exploratory functions which you should know about: `mpg |> names()` prints out a list of names of your data frame's columns, `mpg |> head()` will show you the first six rows of the `mpg` dataset, `mpg |> dim()` will show you the number of rows and columns in the dataset, and `mpg |> glimpse()` will show you information about how R is treating the columns (`int` and `num` denote numerical values while `chr` and `fct` represent character strings and categorical variables respectively). You might also try `mpg |> skim()`, where the `skim()` function comes from the `{skimr}` package. The `skim()` function provides lots of high-level information about your dataset from just a single function call. Use the code block below to try each of these functions and use the output to answer the following questions:

```{webr}
#| exercise: explore-mpg-dataframe
#| caption: Ungraded, Exploration Only
______
```

:::{.hint exercise="explore-mpg-dataframe"}
:::{.callout-note collapse="false"}
## Hint 1

Try some of the functions listed above — `names()`, `head()`, `dim()`, `glimpse()`, and `skim()`.

:::
:::

:::{.hint exercise="explore-mpg-dataframe"}
:::{.callout-note collapse="false"}
## Hint 2

Fill in the blanks with the name of the data frame.

```{r}
#| echo: true
#| eval: false

___ |> names()
___ |> head()
___ |> dim()
___ |> glimpse()
___ |> skim()
```

:::
:::

:::{.hint exercise="explore-mpg-dataframe"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

```{r}
#| echo: true
#| eval: false

mpg |> names()
mpg |> head()
mpg |> dim()
mpg |> glimpse()
mpg |> skim()
```

:::
:::

:::{.callout-caution}
## Check Your Understanding: Variables

How many variables are contained in the `mpg` dataset?

```{ojs}
//| echo: false

viewof q2 = Inputs.radio(
  new Map([
    ["1", 1],
    ["11", 2],
    ["234", 3],
    ["245", 4],
    ["2574", 5]
  ]),
  {value: JSON.parse(localStorage.getItem("q2_selected_t4") ?? "null")}
);

{
  localStorage.setItem("q2_selected_t4", JSON.stringify(q2));
  localStorage.setItem("q2_correct_t4", "2");
  localStorage.setItem("q2_result_t4", q2 === null ? "unattempted" : (q2 == 2 ? "correct" : "incorrect"));
}
ok_response(q2, "2");
```

:::

:::{.callout-caution}
## Check Your Understanding: Observations

How many observations (records) are contained in the `mpg` dataset?

```{ojs}
//| echo: false

viewof q3 = Inputs.radio(
  new Map([
    ["1", 1],
    ["11", 2],
    ["234", 3],
    ["245", 4],
    ["2574", 5]
  ]),
  {value: JSON.parse(localStorage.getItem("q3_selected_t4") ?? "null")}
);

{
  localStorage.setItem("q3_selected_t4", JSON.stringify(q3));
  localStorage.setItem("q3_correct_t4", "3");
  localStorage.setItem("q3_result_t4", q3 === null ? "unattempted" : (q3 == 3 ? "correct" : "incorrect"));
}
ok_response(q3, "3");
```

:::

:::{.callout-caution}
## Check Your Understanding: Numerical Variables

Which of the variables is R interpreting as numerical? Select all that apply.

```{ojs}
//| echo: false

viewof q4 = Inputs.checkbox(
  new Map([
    ["manufacturer", 1],
    ["model", 2],
    ["displ (displacement)", 3],
    ["cyl (cylinders)", 4],
    ["trans (transmission type)", 5],
    ["year", 6],
    ["drv (drive type)", 7],
    ["cty (city fuel economy)", 8],
    ["hwy (highway fuel economy)", 9],
    ["fl (fuel type)", 10],
    ["class (vehicle class)", 11]
  ]),
  {value: JSON.parse(localStorage.getItem("q4_selected_t4") ?? "[]") ?? []}
);

{
  localStorage.setItem("q4_selected_t4", JSON.stringify(q4));
  localStorage.setItem("q4_correct_t4", "3,4,6,8,9");
  localStorage.setItem("q4_result_t4", (!q4 || q4.length === 0) ? "unattempted" : (q4.toString() === "3,4,6,8,9" ? "correct" : "incorrect"));
}
ok_checkbox(q4, "3,4,6,8,9");
```

:::

The `diamonds` data frame is also available to you. See if you can get your basic exploratory functions to help you answer the same questions about the `diamonds` dataset using the code block below.

```{webr}
#| exercise: explore-diamonds
#| caption: Ungraded, Exploration Only
______
```

:::{.hint exercise="explore-diamonds"}
:::{.callout-note collapse="false"}
## Hint 1

Use some of the same functions you used to explore the `mpg` data frame in the previous code block — just swap out the name of the data frame.

:::
:::

Were you able to obtain the information you were looking for? Be sure to ask a question if not!

## Building and Interpreting Plots

The code below creates a plot of highway miles per gallon (`hwy`) against engine displacement `displ` (a measure of the size of an engine).

```{r}
#| echo: true
#| eval: true
#| message: false
#| warning: false
#| results: false
#| fig-alt: "Scatterplot of highway fuel economy (hwy) against engine displacement (displ). The relationship is negative -- as engine displacement increases, highway mpg tends to decrease, though the points are fairly spread out rather than tightly clustered. A few cars with small engines (around 2 liters) stand out with unusually high highway mpg, above 40, well above the surrounding points. One car near 4.7 liters has an unusually low highway mpg, around 12, well below the general trend at that displacement. Several vehicles with large engines (between 5 and 7 liters) have miles per gallon ratings far above the other vehicles in this range."

mpg |> 
  ggplot() + 
  geom_point(aes(x = displ, y = hwy))
```

Use the plot above to answer the following questions.

:::{.callout-caution}
## Check Your Understanding: Reading a Plot I

Which of the following is true?

```{ojs}
//| echo: false

viewof q5 = Inputs.radio(
  new Map([
    ["Highway miles per gallon and engine displacement are independent.", 1],
    ["Highway miles per gallon and engine displacement are associated.", 2],
    ["It is impossible to tell whether the variables are associated or independent.", 3],
    ["Highway miles per gallon and engine displacement are both independent and associated.", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q5_selected_t4") ?? "null")}
);

{
  localStorage.setItem("q5_selected_t4", JSON.stringify(q5));
  localStorage.setItem("q5_correct_t4", "2");
  localStorage.setItem("q5_result_t4", q5 === null ? "unattempted" : (q5 == 2 ? "correct" : "incorrect"));
}
ok_response(q5, "2");
```

:::

:::{.callout-caution}
## Check Your Understanding: Reading a Plot II

Describe the relationship between highway miles per gallon and engine displacement.

```{ojs}
//| echo: false

viewof q6 = Inputs.radio(
  new Map([
    ["Engine displacement and highway miles per gallon share a negative, linear relationship.", 1],
    ["Engine displacement and highway miles per gallon share a positive, linear relationship.", 2],
    ["Engine displacement and highway miles per gallon share a negative, non-linear relationship.", 3],
    ["Engine displacement and highway miles per gallon share a positive, non-linear relationship.", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q6_selected_t4") ?? "null")}
);

{
  localStorage.setItem("q6_selected_t4", JSON.stringify(q6));
  localStorage.setItem("q6_correct_t4", "3");
  localStorage.setItem("q6_result_t4", q6 === null ? "unattempted" : (q6 == 3 ? "correct" : "incorrect"));
}
ok_response(q6, "3");
```

:::

:::{.callout-caution}
## Check Your Understanding: Reading a Plot III

Which is the best practical interpretation of the plot?

```{ojs}
//| echo: false

viewof q7 = Inputs.radio(
  new Map([
    ["Generally, engines with higher displacement are associated with lower average highway miles per gallon ratings.", 1],
    ["Generally, engines with higher displacement are associated with higher average highway miles per gallon ratings.", 2],
    ["Higher engine displacement causes lower average highway miles per gallon ratings.", 3],
    ["Engine displacement causes higher average highway miles per gallon ratings.", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q7_selected_t4") ?? "null")}
);

{
  localStorage.setItem("q7_selected_t4", JSON.stringify(q7));
  localStorage.setItem("q7_correct_t4", "1");
  localStorage.setItem("q7_result_t4", q7 === null ? "unattempted" : (q7 == 1 ? "correct" : "incorrect"));
}
ok_response(q7, "1");
```

:::

**A Note on Plotting Structure:** Using `ggplot()` to create plots in R is daunting at first, but with practice you will notice the structure is very consistent and convenient!

+ Our plot consists of at least two pieces (subject and verb): `mpg |> ggplot()` tells `ggplot` that the subject of our plot is the data contained in the `mpg` table, while the `geom_point()` verb tells `ggplot` that we want our data displayed as points (a scatterplot). The `aes()` inside of `geom_point()` tells `ggplot` some of the adjectives for each individual observation -- here, just the location!
+ In general a simple plot takes the form below -- (where the items in "ALL CAPS" are replaced by your data frame, desired geometry type, and aesthetic mappings).

```{r}
#| echo: true
#| eval: false

DATA |>
  ggplot() + 
  geom_TYPE(aes(MAPPINGS))
```

Use the code block below to make a scatterplot of average city miles per gallon (`cty`) explained by engine displacement (`displ`).

```{webr}
#| exercise: plot-displ-cty
______
```

:::{.hint exercise="plot-displ-cty"}
:::{.callout-note collapse="false"}
## Hint 1

Start by copying and pasting the code from the plot at the start of this subsection.

:::
:::

:::{.hint exercise="plot-displ-cty"}
:::{.callout-note collapse="false"}
## Hint 2

```{r}
#| echo: true
#| eval: false

mpg |> 
  ggplot() + 
  geom_point(aes(x = displ, y = hwy))
```

What needs to be changed here?

:::
:::

:::{.hint exercise="plot-displ-cty"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

Swap out the `hwy` variable for the `cty` variable.

```{r}
#| echo: true
#| eval: false

mpg |> 
  ggplot() + 
  geom_point(aes(x = displ, y = cty))
```

:::
:::

```{webr}
#| exercise: plot-displ-cty
#| solution: true

mpg |>
  ggplot() + 
  geom_point(aes(x = displ, y = cty))
```

```{webr}
#| exercise: plot-displ-cty
#| check: true

exercise_results$plot_displ_cty <<- "fail"
gradethis::grade_this({
  feedback <- gradethis::code_feedback()
  if (is.null(feedback)) {
    exercise_results$plot_displ_cty <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

Now make a scatterplot of average highway miles per gallon (`hwy`) explained by number of cylinders (`cyl`).

```{webr}
#| exercise: plot-hwy-cyl
______
```

:::{.hint exercise="plot-hwy-cyl"}
:::{.callout-note collapse="false"}
## Hint 1

Again, start by copying and pasting an earlier plot as a starting point.

:::
:::

:::{.hint exercise="plot-hwy-cyl"}
:::{.callout-note collapse="false"}
## Hint 2

What needs to be changed now?

```{r}
#| echo: true
#| eval: false

mpg |>
  ggplot() + 
  geom_point(aes(x = displ, y = cty))
```

:::
:::

:::{.hint exercise="plot-hwy-cyl"}
:::{.callout-note collapse="false"}
## Hint 3

Both of the aesthetic mappings need to be changed.

```{r}
#| echo: true
#| eval: false

mpg |>
  ggplot() + 
  geom_point(aes(x = ___, y = ___))
```

:::
:::

:::{.hint exercise="plot-hwy-cyl"}
:::{.callout-note collapse="false"}
## Hint 4 (Solved)

Put cylinders (`cyl`) on the x-axis and highway gas mileage (`hwy`) on the y-axis.

```{r}
#| echo: true
#| eval: false

mpg |>
  ggplot() + 
  geom_point(aes(x = cyl, y = hwy))
```

:::
:::

```{webr}
#| exercise: plot-hwy-cyl
#| solution: true

mpg |>
  ggplot() + 
  geom_point(aes(x = cyl, y = hwy))
```

```{webr}
#| exercise: plot-hwy-cyl
#| check: true

exercise_results$plot_hwy_cyl <<- "fail"
gradethis::grade_this({
  feedback <- gradethis::code_feedback()
  if (is.null(feedback)) {
    exercise_results$plot_hwy_cyl <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

Notice that this plot isn't very useful because the cylinders variable takes on very few levels. We may be better off if we treat `cyl` as if it were a categorical variable here. Check out some of the other geometry layers available to you in ggplot [here](https://ggplot2.tidyverse.org/reference/){target="_blank"}. Try building a set of side-by-side boxplots. If you get an error, be sure to read it -- R suspects that the plot produced wasn't the one you wanted and provides a suggestion.

```{webr}
#| exercise: boxplot-hwy-cyl
______
```

:::{.hint exercise="boxplot-hwy-cyl"}
:::{.callout-note collapse="false"}
## Hint 1

Continue copying and pasting an earlier plot as a starting point.

:::
:::

:::{.hint exercise="boxplot-hwy-cyl"}
:::{.callout-note collapse="false"}
## Hint 2

What needs to be changed here?

```{r}
#| echo: true
#| eval: false

mpg |>
  ggplot() + 
  geom_point(aes(x = cyl, y = hwy))
```

:::
:::

:::{.hint exercise="boxplot-hwy-cyl"}
:::{.callout-note collapse="false"}
## Hint 3

Swap `geom_point()` out with `geom_boxplot()` and generate the plot. Can you decode the warning message that appears?

:::
:::

:::{.hint exercise="boxplot-hwy-cyl"}
:::{.callout-note collapse="false"}
## Hint 4

R is considering both the x- and y-variables to be numeric. The warning message suggests adding a `group` parameter inside of the `aes()` function. Which variable should be mapped to `group`?

```{r}
#| echo: true
#| eval: false

mpg |>
  ggplot() + 
  geom_boxplot(aes(x = cyl, y = hwy, group = ___))
```

:::
:::

:::{.hint exercise="boxplot-hwy-cyl"}
:::{.callout-note collapse="false"}
## Hint 5 (Solved)

We want a boxplot for each cylinder "group." Map `cyl` to the `group` aesthetic.

```{r}
#| echo: true
#| eval: false

mpg |>
  ggplot() + 
  geom_boxplot(aes(
    x = cyl, 
    y = hwy,
    group = cyl))
```

:::
:::

```{webr}
#| exercise: boxplot-hwy-cyl
#| solution: true

mpg |>
  ggplot() + 
  geom_boxplot(aes(
    x = cyl, 
    y = hwy, 
    group = cyl))
```

```{webr}
#| exercise: boxplot-hwy-cyl
#| check: true

exercise_results$boxplot_hwy_cyl <<- "fail"
gradethis::grade_this({
  feedback <- gradethis::code_feedback()
  if (is.null(feedback)) {
    exercise_results$boxplot_hwy_cyl <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

Consider the variables for vehicle class (`class`) and drive type (`drv`) as you answer the following questions.

:::{.callout-caution}
## Check Your Understanding: Appropriate Plot Types I

Why would a scatterplot not be an appropriate visual to use with these two variables?

```{ojs}
//| echo: false

viewof q8 = Inputs.radio(
  new Map([
    ["Trick question -- a scatterplot is always appropriate.", 1],
    ["At least one of the two variables is not numerical.", 2],
    ["A scatterplot would be okay but there are better options.", 3],
    ["There isn't enough data for a scatterplot.", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q8_selected_t4") ?? "null")}
);

{
  localStorage.setItem("q8_selected_t4", JSON.stringify(q8));
  localStorage.setItem("q8_correct_t4", "2");
  localStorage.setItem("q8_result_t4", q8 === null ? "unattempted" : (q8 == 2 ? "correct" : "incorrect"));
}
ok_response(q8, "2");
```

:::

:::{.callout-caution}
## Check Your Understanding: Appropriate Plot Types II

Which of the following plot types would be appropriate for building a visual with the `class` and drive type variables? Select all that apply.

```{ojs}
//| echo: false

viewof q9 = Inputs.checkbox(
  new Map([
    ["A histogram", 1],
    ["A side-by-side boxplot", 2],
    ["A bar graph with fill color", 3],
    ["A heatmap", 4],
    ["A mosaicplot", 5]
  ]),
  {value: JSON.parse(localStorage.getItem("q9_selected_t4") ?? "[]") ?? []}
);

{
  localStorage.setItem("q9_selected_t4", JSON.stringify(q9));
  localStorage.setItem("q9_correct_t4", "3,4,5");
  localStorage.setItem("q9_result_t4", (!q9 || q9.length === 0) ? "unattempted" : (q9.toString() === "3,4,5" ? "correct" : "incorrect"));
}
ok_checkbox(q9, "3,4,5");
```

:::

Use the code block below to make a useful plot of `class` using `geom_bar()` with the aesthetics `x = class` and `fill = drv`. Think about the plot you create -- what does it tell you?

```{webr}
#| exercise: barplot-class-drv
______
```

:::{.hint exercise="barplot-class-drv"}
:::{.callout-note collapse="false"}
## Hint 1

Starting from an earlier plot can take you a long way — look for the simplest one that's closest to what you need.

:::
:::

:::{.hint exercise="barplot-class-drv"}
:::{.callout-note collapse="false"}
## Hint 2

Several things need to be changed. What are they?

```{r}
#| echo: true
#| eval: false

mpg |>
  ggplot() + 
  geom_point(aes(x = displ, y = hwy))
```

:::
:::

:::{.hint exercise="barplot-class-drv"}
:::{.callout-note collapse="false"}
## Hint 3

Four things to be updated: change the geometry layer to `geom_bar()`, map `class` to `x`, remove the `y` aesthetic (bar heights are determined by counts automatically), and map `drv` to `fill`.

```{r}
#| echo: true
#| eval: false

mpg |>
  ggplot() + 
  geom_point(aes(x = displ, y = hwy))
```

:::
:::

:::{.hint exercise="barplot-class-drv"}
:::{.callout-note collapse="false"}
## Hint 4 (Solved)

```{r}
#| echo: true
#| eval: false

mpg |>
  ggplot() + 
  geom_bar(aes(x = class, fill = drv))
```

:::
:::

```{webr}
#| exercise: barplot-class-drv
#| solution: true

mpg |>
  ggplot() + 
  geom_bar(aes(x = class, fill = drv))
```

```{webr}
#| exercise: barplot-class-drv
#| check: true

exercise_results$barplot_class_drv <<- "fail"
gradethis::grade_this({
  feedback <- gradethis::code_feedback()
  if (is.null(feedback)) {
    exercise_results$barplot_class_drv <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

**Note:** There are many different geom types, and different aesthetic properties which can be passed to geoms. We will see examples throughout the rest of this activity, but reading the entirety of Chapter 3 in Hadley Wickham's [R for Data Science](http://r4ds.had.co.nz/data-visualisation.html){target="_blank"} would be a great start for those of you who are more interested in data visualization.

## Plotting More Than Two Variables at Once

Let's go back to our original plot of `hwy` versus `displ`. Maybe we want to color the points in the scatterplot according to the class of the vehicle. Add a third aesthetic, `color = class`, to the code below and re-run the plot.

```{webr}
#| exercise: hwy-displ-class-plot

mpg |>
  ggplot() + 
  geom_point(aes(x = displ, y = hwy))
```

:::{.hint exercise="hwy-displ-class-plot"}
:::{.callout-note collapse="false"}
## Hint 1

Add `color` as a new aesthetic inside `aes()` and map the `class` variable to it.

:::
:::

:::{.hint exercise="hwy-displ-class-plot"}
:::{.callout-note collapse="false"}
## Hint 2

Fill in the blank to map the `class` variable to the `color` aesthetic.

```{r}
#| echo: true
#| eval: false

mpg |> 
  ggplot() + 
  geom_point(aes(
    x = displ, 
    y = hwy, 
    color = ___))
```

:::
:::

:::{.hint exercise="hwy-displ-class-plot"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

```{r}
#| echo: true
#| eval: false

mpg |> 
  ggplot() + 
  geom_point(aes(
    x = displ, 
    y = hwy, 
    color = class))
```

:::
:::

```{webr}
#| exercise: hwy-displ-class-plot
#| solution: true

mpg |> 
  ggplot() + 
  geom_point(aes(
    x = displ, 
    y = hwy, 
    color = class))
```

```{webr}
#| exercise: hwy-displ-class-plot
#| check: true

exercise_results$hwy_displ_class <<- "fail"
gradethis::grade_this({
  feedback <- gradethis::code_feedback()
  if (is.null(feedback)) {
    exercise_results$hwy_displ_class <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

There's a lot going on here, and it is hard to read. Try copying the above plotting command, but appending `facet_wrap(~ class, nrow = 2)` as a new layer to the plot -- notice that plot layers are added with the `+` symbol. Again, think about the resulting plot and how it compares to your previous colored plot.

```{webr}
#| exercise: facet-wrapped-plot

mpg |> 
  ggplot() + 
  geom_point(aes(
    x = displ, 
    y = hwy, 
    color = class))
```

:::{.hint exercise="facet-wrapped-plot"}
:::{.callout-note collapse="false"}
## Hint 1

New layers are added to ggplots with the `+` sign.

:::
:::

:::{.hint exercise="facet-wrapped-plot"}
:::{.callout-note collapse="false"}
## Hint 2

Fill in the blank with the `facet_wrap()` layer.

```{r}
#| echo: true
#| eval: false

mpg |> 
  ggplot() + 
  geom_point(aes(
    x = displ, 
    y = hwy, 
    color = class)) + 
  ___
```

:::
:::

:::{.hint exercise="facet-wrapped-plot"}
:::{.callout-note collapse="false"}
## Hint 3

Fill in the blank with the `facet_wrap()` layer. Check the prompt for how to use `facet_wrap()`.

```{r}
#| echo: true
#| eval: false

mpg |> 
  ggplot() + 
  geom_point(aes(
    x = displ, 
    y = hwy, 
    color = class)) + 
  facet_wrap(___)
```

:::
:::

:::{.hint exercise="facet-wrapped-plot"}
:::{.callout-note collapse="false"}
## Hint 4

The blank after the tilde (`~`) inside of `facet_wrap()` identifies the column whose categories should each get their own plot panel.

```{r}
#| echo: true
#| eval: false

mpg |> 
  ggplot() + 
  geom_point(aes(
    x = displ, 
    y = hwy, 
    color = class)) + 
  facet_wrap(~___, nrow = 2)
```

:::
:::

:::{.hint exercise="facet-wrapped-plot"}
:::{.callout-note collapse="false"}
## Hint 5 (Solved)

We want an individual plot for each of the vehicle classes.

```{r}
#| echo: true
#| eval: false

mpg |> 
  ggplot() + 
  geom_point(aes(
    x = displ, 
    y = hwy, 
    color = class)) + 
  facet_wrap(~ class, nrow = 2)
```

:::
:::

```{webr}
#| exercise: facet-wrapped-plot
#| solution: true

mpg |> 
  ggplot() + 
  geom_point(aes(
    x = displ, 
    y = hwy, 
    color = class)) + 
  facet_wrap(~ class, nrow = 2)
```

```{webr}
#| exercise: facet-wrapped-plot
#| check: true

exercise_results$facet_wrapped <<- "fail"
gradethis::grade_this({
  feedback <- gradethis::code_feedback()
  if (is.null(feedback)) {
    exercise_results$facet_wrapped <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

:::{.callout-caution}
## Check Your Understanding: Investigating Arguments

What does setting the argument `nrow = 2` do? Don't guess -- think about how you can determine what the `nrow` argument here does.

```{ojs}
//| echo: false

viewof q10 = Inputs.radio(
  new Map([
    ["It is impossible to tell.", 1],
    ["It tells ggplot to only use half the data so that plots are less cluttered.", 2],
    ["It tells R to try arranging the plots into two rows.", 3],
    ["It tells R to include two plots per row.", 4],
    ["Nothing -- it is irrelevant and ignored.", 5]
  ]),
  {value: JSON.parse(localStorage.getItem("q10_selected_t4") ?? "null")}
);

{
  localStorage.setItem("q10_selected_t4", JSON.stringify(q10));
  localStorage.setItem("q10_correct_t4", "3");
  localStorage.setItem("q10_result_t4", q10 === null ? "unattempted" : (q10 == 3 ? "correct" : "incorrect"));
}
ok_response(q10, "3");
```

:::

## Categorical Variables

We've seen plots for categorical variables earlier in this workbook. Let's revisit a barplot with fill, similar to the one we constructed between `class` and `drv` earlier and check two alternative plots which give different insights. We'll explore the manual versus automatic transmissions.

```{r}
#| echo: true
#| eval: false

mpg |>
  count(trans)
```

```{r}
#| echo: false
#| eval: true

mpg |>
  count(trans) |> 
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Unfortunately the `trans` variable contains more detailed transmission information than what we are after. Omitting the last four characters from levels in the `trans` column would leave us with the information we want. While this sort of data wrangling is beyond the scope of these notebooks (and you won't be asked to do it on your own), I'll still show you how we can remove those last four characters below.

We can achieve our objective using the `str_sub()` function -- this function *subsets a string*. Again, you won't need this function for our course but, because we're using it here, I'll explain how it works. The first argument is the name of the column whose strings we want to subset, the second argument is the position of the first character we'd like to keep and the third argument is the position of the final character we'd like to keep. Using a negative number here is shorthand for "from the end". That is, we'd like to keep all of the characters starting with the first and stopping five characters from the end of the string.

Run the code below to see the plot.

```{webr}
#| exercise: engineer-trans2
#| fig-alt: "Bar chart of the count of vehicles by transmission type. Automatic transmissions are about twice as common as manual, with roughly 157 automatic vehicles compared to about 77 manual vehicles."

mpg |>
  mutate(trans2 = str_sub(trans, 1, -5)) |>
  ggplot() + 
  geom_bar(aes(x = trans2)) +
  labs(x = "Transmission Type", 
       y = "Count",
       title = "Distribution of Transmission Types")
```

From the plot above we can tell that there are about twice as many vehicles in our dataset with automatic transmissions as there are with manual transmissions. Create a similar plot below, which fills the bars by the `drv` variable. Include the same title and axis labels as the plot above has.

```{webr}
#| exercise: trans2-fill-class
______
```

:::{.hint exercise="trans2-fill-class"}
:::{.callout-note collapse="false"}
## Hint 1

Copy and paste the plotting code above and use it as a starting point.

:::
:::

:::{.hint exercise="trans2-fill-class"}
:::{.callout-note collapse="false"}
## Hint 2

Do you remember how to add a `fill` aesthetic?

```{r}
#| echo: true
#| eval: false

mpg |>
  mutate(trans2 = str_sub(trans, 1, -5)) |>
  ggplot() + 
  geom_bar(aes(x = trans2)) +
  labs(x = "Transmission Type", 
       y = "Count",
       title = "Distribution of Transmission Types")
```

:::
:::

:::{.hint exercise="trans2-fill-class"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

Map the drive variable (`drv`) to the `fill` aesthetic inside `aes()`.

```{r}
#| echo: true
#| eval: false

mpg |>
  mutate(trans2 = str_sub(trans, 1, -5)) |>
  ggplot() + 
  geom_bar(aes(x = trans2, fill = drv)) +
  labs(x = "Transmission Type", 
       y = "Count",
       title = "Distribution of Transmission Types")
```

:::
:::

```{webr}
#| exercise: trans2-fill-class
#| solution: true

mpg |>
  mutate(trans2 = str_sub(trans, 1, -5)) |>
  ggplot() + 
  geom_bar(aes(x = trans2, fill = drv)) +
  labs(x = "Transmission Type", 
       y = "Count",
       title = "Distribution of Transmission Types")
```

```{webr}
#| exercise: trans2-fill-class
#| check: true

exercise_results$trans2_fill_class <<- "fail"
gradethis::grade_this({
  feedback <- gradethis::code_feedback()
  if (is.null(feedback)) {
    exercise_results$trans2_fill_class <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

The problem now is the difficulty in comparing the proportions of drive types within each of these classes. We can fix this by using the `position` argument. Outside of `aes()` but still within `geom_bar()`, add an argument `position = "fill"` to the pre-built plot. You'll need to include a comma after `aes()` since commas separate arguments. Think about what is gained and lost in this new plot.

```{webr}
#| exercise: trans2-drv-plot

mpg |>
  mutate(trans2 = str_sub(trans, 1, -5)) |>
  ggplot() + 
  geom_bar(aes(
    x = trans2, 
    fill = drv)) +
  labs(x = "Transmission Type", 
       y = "Count",
       title = "Distribution of Transmission Types")
```

:::{.hint exercise="trans2-drv-plot"}
:::{.callout-note collapse="false"}
## Hint 1

Since we aren't mapping a column of our data set to the `position` parameter, we add it outside of the `aes()` function.

:::
:::

:::{.hint exercise="trans2-drv-plot"}
:::{.callout-note collapse="false"}
## Hint 2

Fill in the blank to set the position to `"fill"`.

```{r}
#| echo: true
#| eval: false

mpg |>
  mutate(trans2 = str_sub(trans, 1, -5)) |>
  ggplot() + 
  geom_bar(aes(
    x = trans2, 
    fill = drv),
    position = ___) +
  labs(x = "Transmission Type", 
       y = "Count",
       title = "Distribution of Transmission Types")
```

:::
:::

:::{.hint exercise="trans2-drv-plot"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

```{r}
#| echo: true
#| eval: false

mpg |>
  mutate(trans2 = str_sub(trans, 1, -5)) |>
  ggplot() + 
  geom_bar(aes(
    x = trans2, 
    fill = drv), 
    position = "fill") +
  labs(x = "Transmission Type", 
       y = "Count",
       title = "Distribution of Transmission Types")
```

:::
:::

```{webr}
#| exercise: trans2-drv-plot
#| solution: true

mpg |>
  mutate(trans2 = str_sub(trans, 1, -5)) |>
  ggplot() + 
  geom_bar(aes(
    x = trans2, 
    fill = drv), 
    position = "fill") +
  labs(x = "Transmission Type", 
       y = "Count",
       title = "Distribution of Transmission Types")
```

```{webr}
#| exercise: trans2-drv-plot
#| check: true

exercise_results$trans2_drv_plot <<- "fail"
gradethis::grade_this({
  feedback <- gradethis::code_feedback()
  if (is.null(feedback)) {
    exercise_results$trans2_drv_plot <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

Great! But the problem now is that we've lost the idea that there are more automatic vehicles than there are manual vehicles in this dataset. As a third option, we can consider a mosaic plot.

```{r}
#| echo: true
#| eval: true
#| fig-alt: "Mosaic plot showing drive type by transmission type, where bar width represents the proportion of vehicles with each transmission and segment height represents the proportion of each drive type within that transmission. The auto column is about twice as wide as the manual column, reflecting that automatics are roughly twice as common overall. Within automatic transmissions, about 47% are 4-wheel drive, 42% are front-wheel drive, and about 11% are rear-wheel drive. Within manual transmissions, front-wheel drive is more common at about 53%, 4-wheel drive drops to about 36%, and rear-wheel drive is similar at about 12%. The main pattern is that manual transmissions skew more toward front-wheel drive and less toward 4-wheel drive compared to automatics, while rear-wheel drive is roughly equally rare in both."

mpg |>
  mutate(trans2 = str_sub(trans, 1, -5)) |>
  ggplot() + 
  geom_mosaic(aes(
    x = product(trans2), 
    fill = drv)) +
  labs(x = "Transmission Type", 
       y = "Drive Type",
       title = "")
```

The advantage to the mosaic plot is that it contains all of the information from the two separate plots above, but it is all combined into one plot! We've built a visualization that manages to convey that the proportion of automatic vehicles in our data set outweighs the proportion of manuals, and we can also easily compare the distribution of drive types within each of these two classes.

## Submit

If you are part of a course with an instructor who is grading your work on these activities, please copy and submit ***both*** of the hashes below using the method your instructor has requested.

:::{.callout-warning}
## Question Hash

The hash below encodes your responses to the multiple choice and checkbox questions in this activity.

```{ojs}
//| echo: false

function buildQuestionResults() {
  return {
    notebook: "Topic 4: Data Visualization",
    type: "questions",
    timestamp: new Date().toISOString(),
    questions: {
      q1_better_plot: {
        selected: q1,
        correct_answer: "2,3",
        result: (!q1 || q1.length === 0) ? "unattempted" : (q1.toString() === "2,3" ? "correct" : "incorrect")
      },
      q2_variables: {
        selected: q2,
        correct_answer: "2",
        result: q2 === null ? "unattempted" : (q2 == 2 ? "correct" : "incorrect")
      },
      q3_observations: {
        selected: q3,
        correct_answer: "3",
        result: q3 === null ? "unattempted" : (q3 == 3 ? "correct" : "incorrect")
      },
      q4_numerical_variables: {
        selected: q4,
        correct_answer: "3,4,6,8,9",
        result: (!q4 || q4.length === 0) ? "unattempted" : (q4.toString() === "3,4,6,8,9" ? "correct" : "incorrect")
      },
      q5_reading_plot_1: {
        selected: q5,
        correct_answer: "2",
        result: q5 === null ? "unattempted" : (q5 == 2 ? "correct" : "incorrect")
      },
      q6_reading_plot_2: {
        selected: q6,
        correct_answer: "3",
        result: q6 === null ? "unattempted" : (q6 == 3 ? "correct" : "incorrect")
      },
      q7_reading_plot_3: {
        selected: q7,
        correct_answer: "1",
        result: q7 === null ? "unattempted" : (q7 == 1 ? "correct" : "incorrect")
      },
      q8_appropriate_plots_1: {
        selected: q8,
        correct_answer: "2",
        result: q8 === null ? "unattempted" : (q8 == 2 ? "correct" : "incorrect")
      },
      q9_appropriate_plots_2: {
        selected: q9,
        correct_answer: "3,4,5",
        result: (!q9 || q9.length === 0) ? "unattempted" : (q9.toString() === "3,4,5" ? "correct" : "incorrect")
      },
      q10_nrow_argument: {
        selected: q10,
        correct_answer: "3",
        result: q10 === null ? "unattempted" : (q10 == 3 ? "correct" : "incorrect")
      }
    }
  };
}

function toBase64(str) {
  return btoa(unescape(encodeURIComponent(str)));
}

question_hash = {
  q1; q2; q3; q4; q5; q6; q7; q8; q9; q10;
  return toBase64(JSON.stringify(buildQuestionResults()));
}
```

```{ojs}
//| echo: false

html`<div style="font-family: monospace; font-size: 0.85em; background: #f5f5f5; padding: 12px; border-radius: 6px; word-break: break-all; border: 1px solid #ddd; user-select: all; cursor: pointer;" onclick="navigator.clipboard.writeText(this.innerText)">
  ${question_hash}
</div>
<p style="margin-top: 8px; font-size: 0.9em; color: #555;">
  Click the box to copy to clipboard.
</p>`
```

:::

:::{.callout-warning}
## Exercise Hash

Click the button below to generate your exercise submission code. This hash encodes your work on the graded code exercises in this activity.

You must have attempted the graded exercises before clicking — clicking generates a snapshot of your current results. If you have completed the activity over multiple sessions, please go back through and hit the *Run Code* button on each graded exercise before generating the hash below, to ensure your most recent results are recorded.

```{webr}
#| exercise: generate-exercise-hash
#| caption: "Generate Exercise Code"

exercise_results$timestamp <- format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
exercise_results$type <- "exercises"

hash <- base64encode(charToRaw(toJSON(exercise_results, auto_unbox = TRUE)))
cat(hash)
```

:::

## Summary

:::{.callout-tip}
## Main Takeaways

+ **Variable types dictate plot types.** Always consider whether your variables are numerical or categorical before choosing a plot. A scatterplot requires two numerical variables; a histogram, boxplot, or density plot give insight to the distribution of a single numerical variable; a bar plot is appropriate for a single categorical variable and adding a fill color accommodates a second categorical variable; side-by-side boxplots work for numerical vs. categorical comparisons.
+ **Plotting with `ggplot()` follows a grammar of graphics** and a layered structure, with `+` separating each layer. The basic structure is:

```{r}
#| echo: true
#| eval: false

data_frame_name |>
  ggplot() +
  geom_TYPE(aes(MAPPINGS))
```

+ **Aesthetics** (`aes()`) map variables from your data to visual properties of the plot (position, color, fill, size, shape). Properties that don't depend on the data go *outside* `aes()`.
    + The plot type (geom type) determines which aesthetics are required and available.
+ **Additional dimensions** can be added via color, fill, size, shape, or faceting — but more information in a single plot can make it harder to read. Sometimes separate faceted plots communicate more clearly than a single dense plot.
+ **Mosaic plots** are a powerful choice for two categorical variables — they simultaneously convey both the marginal distributions and the conditional distributions within each group.
:::

:::{.callout-tip}
## Looking Ahead

This activity concludes our foundational and "R-onramping" work. Over the next several workbooks, we'll transition to probability and working with probability distributions. We'll start with basic, counting-based probability before moving to the binomial distribution and normal distribution.
:::

:::{.callout-note}
## Loaded Libraries

These interactive workbooks allow setup requirements to be hidden so that we can focus on the task at hand — in this workbook, plotting. When you make the switch to using a standalone R instance you'll need to install and load libraries to expand R's functionality. You can do this with the `install.packages()` and `library()` functions. You would need `{ggplot2}`, `{dplyr}`, `{skimr}`, and `{ggmosaic}` to run the code for this workbook in a local R session.
:::

---

## References

[1] Wickham, H., & Grolemund, G. (2017). R for data science: Import, tidy, transform, visualize, and model data. O'Reilly Media.
[2] Wickham, H. (2016). ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York. [dataset]. https://ggplot2.tidyverse.org/reference/mpg.html
