---
title: "Topic 4: Data Visualization and a Grammar of Graphics"
format: 
  live-html:
    include-in-header:
      text: |
        <style>
        .btn-exercise-solution { display: none !important; }
        </style>
    webr:
      packages:
        - dplyr
        - gradethis
      cell-options: 
        persist: true
    live:
      show-solutions: true
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}
{{< include ./_extensions/r-wasm/live/_gradethis.qmd >}}

```{webr}
#| edit: false
#| output: false
#| define:
#|   - ok_response

library(htmltools)
ok_response <- function(response, n) {
  if (is.na(response)) HTML("")
  else if (response == n) div(HTML("Correct ✓"), style = "color: green")
  else div(HTML("Incorrect ✗"), style = "color: red")
}
```

```{webr}
#| edit: false
#| output: false
#| define:
#|   - ok_checkbox

ok_checkbox <- function(response, n) {
  if (is.na(response)) HTML("")
  else if (response == n) div(HTML("Correct ✓"), style = "color: green")
  else div(HTML("Not Yet! ✗"), style = "color: red")
}
```

```{webr}
#| edit: false
#| output: false
#| define:
#|   - normalize_checkbox_response

normalize_checkbox_response <- function(x) {
  return(cat(sort(x), sep = ","))
}
```

```{webr}
#| context: setup
#| echo: false
#| eval: true
#| message: false
#| warning: false

library(ggplot2)
library(stringr)
library(skimr)
```

:::{.callout-note}
## About

This activity provides an introduction to the grammar of graphics, plotting strategies, and using ggplot. The discussion you'll work through loosely follows Chapter 3 of the first edition Wickham and Grolemund's R for Data Science text. Here's [a link to the updated second edition of the text](https://r4ds.hadley.nz/).
:::

```{r setup, include=FALSE}
#knitr::opts_chunk$set(eval = FALSE)
#library(tidyverse)
library(dplyr)
library(ggplot2)
library(stringr)
library(gradethis)
library(skimr)
library(ggmosaic)
library(kableExtra)

data <- data.frame("FuelType" = c("F1", "F1", "F1", "F2", "F2", "F2"), "EngineBrand" = c("E1", "E2", "E3", "E1", "E2", "E3"))
```

## Data Visualization

Through this workbook you'll learn techniques for data visualization. You'll think about how the type of plot chosen impacts what information is conveyed (or lost). Choosing the wrong plot type can result in a useless plot (at best) or even a plot which is misleading. Think about the *types* (numerical, categorical) of variables you are working with and how they dictate which plot type should be utilized -- for example, a scatterplot is not appropriate in every scenario!

## Objectives

**Workbook Objectives:** After completing this workbook you should be able to:

+ Read a plot, describe any interesting features, and accurately convey the *story* being told by the plot.
+ Identify and utilize the most appropriate plot types for your scenario by considering the *types* of variables you are working with.
+ Utilize color, size, and shape to show additional dimensions of data in a plot, but also recognize that adding more information to a plot negatively impacts our ability to read the graphic. 
+ Use your new found knowledge of data visualization to generate informative plots which uncover a compelling (and truthful) story about the relationships that exist between variables.

Note, much of this notebook is inspired by [Chapter 3 of R for Data Science](https://r4ds.had.co.nz/data-visualisation.html){target="_blank"} by Hadley Wickham and Garrett Grolemund. This is a great freely-available resource. Please check it out.

## The Grammar of Graphics

**Grammar of Graphics:** Jeffrey Gitomer said "Your grammar is a reflection of your image". Here, we take the converse literally. Your image is a reflection of your grammar. Just like a well-written sentence follows the rules of grammar, so does an informative statistical graphic.

* Graphics need a subject (an underlying dataset)
* Graphics need verbs (a type of plotting structure, called a `geom`)
* Graphics need adjectives (attributes attached to the visualized data, called `aesthetics`)
* Graphics need context (a title, scales, axes, legends, etc.)

In this workbook we explore `ggplot2` and think about graphics in terms of their layered grammar. 

**A Note on Plots:** Choosing an appropriate plot is extremely important in data visualization. Some plots don't make sense for certain variable types. For example, a scatterplot in the case of two-categorical predictors is quite a silly choice -- the only thing this plot tells us is that every combination of fuel type and engine brand exists in our dataset. We have no idea which combinations are most or least popular. 

```{r}
#| label: first-plot
#| echo: false
#| eval: true
#| fig.height: 2.25
#| fig.width: 2.25

data |> 
  ggplot() + 
  geom_point(aes(x = FuelType, y = EngineBrand), size = 2) + 
  labs(y = "Engine Brand", 
       x = "Fuel Type")
```

The following are some recommended [basic] plots types under certain scenarios:

  + A single numerical variable: A histogram, boxplot, or density
  + A single categorical variable: A bar plot
  + Numerical versus Numerical: A scatterplot, heatmap
  + Numerical versus categorical: side-by-side boxplots, stacked or overlayed densities, stacked or overlayed histograms
  + Categorical versus categorical: barplot with fill color, mosaic plot, heatmap

:::{.callout-caution}
## Check Your Understanding: A Better Plot

Which of the following plot types would have been a better choice to display the engine brand and fuel type data?

```{ojs}
//| echo: false

mutable ok_checkbox = (response, n) => { return html`Loading...` };
viewof q1 = Inputs.checkbox(
  new Map([
    ["A histogram", 1],
    ["A barplot with fill color", 2],
    ["A heatmap", 3],
    ["A dotplot", 4],
    ["Side-by-side boxplots", 5]
  ])
);

ok_checkbox(q1.toString(), "2,3");
```

:::

## Getting Started: Exploring Data

Throughout this workbook, we will explore fuel efficiencies of multiple classes of vehicles using the `mpg` data frame.

1. Remember that a *data frame* is like an Excel spreadsheet. Explore the `mpg` data frame by typing `mpg` in the following code block and running it.

```{webr}
#| exercise: print-mpg-dataframe
______
```

```{webr}
#| exercise: print-mpg-dataframe
#| solution: true
mpg
```

```{webr}
#| exercise: print-mpg-dataframe
#| check: true

gradethis::grade_this_code()
```

Notice that when you execute a line of code which just calls the name of a data frame, a snippet of that data frame is printed out. This is true for other objects (variables, vectors, functions, etc) in R as well.

**Some Basic Exploratory Functions:** It is useful to know more about your dataset when you first start working with data. R has a few exploratory functions which you should know about: `mpg |> names()` prints out a list of names of your data frame's columns, `mpg |> head()` will show you the first six rows of the `mpg` dataset, `mpg |> dim()` will show you the number of rows and columns in the dataset, and `mpg |> glimpse()` will show you information about how R is treating the columns (`int` and `num` denote numerical values while `chr` and `fct` represent character strings and categorical variables respectively). You might also try `mpg |> skim()`, where the `skim()` function comes from the `{skimr}` package. The `skim()` function provides lots of high-level information about your dataset from just a single function call. Use the code block below to try each of these functions and use the output to answer the following questions:

```{webr}
#| exercise: explore-mpg-dataframe
______
```

:::{.hint exercise="explore-mpg-dataframe"}
:::{.callout-note collapse="false"}
## Hint 1

```{r}
#| echo: true
#| eval: false
#Try some of the functions listed above.
```

:::
:::

:::{.hint exercise="explore-mpg-dataframe"}
:::{.callout-note collapse="false"}
## Hint 2

```{r}
#| echo: true
#| eval: false
#Fill in the blanks with the name of 
#the data frame.

___ |>
  names()

___ |>
  head()

___ |>
  dim()

___ |>
  glimpse()
  
___ |> 
  skim()
```

:::
:::

:::{.hint exercise="explore-mpg-dataframe"}
:::{.callout-note collapse="false"}
## Hint 3

```{r}
#| echo: true
#| eval: false
mpg |>
  names()

mpg |>
  head()

mpg |>
  dim()

mpg |>
  glimpse()

mpg |> 
  skim()
```

:::
:::

:::{.callout-caution}
## Check Your Understanding: Variables

How many variables are contained in the mpg dataset?

```{ojs}
//| echo: false

mutable ok_response = (response, n) => { return html`Loading...` };
viewof q2 = Inputs.radio(
  new Map([
    ["1", 1],
    ["11", 2],
    ["234", 3],
    ["245", 4],
    ["2574", 5]
  ])
);

ok_response(q2, "2");
```

:::

:::{.callout-caution}
## Check Your Understanding: Observations

How many observations (records) are contained in the mpg dataset?

```{ojs}
//| echo: false

viewof q3 = Inputs.radio(
  new Map([
    ["1", 1],
    ["11", 2],
    ["234", 3],
    ["245", 4],
    ["2574", 5]
  ])
);

ok_response(q3, "3");
```

:::

:::{.callout-caution}
## Check Your Understanding: Numerical Variables

Which of the variables is R interpreting as numerical?

```{ojs}
//| echo: false

viewof q4 = Inputs.checkbox(
  new Map([
    ["manufacturer", 1],
    ["model", 2],
    ["displ (displacement)", 3],
    ["cyl (cylinders)", 4],
    ["trans (transmission type)", 5],
    ["year", 6],
    ["drv (drive type)", 7],
    ["cty (city fuel economy)", 8],
    ["hwy (highway fuel economy)", 9],
    ["fl (fuel type)", 10],
    ["class (vehicle class)", 11]
  ])
);

ok_checkbox(q4.toString(), "3,4,6,8,9");
```

:::

The `diamonds` data frame is also available to you. See if you can get your basic exploratory functions to help you answer the same questions about the `diamonds` dataset using the code block below.

```{webr}
#| exercise: explore-diamonds
______
```

:::{.hint exercise="explore-diamonds"}
:::{.callout-note collapse="false"}

```{r}
#| echo: true
#| eval: false
#Use some of the same functions you used 
#to explore the mpg data frame in the 
#previous code chunk.
```

:::
:::

Were you able to obtain the information you were looking for? Be sure to ask a question if not!

## Building and interpreting plots

The code below creates a plot of highway miles per gallon (`hwy`) against engine displacement `displ` (a measure of the size of an engine).

```{r echo=TRUE, eval=TRUE, message = FALSE, warning = FALSE, results = FALSE}
#| label: displ-hwy-plot
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| results: false

mpg |> 
  ggplot() + 
  geom_point(aes(x = displ, y = hwy))
```

Use the plot above to answer the following questions.

:::{.callout-caution}
## Check Your Understanding: Reading a Plot I

Which of the following is true?

```{ojs}
//| echo: false

viewof q5 = Inputs.radio(
  new Map([
    ["Highway miles per gallon and engine displacement are independent", 1],
    ["Highway miles per gallon and engine displacement are associated", 2],
    ["It is impossible to tell whenther the variables are associated or independent", 3],
    ["Highway miles per gallon and engine displacement are both independent and associated", 4]
  ])
);

ok_response(q5, "2");
```

:::

:::{.callout-caution}
## Check Your Understanding: Reading a Plot II

Describe the relationship between highway miles per gallon and engine displacement.

```{ojs}
//| echo: false

viewof q6 = Inputs.radio(
  new Map([
    ["Engine displacement and highway miles per gallon share a negative, linear relationship", 1],
    ["Engine displacement and highway miles per gallon share a positive, linear relationship", 2],
    ["Engine displacement and highway miles per gallon share a negative, non-linear relationship", 3],
    ["Engine displacement and highway miles per gallon share a positive, non-linear relationship", 4]
  ])
);

ok_response(q6, "3");
```

:::

:::{.callout-caution}
## Check Your Understanding: Reading a Plot III

Which is the best practical interpretation of the plot?

```{ojs}
//| echo: false

viewof q7 = Inputs.radio(
  new Map([
    ["Generally, engines with higher displacement are associated with lower average highway miles per gallon ratings", 1],
    ["Generally, engines with higher displacement are associated with higher average highway miles per gallon ratings", 2],
    ["Higher engine displacement causes lower average highway miles per gallon ratings", 3],
    ["Engine displacement causes higher average highway miles per gallon ratings", 4]
  ])
);

ok_response(q7, "1");
```

:::

**A Note on Plotting Structure:** Using `ggplot()` to create plots in R is daunting at first, but with practice you will notice the structure is very consistent and convenient!

+ Our plot consists of at least two pieces (subject and verb), `mpg |> ggplot()` tells `ggplot` that the subject of our plot is the data contained in the `mpg` table, while the `geom_point()` verb tells `ggplot` that we want our data displayed as points (a scatterplot). The `aes()` inside of `geom_point()` tells `ggplot` some of the adjectives for each individual observation -- here, just the location!
+ In general a simple plot takes the form below -- (where the items in "ALL CAPS" are replaced by your data frame, desired geometry type, and aesthetic mappings).

```{r}
#| label: sample-plot
#| echo: true
#| eval: false

DATA |>
  ggplot() + 
  geom_TYPE(aes(MAPPINGS))
```

Use the code block below to make a scatterplot of average city miles per gallon (`cty`) explained by engine displacement (`displ`).

```{webr plot-displ-cty}
#| exercise: plot-displ-cty
______
```

:::{.hint exercise="plot-displ-cty"}
:::{.callout-note collapse="false"}
## Hint 1

```{r}
#| echo: true
#| eval: false
#Start by copying and pasting the code from 
#the plot at the start of this subsection.

```

:::
:::

:::{.hint exercise="plot-displ-cty"}
:::{.callout-note collapse="false"}
## Hint 2

```{r}
#| echo: true
#| eval: false

mpg |> 
  ggplot() + 
  geom_point(aes(x = displ, y = hwy))

#What needs to be changed here?
```

:::
:::

:::{.hint exercise="plot-displ-cty"}
:::{.callout-note collapse="false"}
## Hint 3

```{webr}
#| echo: true
#| eval: false
#Swap out the hwy variable for the cty 
#variable and generate the new plot.
mpg |> 
  ggplot() + 
  geom_point(aes(x = displ, y = hwy))
```

:::
:::

```{webr}
#| exercise: plot-displ-cty
#| solution: true
mpg |>
  ggplot() + 
  geom_point(aes(x = displ, y = cty))
```

```{webr}
#| exercise: plot-displ-cty
#| check: true

gradethis::grade_this_code()
```

Now make a scatterplot of average highway miles per gallon (`hwy`) explained by number of cylinders (`cyl`).

```{webr plot-hwy-cyl, exercise = TRUE}
#| exercise: plot-hwy-cyl
______
```


:::{.hint exercise="plot-hwy-cyl"}
:::{.callout-note collapse="false"}
## Hint 1

```{r}
#| echo: true
#| eval: false
#Again, start by copying and pasting an 
#earlier plot.

```

:::
:::

:::{.hint exercise="plot-hwy-cyl"}
:::{.callout-note collapse="false"}
## Hint 2

```{r}
#| echo: true
#| eval: false
mpg |>
  ggplot() + 
  geom_point(aes(x = displ, y = cty))

#What needs to be changed now?
```

:::
:::

:::{.hint exercise="plot-hwy-cyl"}
:::{.callout-note collapse="false"}
## Hint 3

```{webr}
#| echo: true
#| eval: false
#Put cylinders (cyl) on the x-axis and 
#highway gas mileage (hwy) on the y-axis.
mpg |>
  ggplot() + 
  geom_point(aes(x = displ, y = cty))
```

:::
:::

```{webr}
#| exercise: plot-hwy-cyl
#| solution: true
mpg |>
  ggplot() + 
  geom_point(aes(x = cyl, y = hwy))
```

```{webr}
#| exercise: plot-hwy-cyl
#| check: true
gradethis::grade_this_code()
```

Notice that this plot isn't very useful because the cylinders variable takes on very few levels. We may be better off if we treat `cyl` as if it were a categorical variable here. Check out some of the other geometry layers available to you in ggplot [here](https://ggplot2.tidyverse.org/reference/). Try building a set of side-by-side boxplots. If you get an error, be sure to read it -- R suspects that the plot produced wasn't the one you wanted and provides a suggestion.

```{webr}
#| exercise: boxplot-hwy-cyl
______
```

:::{.hint exercise="boxplot-hwy-cyl"}
:::{.callout-note collapse="false"}
## Hint 1

```{r}
#| echo: true
#| eval: false
#Continue copying/pasting an earlier plot 
#as a starting point for now. You'll be 
#able to construct your own plots from 
#scratch with practice.

```

:::
:::

:::{.hint exercise="boxplot-hwy-cyl"}
:::{.callout-note collapse="false"}
## Hint 2

```{r}
#| echo: true
#| eval: false
mpg |>
  ggplot() + 
  geom_point(aes(x = cyl, y = hwy))

#What needs to be changed here?
```

:::
:::

:::{.hint exercise="boxplot-hwy-cyl"}
:::{.callout-note collapse="false}
## Hint 3

```{r}
#| echo: true
#| eval: false
#Swap geom_point() out with geom_boxplot()
#and generate the plot.
mpg |>
  ggplot() + 
  geom_point(aes(x = cyl, y = hwy))
```

:::
:::

:::{.hint exercise="boxplot-hwy-cyl"}
:::{.callout-note collapse="false"}
## Hint 4

```{r boxplot-hwy-cyl-hint-4}
#Can you decode the warning message?
mpg |>
  ggplot() + 
  geom_boxplot(aes(x = cyl, y = hwy))
```

:::
:::

:::{.hint exercise="boxplot-hwy-cyl"}
:::{.callout-note collapse="false"}
## Hint 5

```{r}
#| echo: true
#| eval: false
#R is considering both the x- and y-variables
#to be numeric. The message suggests adding a 
#group parameter inside of the aes() function.
#Which variable should be mapped to group?
mpg |>
  ggplot() + 
  geom_boxplot(aes(x = cyl, y = hwy))
```

:::
:::

:::{.hint exercise="boxplot-hwy-cyl"}
:::{.callout-note collapse="false"}
## Hint 6

```{r}
#| echo: true
#| eval: false
#We want a boxplot for each cylinder "group". 
#Fill in the blank below to map cyl to group
#and then generate the plot.
mpg |>
  ggplot() + 
  geom_boxplot(aes(
    x = cyl, 
    y = hwy,
    group = ___))
```

:::
:::

```{webr}
#| exercise: boxplot-hwy-cyl
#| solution: true
mpg |>
  ggplot() + 
  geom_boxplot(aes(
    x = cyl, 
    y = hwy, 
    group = cyl))
```

```{webr}
#| exercise: boxplot-hwy-cyl
#| check: true

gradethis::grade_this_code()
```

Consider the variables for vehicle class (`class`) and drive type (`drv`) as you answer the following questions. 

:::{.callout-caution}
## Check Your Understanding: Appropriate Plot Types I

Why would a scatterplot not be an appropriate visual to use with these two variables?

```{ojs}
//| echo: false

viewof q8 = Inputs.radio(
  new Map([
    ["Trick question -- a scatterplot is always appropriate", 1],
    ["At least one of the two variables is not numerical", 2],
    ["A scatterplot would be okay but there are better options", 3],
    ["There isn't enough data for a scatterplot", 4]
  ])
);

ok_response(q8, "2");
```

:::

:::{.callout-caution}
## Check Your Understanding: Appropriate Plot Types II

Which of the following plot types would be appropriate for building a visual with the class and drive type variables?

```{ojs}
//| echo: false

viewof q9 = Inputs.checkbox(
  new Map([
    ["A histogram", 1],
    ["A side-by-side boxplot", 2],
    ["A bar graph with fill color", 3],
    ["A heatmap", 4],
    ["A mosaicplot", 5]
  ])
);

ok_checkbox(q9.toString(), "3,4,5");
```

:::

Use the code block below to make a useful plot of `class` using `geom_bar()` with the aesthetics `x = class` and `fill = drv`. Think about the plot you create -- what does it tell you?

```{webr}
#| exercise: barplot-class-drv
______
```

:::{.hint exercise="barplot-class-drv"}
:::{.callout-note collapse="false"}
## Hint 1

```{r}
#| echo: true
#| eval: false
#Starting from an earlier plot can take you 
#a long way!

```

:::
:::

:::{.hint exercise="barplot-class-drv"}
:::{.callout-note collapse="false"}
## Hint 2

```{r}
#| echo: true
#| eval: false
#A relatively simple plot is the best starting point...

```

:::
:::

:::{.hint exercise="barplot-class-drv"}
:::{.callout-note collapse="false"}
## Hint 3

```{r}
#| echo: true
#| eval: false
mpg |>
  ggplot() + 
  geom_point(aes(x = displ, y = hwy))

#Several things need to be changed now.
#What are they?
```

:::
:::

:::{.hint exercise="barplot-class-drv"}
:::{.callout-note collapse="false"}
## Hint 4

```{r}
#| echo: true
#| eval: false
#1) Change the geometry layer to geom_bar()
#2) Map the class variable to x
#3) The heights of the bars in a barplot are 
#determined by the frequencies (counts) of 
#observations in each category. We can't 
#prescribe a y value -- remove it.
#4) Map the drive variable (drv) to the fill 
#parameter.
mpg |>
  ggplot() + 
  geom_point(aes(x = displ, y = hwy))
```

:::
:::

:::{.hint exercise="barplot-class-drv"}
:::{.callout-note collapse="false"}
## Hint 5 (Solved)

```{r}
#| echo: true
#| eval: false

mpg |>
  ggplot() + 
  geom_bar(aes(x = class, fill = drv))
```

:::
:::

```{webr}
#| exercise: barplot-class-drv
#| solution: true

mpg |>
  ggplot() + 
  geom_bar(aes(x = class, fill = drv))
```

```{webr}
#| exercise: barplot-class-drv
#| check: true

gradethis::grade_this_code()
```

**Note:** There are many different geom types, and different aesthetic properties which can be passed to geoms. We will see examples throughout the rest of this activity, but reading the entirety of Chapter 3 in Hadley Wickham's [R for Data Science](http://r4ds.had.co.nz/data-visualisation.html) would be a great start for those of you who are more interested in data visualization.

## Plotting more than two variables at once

Let's go back to our original plot of `hwy` versus `displ`. Maybe we want to color the points in the scatterplot according to the class of the vehicle. Add a third asthetic, `color = class` to the code below and re-run the plot.

```{webr}
#| exercise: hwy-displ-class-plot
mpg |>
  ggplot() + 
  geom_point(aes(x = displ, y = hwy))
```

:::{.hint exercise="hwy-displ-class-plot"}
:::{.callout-note collapse="false"}
## Hint 1

```{r}
#| echo: true
#| eval: false
#Add color as a new aesthetic and map the 
#class variable to it.
```

:::
:::

:::{.hint exercise="hwy-displ-class-plot"}
:::{.callout-note collapse="false"}
## Hint 2

```{r}
#| echo: true
#| eval: false
#Fill in the blank to map the class variable 
#to the color aesthetic.
mpg |> 
  ggplot() + 
  geom_point(aes(
    x = displ, 
    y = hwy, 
    color = ___))
```

:::
:::

:::{.hint exercise="hwy-displ-class-plot"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

```{r}
#| echo: true
#| eval: false
mpg |> 
  ggplot() + 
  geom_point(aes(
    x = displ, 
    y = hwy, 
    color = class))
```

:::
:::

```{webr}
#| exercise: hwy-displ-class-plot
#| solution: true
mpg |> 
  ggplot() + 
  geom_point(aes(
    x = displ, 
    y = hwy, 
    color = class))
```

```{webr}
#| exercise: hwy-displ-class-plot
#| check: true

gradethis::grade_this_code()
```

There's a lot going on here, and it is hard to read. Try copying the above plotting command, but appending `facet_wrap(~ class, nrow = 2)` as a new layer to the plot -- notice that plot layers are added with the `+` symbol. Again, think about the resulting plot and how it compares to your previous colored plot.

```{webr}
#| exercise: facet-wrapped-plot
mpg |> 
  ggplot() + 
  geom_point(aes(
    x = displ, 
    y = hwy, 
    color = class))
```

:::{.hint exercise="facet-wrapped-plot"}
:::{.callout-note collapse="false"}
## Hint 1

```{r}
#| echo: true
#| eval: false
#You add new layers to ggplots with the 
#plus (+) sign.
```

:::
:::

:::{.hint exercise="facet-wrapped-plot"}
:::{.callout-note collapse="false"}
## Hint 2

```{r}
#| echo: true
#| eval: false
#Fill in the blank with the facet_wrap() 
#layer.
mpg |> 
  ggplot() + 
  geom_point(aes(
    x = displ, 
    y = hwy, 
    color = class)) + 
  ___

```

:::
:::

:::{.hint exercise="facet-wrapped-plot"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

```{r}
#| echo: true
#| eval: false
mpg |> 
  ggplot() + 
  geom_point(aes(
    x = displ, 
    y = hwy, 
    color = class)) + 
  facet_wrap(~ class, nrow = 2)
```

:::
:::

```{webr}
#| exercise: facet-wrapped-plot
#| solution: true
mpg |> 
  ggplot() + 
  geom_point(aes(
    x = displ, 
    y = hwy, 
    color = class)) + 
  facet_wrap(~ class, nrow = 2)
```

```{webr}
#| exercise: facet-wrapped-plot
#| check: true

gradethis::grade_this_code()
```

:::{.callout-caution}
## Check Your Understanding: Investigating Arguments

What does setting the argument `nrow = 2` do? Don't guess -- think about how you can determine what the nrow argument here does.

```{ojs}
//| echo: false

viewof q10 = Inputs.radio(
  new Map([
    ["It is impossible to tell", 1],
    ["It tells the ggplot to only use half the data so that plots are less cluttered", 2],
    ["It tells R to try arranging the plots into two rows", 3],
    ["It tells R to include two plots per row", 4],
    ["Nothing -- it is irrelevant and ignored", 5]
  ])
);

ok_response(q10, "3");
```

:::

## Categorical Variables

We've seen plots for categorical variables earlier in this workbook. Let's revisit a barplot with fill, similar to the one we constructed between `class` and `drv` earlier and check two alternative plots which give different insights. We'll explore the manual versus automatic transmissions.

```{r}
#| echo: true
#| eval: false
mpg |>
  count(trans)
```

```{r}
#| echo: false
#| eval: true
mpg |>
  count(trans) |> 
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Unfortunately the `trans` variable contains more detailed transmission information than what we are after. Omitting the last four characters from levels in the `trans` column would leave us with the information we want. While this sort of data wrangling is beyond the scope of these notebooks (and you won't be asked to do it on your own), I'll still show you how we can remove those last four characters below.

We can achieve our objective using the `str_sub()` function -- this function *subsets a string*. You won't need this function for our course but, because we're using it here, I'll explain how it works. The first argument is the name of the column whose strings we want to subset, the second argument is the position of the first character we'd like to keep and the third argument is the position of the final character we'd like to keep. Using a negative number here is shorthand for "from the end". That is, we'd like to keep all of the characters starting with the first and stopping five characters from the end of the string.

Run the code below to see the plot.

```{webr}
#| exercise: engineer-trans2
mpg |>
  mutate(trans2 = str_sub(trans, 1, -5)) |>
  ggplot() + 
  geom_bar(aes(x = trans2)) +
  labs(x = "Transmission Type", 
       y = "Count",
       title = "Distribution of Transmission Types")
```

From the plot above we can tell that there are about twice as many vehicles in our dataset with automatic transmissions as there are with manual transmissions. Create a similar plot below, which fills the bars by the `drv` variable.

```{webr trans2-fill-class, exercise = TRUE}
#| exercise: trans2-fill-class
______
```

:::{.hint exercise="trans2-fill-class"}
:::{.callout-note collapse="false"}
## Hint 1

```{r}
#| echo: true
#| eval: false
#Copy and paste the plotting code above and 
#use it as a starting point.

```

:::
:::

:::{.hint exercise="trans2-fill-class"}
:::{.callout-note collapse="false"}
## Hint 2

```{r}
#| echo: true
#| eval: false
mpg |>
  mutate(trans2 = str_sub(trans, 1, -5)) |>
  ggplot() + 
  geom_bar(aes(x = trans2)) +
  labs(x = "Transmission Type", 
       y = "Count",
       title = "Distribution of Transmission Types")

#Do you remember how to add a fill aesthetic?
```

:::
:::

:::{.hint exercise="trans2-fill-class"}
:::{.callout-note collapse="false"}
## Hint 3

```{r}
#| echo: true
#| eval: false
#Fill in the blank to map the drive variable 
#(drv) to the fill aesthetic.
mpg |>
  mutate(trans2 = str_sub(trans, 1, -5)) |>
  ggplot() + 
  geom_bar(aes(
    x = trans2,
    fill = ___)) +
  labs(x = "Transmission Type", 
       y = "Count",
       title = "Distribution of Transmission Types")

```

:::
:::

:::{.hint exercise="trans2-fill-class"}
:::{.callout-note collapse="false"}
## Hint 4 (Solved)

```{r}
#| echo: true
#| eval: false
mpg |>
  mutate(trans2 = str_sub(trans, 1, -5)) |>
  ggplot() + 
  geom_bar(aes(x = trans2, fill = drv)) +
  labs(x = "Transmission Type", 
       y = "Count",
       title = "Distribution of Transmission Types")
```

:::
:::

```{webr}
#| exercise: trans2-fill-class
#| solution: true
mpg |>
  mutate(trans2 = str_sub(trans, 1, -5)) |>
  ggplot() + 
  geom_bar(aes(x = trans2, fill = drv)) +
  labs(x = "Transmission Type", 
       y = "Count",
       title = "Distribution of Transmission Types")
```

```{webr}
#| exercise: trans2-fill-class
#| check: true

gradethis::grade_this_code()
```

The problem now is the difficulty in comparing the proportions of drive types within each of these classes. We can fix this by using the position argument. Outside of `aes()` but still within `mapping()` add an argument `position = "fill"` to the pre-built plot -- you'll need to include a comma after `aes()` since commas separate arguments. Think about what is gained and lost in this new plot.

```{webr}
#| exercise: trans2-drv-plot
mpg |>
  mutate(trans2 = str_sub(trans, 1, -5)) |>
  ggplot() + 
  geom_bar(aes(
    x = trans2, 
    fill = drv)) +
  labs(x = "Transmission Type", 
       y = "Count",
       title = "Distribution of Transmission Types")
```

:::{.hint exercise="trans2-drv-plot"}
:::{.callout-note collapse="false"}
## Hint 1

```{r}
#| echo: true
#| eval: false
#Since we aren't mapping a column of our 
#data set to the position parameter, we add 
#it outside of the aesthetics -- the aes() function.

```

:::
:::

:::{.hint exercise="trans2-drv-plot"}
:::{.callout-note collapse="false"}
## Hint 2

```{r}
#| echo: true
#| eval: false
#Fill in the blank to set the position to "fill"
mpg |>
  mutate(trans2 = str_sub(trans, 1, -5)) |>
  ggplot() + 
  geom_bar(aes(
    x = trans2, 
    fill = drv),
    position = ___) +
  labs(x = "Transmission Type", 
       y = "Count",
       title = "Distribution of Transmission Types")
```

:::
:::

:::{.hint exercise="trans2-drv-plot"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

```{r}
#| echo: true
#| eval: false
mpg |>
  mutate(trans2 = str_sub(trans, 1, -5)) |>
  ggplot() + 
  geom_bar(aes(
    x = trans2, 
    fill = drv), 
    position = "fill") +
  labs(x = "Transmission Type", 
       y = "Count",
       title = "Distribution of Transmission Types")
```

:::
:::

```{webr}
#| exercise: trans2-drv-plot
#| solution: true
mpg |>
  mutate(trans2 = str_sub(trans, 1, -5)) |>
  ggplot() + 
  geom_bar(aes(
    x = trans2, 
    fill = drv), 
    position = "fill") +
  labs(x = "Transmission Type", 
       y = "Count",
       title = "Distribution of Transmission Types")
```

```{webr}
#| exercise: trans2-drv-plot
#| check: true

gradethis::grade_this_code()
```

Great! But the problem now is that we've lost the idea that there are more automatic vehicles than there are manual vehicles in this dataset. As a third option, we can consider a mosaic plot.

```{r echo = TRUE, eval = TRUE}
mpg |>
  mutate(trans2 = str_sub(trans, 1, -5)) |>
  ggplot() + 
  geom_mosaic(aes(
    x = product(trans2), 
    fill = drv)) +
  labs(x = "Transmission Type", 
       y = "Drive Type",
       title = "")
```

The advantage to the mosaic plot is that it contains all of the information from the two separate plots above, but it is all combined into one plot! We've built a visualization that manages to convey that the proportion of automatic vehicles in our data set outweighs the proportion of manuals, and we can also easily compare the distribution of drive types within each of these two classes.

## Submit

:::{.callout-warning}

Grading function is not currently available for these web-based notebooks. If you would like to utilize grading functionality, then please use either the [Posit Cloud pre-built instance](https://posit.cloud/content/6328402) or [install the `{learnr}` notebooks locally](https://agmath.github.io/IntroductoryStatistics/AccessingInteractiveNotes.html).

:::

## Summary

###

Congratulations! You made it through a first discussion on data visualization. You should know, however, that this is just the "tip of the iceberg" -- there's much more to learn. I've already suggested the Data Visualization chapter of Hadley Wickham's *R for Data Science*, but he also has an entire book devoted to the `ggplot2` package in R -- check it out if you are interested in learning more! Additionally, you might follow `#TidyTuesday` on Twitter to see (and maybe even participate in) a social community on data visualization. 

###

In this workbook you learned the following:

+ Variable types dictate the kinds of plots which can be used to create effective visuals. Always consider your variable types!
+ Plotting with `ggplot()` follows a *grammar of graphics* and layered plotting structure, with "`+`" separating each plot layer. The structure of a `ggplot` is as follows:
```{r sample-ggplot-code, echo = TRUE, eval = FALSE}
data_frame_name |> 
  ggplot() + 
  geom_TYPE(aes(MAPPINGS))
```

  + Note that the `MAPPINGS` you need are determined by the geometry `TYPE` you are using.

+ There are entire courses devoted to data visualization. You've had a crash course here but there is much more to be learned. Check out the two Wickham books, use the `ggplot2` [documentation](https://ggplot2.tidyverse.org/reference/), or check out [this cheatsheet](https://posit.co/wp-content/uploads/2022/10/data-visualization-1.pdf) for more inspiration!

###

**A final note:** These interactive workbooks allow me to hide some of the setup requirements from you so that we can focus on the task at hand -- in this workbook, plotting. When you make the switch to using a standalone R instance you'll need to install and load libraries to expand R's functionality. You can do this with the `install.packages()` and `library()` functions. You would need `{ggplot2}` as well as `{dplyr}`, `{skimr}`, and `{ggmosaic}` to run the code for this workbook in a local R session.
