---
title: "Topic 5: Discrete Probability Distributions (Dynamic)"
format: 
  live-html:
    include-in-header:
      text: |
        <style>
        .btn-exercise-solution { display: none !important; }
        </style>
    webr:
      packages:
        - dplyr
        - gradethis
        - jsonlite
        - base64enc
      cell-options: 
        persist: true
    live:
      show-solutions: true
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}
{{< include ./_extensions/r-wasm/live/_gradethis.qmd >}}

```{webr}
#| edit: false
#| output: false
#| define:
#|   - ok_response

library(htmltools)
ok_response <- function(response, n) {
  if (is.na(response)) div(HTML("You haven't answered yet."), style = "color: purple")
  else if (response == n) div(HTML("Correct ✓"), style = "color: green")
  else div(HTML("Incorrect ✗"), style = "color: red")
}
```

```{ojs}
//| echo: false

function ok_checkbox(response, n) {
  if (!response || response.length === 0)
    return html`<span style="color:purple">You haven't answered yet.</span>`;
  if (response.toString() === n)
    return html`<span style="color:green">Correct ✓</span>`;
  return html`<span style="color:red">Not Yet! ✗</span>`;
}
```

```{webr}
#| edit: false
#| output: false
#| define:
#|   - normalize_checkbox_response

normalize_checkbox_response <- function(x) {
  return(cat(sort(x), sep = ","))
}
```

```{webr}
#| context: setup
#| echo: false
#| output: false
#| define:
#|   - n_students
#|   - p_assign
#|   - p_pct
#|   - thresh_atmost
#|   - thresh_atleast
#|   - thresh_between_lo
#|   - thresh_between_hi

library(jsonlite)
library(base64enc)

set.seed(NULL)

# Sample size: integer between 15 and 25
n_students <- sample(15:25, 1)

# Probability: one of several "clean" percentage values
p_assign <- sample(c(0.25, 0.28, 0.30, 0.32, 0.34, 0.36, 0.38, 0.40), 1)
p_pct <- round(p_assign * 100)

# Thresholds derived from n so they are always pedagogically sensible
thresh_atmost <- round(n_students * p_assign) + sample(1:3, 1)
thresh_atleast <- round(n_students * p_assign) + sample(4:7, 1)
thresh_atleast <- min(thresh_atleast, n_students - 1)
thresh_between_lo <- max(2, round(n_students * p_assign) - sample(2:4, 1))
thresh_between_hi <- min(n_students - 1, round(n_students * p_assign) + sample(3:6, 1))

# Initialise exercise results
exercise_results <- list(
  notebook              = "Topic 5: Discrete Probability Distributions",
  prob_a_and_b          = "unattempted",
  chem_quiz_binom       = "unattempted",
  binom_assignments_ex1 = "unattempted",
  binom_assignments_ex2 = "unattempted",
  binom_assignments_ex3 = "unattempted",
  binom_assignments_ex4 = "unattempted",
  binom_assignments_ex5 = "unattempted",
  binom_assignments_ex6 = "unattempted",
  binom_assignments_ex7 = "unattempted"
)
```

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(gradethis)
```

:::{.callout-note}
## About

This activity provides an introduction to discrete probability, including basic probability through counting outcomes, and calculating probabilities associated with outcomes of binomial experiments using the binomial distribution.
:::

## Discrete Probability Distributions

Throughout this activity you'll be introduced to the notion of probability and will explore applications of probability and discrete random variables. After developing some intuition using foundational probability ideas, we'll focus on binomial experiments and using the binomial distribution to find probabilities of prescribed outcomes.

:::{.callout-tip}
## Limitations

There are entire courses devoted to probability -- we will only cover probability to the extent that it is necessary for use in this course. If you are interested in a more detailed treatment of probability, seek out one of the many great courses available.
:::

## Objectives

**Activity Objectives:** After completing this workbook you should be able to:

+ Define, discuss, and interpret the probability of an event as its likelihood.
+ Apply fundamental counting principles and the notion of independence to compute the probability associated with the occurrence of a sequence of events.
+ Use the definition of binomial experiments to identify scenarios to which the binomial distribution can be applied.
+ Apply the binomial distribution in appropriate scenarios to find probabilities associated with specified outcomes.
+ Given a binomial experiment, compute the expected number of successful outcomes as well as the standard deviation for number of successes.

## Basic Probability

**Definition of Probability (*frequentist*):** For a given random process, the probability of an event $A$ is the proportion of time we would observe outcome $A$ if the random process were repeated an infinite number of times.

**Example:** Given a fair coin, the probability of a flip turning up heads is $0.5$ (or 50%). Similarly, given a fair six-sided die, the probability of a roll resulting in a number greater than four is $1/3$ (or about 33.3%) because there are two outcomes satisfying the criteria (rolling a 5 or rolling a 6) out of the six total possible outcomes.

**Try It!** Now it is your turn. Try the next few problems. Be sure to note any questions you have as you work through them.

:::{.callout-caution}
## Check Your Understanding: Basic Probability I

Given one fair, six-sided die, what is the probability of rolling a three?

```{ojs}
//| echo: false

mutable ok_response = (response, n) => { return html`Loading...` };
viewof q1 = Inputs.radio(
  new Map([
    ["0 / 6", 1],
    ["1 / 6", 2],
    ["2 / 6", 3],
    ["3 / 6", 4],
    ["4 / 6", 5],
    ["5 / 6", 6],
    ["6 / 6", 7]
  ]),
  {value: JSON.parse(localStorage.getItem("q1_selected") ?? "null")}
);

{
  localStorage.setItem("q1_selected", JSON.stringify(q1));
  localStorage.setItem("q1_correct", "2");
  localStorage.setItem("q1_result", q1 === null ? "unattempted" : (q1 == 2 ? "correct" : "incorrect"));
}
ok_response(q1, "2");
```

:::

:::{.callout-caution}
## Check Your Understanding: Basic Probability II

Given one fair, six-sided die, what is the probability of rolling a two, four, or six?

```{ojs}
//| echo: false

viewof q2 = Inputs.radio(
  new Map([
    ["0 / 6", 1],
    ["1 / 6", 2],
    ["2 / 6", 3],
    ["3 / 6", 4],
    ["4 / 6", 5],
    ["5 / 6", 6],
    ["6 / 6", 7]
  ]),
  {value: JSON.parse(localStorage.getItem("q2_selected") ?? "null")}
);

{
  localStorage.setItem("q2_selected", JSON.stringify(q2));
  localStorage.setItem("q2_correct", "4");
  localStorage.setItem("q2_result", q2 === null ? "unattempted" : (q2 == 4 ? "correct" : "incorrect"));
}
ok_response(q2, "4");
```

:::

:::{.callout-caution}
## Check Your Understanding: Basic Probability III

Given two fair, six-sided dice, which is larger?

```{ojs}
//| echo: false

viewof q3 = Inputs.radio(
  new Map([
    ["The probability of rolling a total of two.", 1],
    ["The probability of rolling a total of five.", 2]
  ]),
  {value: JSON.parse(localStorage.getItem("q3_selected") ?? "null")}
);

{
  localStorage.setItem("q3_selected", JSON.stringify(q3));
  localStorage.setItem("q3_correct", "2");
  localStorage.setItem("q3_result", q3 === null ? "unattempted" : (q3 == 2 ? "correct" : "incorrect"));
}
ok_response(q3, "2");
```

:::

Good work on that last set of questions. In those problems you could find the probability by counting the number of ways the desired outcome could occur and then dividing that number by the total number of outcomes possible. In the last question, there were simply more ways to roll a five (four ways to do it) than to roll a two (just one way). What if we try doing something a bit more complicated? Say we wanted to know the probability of rolling at least a two on a single roll of a die *and then* flipping a "tails" on a single flip of a coin?

:::{.callout-tip}
## Probability and Independent Events
If $A$ and $B$ are independent events (that is, the probability that $B$ occurs does not depend on whether or not $A$ occurred, and vice-versa), then the probability of $A$ and $B$ occurring is the product of the probability of $A$ occurring and the probability of $B$ occurring. Mathematically, we write: $\mathbb{P}\left[A~\text{and}~B\right] = \mathbb{P}\left[A\right]\cdot\mathbb{P}\left[B\right]$.

:::

:::{.callout-caution}
## Check Your Understanding: Probability and Independent Events I

Given a single roll of a fair, six-sided die, what is the probability of rolling at least a two?

```{ojs}
//| echo: false

viewof q4 = Inputs.radio(
  new Map([
    ["0 / 6", 1],
    ["1 / 6", 2],
    ["2 / 6", 3],
    ["3 / 6", 4],
    ["4 / 6", 5],
    ["5 / 6", 6],
    ["6 / 6", 7]
  ]),
  {value: JSON.parse(localStorage.getItem("q4_selected") ?? "null")}
);

{
  localStorage.setItem("q4_selected", JSON.stringify(q4));
  localStorage.setItem("q4_correct", "6");
  localStorage.setItem("q4_result", q4 === null ? "unattempted" : (q4 == 6 ? "correct" : "incorrect"));
}
ok_response(q4, "6");
```

:::

:::{.callout-caution}
## Check Your Understanding: Probability and Independent Events II

Given a single flip of a fair coin, what is the probability of the coin landing with *tails* facing upwards?

```{ojs}
//| echo: false

viewof q5 = Inputs.radio(
  new Map([
    ["0 / 2", 1],
    ["1 / 2", 2],
    ["2 / 2", 3]
  ]),
  {value: JSON.parse(localStorage.getItem("q5_selected") ?? "null")}
);

{
  localStorage.setItem("q5_selected", JSON.stringify(q5));
  localStorage.setItem("q5_correct", "2");
  localStorage.setItem("q5_result", q5 === null ? "unattempted" : (q5 == 2 ? "correct" : "incorrect"));
}
ok_response(q5, "2");
```

:::

:::{.callout-caution}
## Check Your Understanding: Probability and Independent Events III

Use the code block below to compute the probability that in a single roll of a fair die and a flip of a coin we observe a roll of at least two and a flip of tails.

```{webr}
#| exercise: prob-a-and-b
______
```

:::{.hint exercise="prob-a-and-b"}
:::{.callout-note collapse="false"}
## Hint 1

Use the probabilities you identified above.

:::
:::

:::{.hint exercise="prob-a-and-b"}
:::{.callout-note collapse="false"}
## Hint 2

Use the probabilities you identified above. The flip of a coin and roll of a fair die are independent events.

:::
:::

:::{.hint exercise="prob-a-and-b"}
:::{.callout-note collapse="false"}
## Hint 3

Since the events are independent, we can multiply the probabilities of the individual outcomes together.

```{r}
#| echo: true
#| eval: false

(___) * (___)
```

:::
:::

:::{.hint exercise="prob-a-and-b"}
:::{.callout-note collapse="false"}
## Hint 4 (Solved)

Since the events are independent, we can multiply the probabilities of the individual outcomes together.

```{r}
#| echo: true
#| eval: false

(5/6) * (1/2)
```

:::
:::

```{webr}
#| exercise: prob-a-and-b
#| solution: true

(5/6)*(1/2)
```

```{webr}
#| exercise: prob-a-and-b
#| check: true

exercise_results$prob_a_and_b <<- "fail"
gradethis::grade_this({
  if (abs(.result - .solution) < 0.001) {
    exercise_results$prob_a_and_b <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

:::

Good work so far. Let's say you forgot to study for your chemistry quiz today. It is a four question multiple choice quiz with answer options $a)$ through $e)$ on each question. You decide that your best option is to guess randomly on each of the questions. Answer the following, using the empty code block below to carry out any necessary computations.

```{webr}
#| exercise: calc-box
# Use this space for scratch calculations

```

:::{.hint exercise="calc-box"}
:::{.callout-note collapse="false"}
## Hint 1 (Guessing, Part I)

There are five answer options possible and only one of them is correct.

:::
:::

:::{.hint exercise="calc-box"}
:::{.callout-note collapse="false"}
## Hint 2 (Guessing, Part I)

There are five answer options possible and only one of them is correct. You're guessing randomly, so no one choice is more likely than any of the others.

:::
:::

:::{.hint exercise="calc-box"}
:::{.callout-note collapse="false"}
## Hint 3 (Guessing, Part I)

There are five answer options possible and only one of them is correct. You're guessing randomly, so no one choice is more likely than any of the others. By choosing randomly, you have a one out of five (20% or 0.20) probability of selecting the correct answer on any one question.

:::
:::

:::{.hint exercise="calc-box"}
:::{.callout-note collapse="false"}
## Hint 1 (Guessing, Part II)

The questions are independent events here.

:::
:::

:::{.hint exercise="calc-box"}
:::{.callout-note collapse="false"}
## Hint 2 (Guessing, Part II)

The questions are independent events here. Multiply the probability associated with the outcome on each individual question together.

```{r}
#| echo: true
#| eval: false
(___)*(___)*(___)*(___)
```

:::
:::

:::{.hint exercise="calc-box"}
:::{.callout-note collapse="false"}
## Hint 3 (Guessing, Part II)

The questions are independent events here. Multiply the probability associated with the outcome on each individual question together. The probability of getting any individual question correct is 0.20.

```{r}
#| echo: true
#| eval: false
(0.20)*(0.20)*(0.20)*(0.20)
```

:::
:::

:::{.hint exercise="calc-box"}
:::{.callout-note collapse="false"}
## Hint 1 (Guessing, Part III)

Use the same approach here, but now every question is being answered incorrectly.

```{r}
#| echo: true
#| eval: false
(___)*(___)*(___)*(___)
```

:::
:::

:::{.hint exercise="calc-box"}
:::{.callout-note collapse="false"}
## Hint 2 (Guessing, Part III)

If the probability of getting any individual question correct is 0.20, then the probability of getting it wrong must be 0.80.

```{r}
#| echo: true
#| eval: false
(___)*(___)*(___)*(___)
```

:::
:::

:::{.hint exercise="calc-box"}
:::{.callout-note collapse="false"}
## Hint 3 (Guessing, Part III)

If the probability of getting any individual question correct is 0.20, then the probability of getting it wrong must be 0.80.

```{r}
#| echo: true
#| eval: false
(0.80)*(0.80)*(0.80)*(0.80)
```

:::
:::

:::{.hint exercise="calc-box"}
:::{.callout-note collapse="false"}
## Hint 1 (Guessing, Part IV)

We'll start with the same setup as for the previous two parts.

```{r}
#| echo: true
#| eval: false
(___)*(___)*(___)*(___)
```

:::
:::

:::{.hint exercise="calc-box"}
:::{.callout-note collapse="false"}
## Hint 2 (Guessing, Part IV)

We don't care about the result to question 4. We could get it right or get it wrong, and it makes no difference to whether or not our even of interest occurs. 

```{r}
#| echo: true
#| eval: false
(___)*(___)*(___)*(___)
```

:::
:::

:::{.hint exercise="calc-box"}
:::{.callout-note collapse="false"}
## Hint 3 (Guessing, Part IV)

We don't care about the result to question 4. We could get it right or get it wrong, and it makes no difference to whether or not our even of interest occurs. Since, on the fourth question, our event of interest is "getting the question right or wrong", the probability of that outcome is 100% (or 1), since no other outcome is possible.

```{r}
#| echo: true
#| eval: false
(___)*(___)*(___)*(1)
```

:::
:::

:::{.hint exercise="calc-box"}
:::{.callout-note collapse="false"}
## Hint 4 (Guessing, Part IV)

If the first question that we get right is question 3, what must have been the outcome for each of the first two questions?

```{r}
#| echo: true
#| eval: false
(___)*(___)*(___)*(1)
```

:::
:::

:::{.hint exercise="calc-box"}
:::{.callout-note collapse="false"}
## Hint 4 (Guessing, Part IV)

If the first question that we get right is question 3, what must have been the outcome for each of the first two questions? We must have gotten both of the first two questions wrong. The probability of getting any individual question wrong was 0.80.

```{r}
#| echo: true
#| eval: false
(0.80)*(0.80)*(___)*(1)
```

:::
:::

:::{.hint exercise="calc-box"}
:::{.callout-note collapse="false"}
## Hint 5 (Guessing, Part IV)

We must have gotten the third question right. The probability of guessing correctly was 0.20.

```{r}
#| echo: true
#| eval: false
(0.80)*(0.80)*(0.20)*(1)
```

:::
:::

:::{.hint exercise="calc-box"}
:::{.callout-note collapse="false"}
## Hint 1 (Guessing, Part V)

Will the approach we've taken to each of the previous three questions work here? Why or why not?

```{r}
#| echo: true
#| eval: false
(___)*(___)*(___)*(___)
```

:::
:::

:::{.callout-caution}
## Check Your Understanding: Guessing on a Quiz I

For a single question, what is the probability that you get that question correct?

```{ojs}
//| echo: false

viewof q6 = Inputs.radio(
  new Map([
    ["0", 1],
    ["0.10", 2],
    ["0.20", 3],
    ["0.25", 4],
    ["0.50", 5],
    ["0.80", 6],
    ["1", 7]
  ]),
  {value: JSON.parse(localStorage.getItem("q6_selected") ?? "null")}
);

{
  localStorage.setItem("q6_selected", JSON.stringify(q6));
  localStorage.setItem("q6_correct", "3");
  localStorage.setItem("q6_result", q6 === null ? "unattempted" : (q6 == 3 ? "correct" : "incorrect"));
}
ok_response(q6, "3");
```

:::

:::{.callout-caution}
## Check Your Understanding: Guessing on a Quiz II

What is the probability that you get every one of the questions correct?

```{ojs}
//| echo: false

viewof q7 = Inputs.radio(
  new Map([
    ["0", 1],
    ["0.0016", 2],
    ["0.01", 3],
    ["0.20", 4],
    ["0.4096", 5],
    ["0.5555", 6],
    ["0.998", 7],
    ["1", 8]
  ]),
  {value: JSON.parse(localStorage.getItem("q7_selected") ?? "null")}
);

{
  localStorage.setItem("q7_selected", JSON.stringify(q7));
  localStorage.setItem("q7_correct", "2");
  localStorage.setItem("q7_result", q7 === null ? "unattempted" : (q7 == 2 ? "correct" : "incorrect"));
}
ok_response(q7, "2");
```

:::

:::{.callout-caution}
## Check Your Understanding: Guessing on a Quiz III

What is the probability that you get every one of the questions wrong?

```{ojs}
//| echo: false

viewof q8 = Inputs.radio(
  new Map([
    ["0", 1],
    ["0.0016", 2],
    ["0.01", 3],
    ["0.20", 4],
    ["0.4096", 5],
    ["0.5555", 6],
    ["0.998", 7],
    ["1", 8]
  ]),
  {value: JSON.parse(localStorage.getItem("q8_selected") ?? "null")}
);

{
  localStorage.setItem("q8_selected", JSON.stringify(q8));
  localStorage.setItem("q8_correct", "5");
  localStorage.setItem("q8_result", q8 === null ? "unattempted" : (q8 == 5 ? "correct" : "incorrect"));
}
ok_response(q8, "5");
```

:::

:::{.callout-caution}
## Check Your Understanding: Guessing on a Quiz IV

What is the probability that the first one you get wrong is question three?

```{ojs}
//| echo: false

viewof q9 = Inputs.radio(
  new Map([
    ["0", 1],
    ["0.015", 2],
    ["0.032", 3],
    ["0.0064", 4],
    ["0.20", 5],
    ["1.2", 6]
  ]),
  {value: JSON.parse(localStorage.getItem("q9_selected") ?? "null")}
);

{
  localStorage.setItem("q9_selected", JSON.stringify(q9));
  localStorage.setItem("q9_correct", "3");
  localStorage.setItem("q9_result", q9 === null ? "unattempted" : (q9 == 3 ? "correct" : "incorrect"));
}
ok_response(q9, "3");
```

:::

:::{.callout-caution}
## Check Your Understanding: Guessing on a Quiz V

What is the probability that you get exactly two questions right?

```{ojs}
//| echo: false

viewof q10 = Inputs.radio(
  new Map([
    ["0", 1],
    ["0.04", 2],
    ["0.08", 3],
    ["0.256", 4],
    ["0.20", 5],
    ["0.40", 6],
    ["0.64", 7],
    ["None of These", 8]
  ]),
  {value: JSON.parse(localStorage.getItem("q10_selected") ?? "null")}
);

{
  localStorage.setItem("q10_selected", JSON.stringify(q10));
  localStorage.setItem("q10_correct", "8");
  localStorage.setItem("q10_result", q10 === null ? "unattempted" : (q10 == 8 ? "correct" : "incorrect"));
}
ok_response(q10, "8");
```

:::

So in the last question, none of the choices were correct -- but why? There are lots of ways that we could get two of the questions right. We could get the first two right, the first and last right, the middle two right, and more! We need to account for all of these possibilities.

## Binomial Experiments and the Binomial Distribution

**Binomial Experiments:** A binomial experiment satisfies each of the following three criteria:

+ There are $n$ repeated trials.
+ Each trial has two possible outcomes (usually called *success* and *failure* for convenience).
+ The trials are independent of one another. That is, for each trial, the probability of success is $p$ (which remains constant).

**Binomial Distribution:** Let $X$ be the number of successes resulting from a binomial experiment with $n$ trials. We can compute the following probabilities:

+ The probability of <u>exactly</u> $k$ successes is given by
<center> $\displaystyle{\mathbb{P}\left[X = k\right] = \binom{n}{k}\cdot p^k\left(1 - p\right)^{n-k} \approx \tt{dbinom(k, n, p)}}$</center>
+ The probability of <u>at most</u> $k$ successes is given by
<center> $\displaystyle{\mathbb{P}\left[X \leq k\right] = \sum_{i=0}^{k}{\binom{n}{i}\cdot p^i\left(1 - p\right)^{n-i}} \approx \tt{pbinom(k, n, p)}}$</center>

In the equations above, $\binom{n}{k} = \frac{n!}{k!\left(n-k\right)!}$ counts the number of ways to arrange the $k$ successes amongst the $n$ trials. That being said, the R functionality, `dbinom()` and `pbinom()` allow us to bypass the messy formulas -- but you'll still need to know what these functions do in order to use them correctly!

:::{.callout-tip}
## Tip: Binomial Distribution

We need to use the binomial distribution to find probabilities associated with numbers of successful (or failing) outcomes in which *we do not know for certain the trials on which the successes (or failures) occur*.
:::

The code block below is set up to find the probability of exactly two flips of a coin landing heads-up out of seven total flips. Edit the code block so that it finds the probability that you got exactly two of the four questions on your chemistry quiz from earlier correct. As a reminder, there were five answer options for each question and you were guessing randomly.

```{webr}
#| exercise: chem-quiz-binom

dbinom(2, 7, 0.5)
```

:::{.hint exercise="chem-quiz-binom"}
:::{.callout-note collapse="false"}
## Hint 1

The arguments to `dbinom()` are, in order:

+ number of successful outcomes
+ total number of trials
+ probability of a successful outcome on one trial

:::
:::

:::{.hint exercise="chem-quiz-binom"}
:::{.callout-note collapse="false"}
## Hint 2

You are interested in two successful outcomes, so the first argument doesn't need changing. The other two will need to be changed though.

:::
:::

:::{.hint exercise="chem-quiz-binom"}
:::{.callout-note collapse="false"}
## Hint 3

For the second argument, how many questions are on the quiz?

```{r}
#| echo: true
#| eval: false

dbinom(2, ___, ___)
```

:::
:::

:::{.hint exercise="chem-quiz-binom"}
:::{.callout-note collapse="false"}
## Hint 4

For the third argument, what is the probability of guessing correctly on a single question?

```{r}
#| echo: true
#| eval: false

dbinom(2, 4, ___)
```

:::
:::

:::{.hint exercise="chem-quiz-binom"}
:::{.callout-note collapse="false"}
## Hint 5 (Solved)

Since each question has five possible choices, one of which is correct, the probability of guessing a correct response on one question is 0.2.

```{r}
#| echo: true
#| eval: false

dbinom(2, 4, 0.2)
```

:::
:::

```{webr}
#| exercise: chem-quiz-binom
#| solution: true

dbinom(2, 4, 0.2)
```

```{webr}
#| exercise: chem-quiz-binom
#| check: true

exercise_results$chem_quiz_binom <<- "fail"
gradethis::grade_this({
  if (abs(.result - .solution) < 0.001) {
    exercise_results$chem_quiz_binom <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

Good work. Now you'll get to try a few more problems! As you work through the next set of questions, you may want to check out [this example and solution](www/BinomialExample.pdf){target="_blank"}. Note that in that document, I mention that drawing a simple picture for each problem will help you decide which function(s) you might use and whether you might need to make multiple computations. This is a really important strategy that will help you develop an approach to solve each problem.

**Practice:** For each of the following, consider a scenario in which a random sample of `{ojs} n_students` students is asked (in private) whether they've failed to hand in at least one assignment this semester. We assume that about `{ojs} p_pct`% of students fail to hand in at least one assignment.

1. Given a single, randomly chosen student, what is the probability that the student will have failed to hand in at least one assignment this semester?

```{webr}
#| exercise: binom-assignments-ex1
#| input:
#|   - p_assign
______
```

:::{.hint exercise="binom-assignments-ex1"}
:::{.callout-note collapse="false"}
## Hint 1

If we had 100 randomly chosen students, how many might you expect failed to hand in at least one assignment? How did you know that? What, mathematically, did you do?

:::
:::

:::{.hint exercise="binom-assignments-ex1"}
:::{.callout-note collapse="false"}
## Hint 2

The percentage given in the problem statement directly tells you the probability for a single randomly selected student. Convert `{ojs} p_pct`% to a decimal.

:::
:::

:::{.hint exercise="binom-assignments-ex1"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

`{ojs} p_pct`% as a decimal is `{ojs} p_assign`.

:::
:::

```{webr}
#| exercise: binom-assignments-ex1
#| solution: true
#| input:
#|   - p_assign

p_assign
```

```{webr}
#| exercise: binom-assignments-ex1
#| check: true

exercise_results$binom_assignments_ex1 <<- "fail"
gradethis::grade_this({
  if (.result == .solution) {
    exercise_results$binom_assignments_ex1 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

2. Find the probability that exactly 7 of the `{ojs} n_students` students have failed to hand in at least one assignment.

```{webr}
#| exercise: binom-assignments-ex2
#| input:
#|   - n_students
#|   - p_assign
______
```

:::{.hint exercise="binom-assignments-ex2"}
:::{.callout-note collapse="false"}
## Hint 1

We don't know exactly which of the students are failing to hand in at least one assignment here. We'll need a special function to account for the different combinations of students.

:::
:::

:::{.hint exercise="binom-assignments-ex2"}
:::{.callout-note collapse="false"}
## Hint 2

Which function is more appropriate, `dbinom()` or `pbinom()`?

:::
:::

:::{.hint exercise="binom-assignments-ex2"}
:::{.callout-note collapse="false"}
## Hint 3

We'll use `dbinom()` since we want the probability of *exactly* 7 students failing to hand in an assignment.

:::
:::

:::{.hint exercise="binom-assignments-ex2"}
:::{.callout-note collapse="false"}
## Hint 4

Fill in the blanks to calculate the desired probability.

```{r}
#| echo: true
#| eval: false

dbinom(___, ___, ___)
```

:::
:::

:::{.hint exercise="binom-assignments-ex2"}
:::{.callout-note collapse="false"}
## Hint 5

The first argument is the number of successes (students not handing in an assignment) we are interested in — that's 7 here.

```{r}
#| echo: true
#| eval: false

dbinom(7, ___, ___)
```

:::
:::

:::{.hint exercise="binom-assignments-ex2"}
:::{.callout-note collapse="false"}
## Hint 6

The second argument is the total number of trials. There are `{ojs} n_students` students in the sample, so fill in the second blank with `n_students`.

```{r}
#| echo: true
#| eval: false

dbinom(7, ___, ___)
```

:::
:::

:::{.hint exercise="binom-assignments-ex2"}
:::{.callout-note collapse="false"}
## Hint 7 (Solved)

The third argument is the probability of a single student failing to hand in at least one assignment, which is `{ojs} p_assign`. So the full call has the second and third blanks filled with `{ojs} n_students` and `{ojs} p_assign`, respectively.

```{r}
#| echo: true
#| eval: false

dbinom(7, ___, ___)
```

:::
:::

```{webr}
#| exercise: binom-assignments-ex2
#| solution: true
#| input:
#|   - n_students
#|   - p_assign

dbinom(7, n_students, p_assign)
```

```{webr}
#| exercise: binom-assignments-ex2
#| check: true

exercise_results$binom_assignments_ex2 <<- "fail"
gradethis::grade_this({
  if (abs(.result - .solution) < 0.0001) {
    exercise_results$binom_assignments_ex2 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

3. Find the probability that at most `{ojs} thresh_atmost` of the `{ojs} n_students` students have failed to hand in at least one assignment.

```{webr}
#| exercise: binom-assignments-ex3
#| input:
#|   - n_students
#|   - p_assign
#|   - thresh_atmost
______
```

:::{.hint exercise="binom-assignments-ex3"}
:::{.callout-note collapse="false"}
## Hint 1

Similar to the previous question, we don't know exactly which of the students are failing to hand in at least one assignment here. We'll need a special function to account for the different combinations of students.

:::
:::

:::{.hint exercise="binom-assignments-ex3"}
:::{.callout-note collapse="false"}
## Hint 2

Which function is more appropriate here, `dbinom()` or `pbinom()`?

:::
:::

:::{.hint exercise="binom-assignments-ex3"}
:::{.callout-note collapse="false"}
## Hint 3

We'll use `pbinom()` since we want the probability of *at most* a certain number of students failing to hand in an assignment.

:::
:::

:::{.hint exercise="binom-assignments-ex3"}
:::{.callout-note collapse="false"}
## Hint 4

Fill in the blanks to calculate the desired probability.

```{r}
#| echo: true
#| eval: false

pbinom(___, ___, ___)
```

:::
:::

:::{.hint exercise="binom-assignments-ex3"}
:::{.callout-note collapse="false"}
## Hint 5

For `pbinom()`, the first argument is the maximum number of successes you are willing to consider — that's `thresh_atmost` here. The second and third arguments are the number of trials and the probability of success.

```{r}
#| echo: true
#| eval: false

pbinom(___, ___, ___)
```

:::
:::

:::{.hint exercise="binom-assignments-ex3"}
:::{.callout-note collapse="false"}
## Hint 6 (Solved)

Fill in `{ojs} thresh_atmost` for the maximum number of successes, `{ojs} n_students` for the number of trials, and `{ojs} p_assign` for the probability of success.

```{r}
#| echo: true
#| eval: false

pbinom(___, ___, ___)
```

:::
:::

```{webr}
#| exercise: binom-assignments-ex3
#| solution: true
#| input:
#|   - n_students
#|   - p_assign
#|   - thresh_atmost

pbinom(thresh_atmost, n_students, p_assign)
```

```{webr}
#| exercise: binom-assignments-ex3
#| check: true

exercise_results$binom_assignments_ex3 <<- "fail"
gradethis::grade_this({
  if (abs(.result - .solution) < 0.0001) {
    exercise_results$binom_assignments_ex3 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

4. Find the probability that at least `{ojs} thresh_atleast` of the `{ojs} n_students` students have failed to hand in at least one assignment.

```{webr}
#| exercise: binom-assignments-ex4
#| input:
#|   - n_students
#|   - p_assign
#|   - thresh_atleast
______
```

:::{.hint exercise="binom-assignments-ex4"}
:::{.callout-note collapse="false"}
## Hint 1

Again, we'll need a special function because we don't know which of the students will have failed to hand in an assignment.

:::
:::

:::{.hint exercise="binom-assignments-ex4"}
:::{.callout-note collapse="false"}
## Hint 2

Unfortunately, neither `dbinom()` nor `pbinom()` are a perfect fit for this scenario. Is either one *better*-suited to it, though?

:::
:::

:::{.hint exercise="binom-assignments-ex4"}
:::{.callout-note collapse="false"}
## Hint 3

The `pbinom()` function can handle multiple possible outcomes, while `dbinom()` is most useful when we are interested in exactly one outcome.

:::
:::

:::{.hint exercise="binom-assignments-ex4"}
:::{.callout-note collapse="false"}
## Hint 4

Is there a way that we could utilize the `pbinom()` function here?

:::
:::

:::{.hint exercise="binom-assignments-ex4"}
:::{.callout-note collapse="false"}
## Hint 5

The challenge is that `pbinom()` will find the probability of *at most* some number of successes, not the probability of *at least* some number of successes.

:::
:::

:::{.hint exercise="binom-assignments-ex4"}
:::{.callout-note collapse="false"}
## Hint 6

The approach below, filling the blanks with `{ojs} thresh_atleast`, `{ojs} n_students`, and `{ojs} p_assign`, results in the wrong probability — why?

```{r}
#| echo: true
#| eval: false

pbinom(___, ___, ___)
```

:::
:::

:::{.hint exercise="binom-assignments-ex4"}
:::{.callout-note collapse="false"}
## Hint 7

Filling the blanks with `{ojs} thresh_atleast`, `{ojs} n_students`, and `{ojs} p_assign` calculates the probability of 0 through `{ojs} thresh_atleast` successes. That's not what we want. We want `{ojs} thresh_atleast` *or more*.

```{r}
#| echo: true
#| eval: false

pbinom(___, ___, ___)
```

:::
:::

:::{.hint exercise="binom-assignments-ex4"}
:::{.callout-note collapse="false"}
## Hint 8

Could we start with the probability of any (all) outcomes, and then just remove the probability of the events we don't want?

:::
:::

:::{.hint exercise="binom-assignments-ex4"}
:::{.callout-note collapse="false"}
## Hint 9

Since we only have `{ojs} n_students` students, it must be the case that somewhere between 0 and `{ojs} n_students` students fail to hand in at least one assignment. This probability is 1 (or 100%).

:::
:::

:::{.hint exercise="binom-assignments-ex4"}
:::{.callout-note collapse="false"}
## Hint 10

Fill in the blanks to answer the question.

```{r}
#| echo: true
#| eval: false

1 - pbinom(___, ___, ___)
```

:::
:::

:::{.hint exercise="binom-assignments-ex4"}
:::{.callout-note collapse="false"}
## Hint 11

Something is still wrong if we fill in the blanks with `{ojs} thresh_atleast`, `{ojs} n_students`, and `{ojs} p_assign`. What needs to be adjusted?

```{r}
#| echo: true
#| eval: false

1 - pbinom(___, ___, ___)
```

:::
:::

:::{.hint exercise="binom-assignments-ex4"}
:::{.callout-note collapse="false"}
## Hint 12

That call removes the probability of at most `{ojs} thresh_atleast` successes, leaving only the probability of *more than* `{ojs} thresh_atleast`. We want *at least* `{ojs} thresh_atleast`, which includes `{ojs} thresh_atleast` itself. Replace the first argument with `thresh_atleast - 1`.

```{r}
#| echo: true
#| eval: false

1 - pbinom(___, ___, ___)
```

:::
:::

:::{.hint exercise="binom-assignments-ex4"}
:::{.callout-note collapse="false"}
## Hint 13 (Solved)

Use `{ojs} thresh_atleast - 1` as the first argument, `{ojs} n_students` as the second, and `{ojs} p_assign` as the third.

```{r}
#| echo: true
#| eval: false

1 - pbinom(___, ___, ___)
```

:::
:::

```{webr}
#| exercise: binom-assignments-ex4
#| solution: true
#| input:
#|   - n_students
#|   - p_assign
#|   - thresh_atleast

1 - pbinom(thresh_atleast - 1, n_students, p_assign)
```

```{webr}
#| exercise: binom-assignments-ex4
#| check: true

exercise_results$binom_assignments_ex4 <<- "fail"
gradethis::grade_this({
  if (abs(.result - .solution) < 0.0001) {
    exercise_results$binom_assignments_ex4 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

5. Find the probability that between a minimum of `{ojs} thresh_between_lo` and a maximum of `{ojs} thresh_between_hi` out of the `{ojs} n_students` students have failed to hand in at least one assignment.

```{webr}
#| exercise: binom-assignments-ex5
#| input:
#|   - n_students
#|   - p_assign
#|   - thresh_between_lo
#|   - thresh_between_hi
______
```

:::{.hint exercise="binom-assignments-ex5"}
:::{.callout-note collapse="false"}
## Hint 1

Similar to the previous question, neither `dbinom()` nor `pbinom()` are perfect fits here. We should prefer `pbinom()` though, since we are interested in a collection of outcomes rather than exactly a single outcome.

:::
:::

:::{.hint exercise="binom-assignments-ex5"}
:::{.callout-note collapse="false"}
## Hint 2

The strategy we used in the previous question won't *exactly* work either. Maybe we can use a similar idea, though!

:::
:::

:::{.hint exercise="binom-assignments-ex5"}
:::{.callout-note collapse="false"}
## Hint 3

Which of the following corresponds to a probability that is definitely too big — filling the first argument of `pbinom()` with `thresh_between_lo` or with `thresh_between_hi`?

:::
:::

:::{.hint exercise="binom-assignments-ex5"}
:::{.callout-note collapse="false"}
## Hint 4

Filling the first argument with `{ojs} thresh_between_hi` results in a probability that is too large because it includes outcomes below our minimum of `{ojs} thresh_between_lo`. Can we subtract out some of the probability to get only the outcomes we care about?

:::
:::

:::{.hint exercise="binom-assignments-ex5"}
:::{.callout-note collapse="false"}
## Hint 5

Fill in the blanks:

```{r}
#| echo: true
#| eval: false

pbinom(___, ___, ___) - ___(___, ___, ___)
```

The first `pbinom()` call uses `{ojs} thresh_between_hi` as its first argument, `{ojs} n_students` as its second, and `{ojs} p_assign` as its third.

:::
:::

:::{.hint exercise="binom-assignments-ex5"}
:::{.callout-note collapse="false"}
## Hint 6

We want to remove a collection of outcomes (not just one), so use `pbinom()` for the subtracted term as well.

```{r}
#| echo: true
#| eval: false

pbinom(___, ___, ___) - pbinom(___, ___, ___)
```

:::
:::

:::{.hint exercise="binom-assignments-ex5"}
:::{.callout-note collapse="false"}
## Hint 7

Filling the blanks in the first call to `pbinom()` with `{ojs} thresh_between_hi`, `{ojs} n_students`, and `{ojs} p_assign` is a good start. If we fill in the blanks of the second call to `pbinom()` with `{ojs} thresh_between_lo`, `{ojs} n_students`, and `{ojs} p_assign`, we obtain the wrong probability though. Why is that?

```{r}
#| echo: true
#| eval: false

pbinom(___, ___, ___) - pbinom(___, ___, ___)
```

:::
:::

:::{.hint exercise="binom-assignments-ex5"}
:::{.callout-note collapse="false"}
## Hint 8

Filling in the blanks of the second call to `pbinom()` with `{ojs} thresh_between_lo`, `{ojs} n_students`, and `{ojs} p_assign` subtracts the probability of at most `{ojs} thresh_between_lo` successes, which *excludes* `{ojs} thresh_between_lo` itself from the answer. We want to *include* `{ojs} thresh_between_lo`, so replace the first argument of that second call to `pbinom()` with `{ojs} thresh_between_lo - 1`.

```{r}
#| echo: true
#| eval: false

pbinom(___, ___, ___) - pbinom(___, ___, ___)
```

:::
:::

:::{.hint exercise="binom-assignments-ex5"}
:::{.callout-note collapse="false"}
## Hint 9 (Solved)

Use `{ojs} thresh_between_hi` and `{ojs} thresh_between_lo - 1` as the first arguments of the two `pbinom()` calls, with `{ojs} n_students` and `{ojs} p_assign` as the remaining arguments in both.

```{r}
#| echo: true
#| eval: false

pbinom(___, ___, ___) - pbinom(___, ___, ___)
```

:::
:::

```{webr}
#| exercise: binom-assignments-ex5
#| solution: true
#| input:
#|   - n_students
#|   - p_assign
#|   - thresh_between_lo
#|   - thresh_between_hi

pbinom(thresh_between_hi, n_students, p_assign) - pbinom(thresh_between_lo - 1, n_students, p_assign)
```

```{webr}
#| exercise: binom-assignments-ex5
#| check: true

exercise_results$binom_assignments_ex5 <<- "fail"
gradethis::grade_this({
  if (abs(.result - .solution) < 0.0001) {
    exercise_results$binom_assignments_ex5 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

:::{.callout-important}
## Don't Memorize Approaches

In several of the previous scenarios, we needed to think about the correct "first argument" being passed to `pbinom()`. Don't try to memorize when to subtract one, when to add one, when to leave the number the same as it appeared in the problem, etc. The *language* is what matters, and there are lots of ways to express which outcomes we are most interested in. If you insist on memorizing, you'll become frustrated quickly.

Instead of memorizing, take the time to *draw a picture* to help you. Examples of what these pictures might look like can be seen in [the example and solution document](www/BinomialExample.pdf){target="_blank"}, which I pointed you to earlier.
:::

6. The expected number of successes in a binomial experiment is sometimes denoted by $\mathbb{E}\left[X\right]$ and can be computed as $\mathbb{E}\left[X\right] = n\cdot p$, where $n$ denotes the number of trials run and $p$ denotes the probability of success on a single trial. Sometimes it is convenient to think of the expected number of successes as "the mean". Use the code block below to compute the expected number of students (out of `{ojs} n_students`) who have failed to hand in at least one assignment:

```{webr}
#| exercise: binom-assignments-ex6
#| input:
#|   - n_students
#|   - p_assign
______
```

:::{.hint exercise="binom-assignments-ex6"}
:::{.callout-note collapse="false"}
## Hint 1

Use the formula $\mathbb{E}\left[X\right] = n\cdot p$ from the problem statement to compute the expected value.

:::
:::

:::{.hint exercise="binom-assignments-ex6"}
:::{.callout-note collapse="false"}
## Hint 2

$n$ is the number of trials. Here that's `{ojs} n_students` students.

:::
:::

:::{.hint exercise="binom-assignments-ex6"}
:::{.callout-note collapse="false"}
## Hint 3

$p$ is the probability of success on a single trial, which is `{ojs} p_assign` in this scenario.

:::
:::

:::{.hint exercise="binom-assignments-ex6"}
:::{.callout-note collapse="false"}
## Hint 4 (Solved)

$n \cdot p$ = `{ojs} n_students` * `{ojs} p_assign`

```{r}
#| echo: true
#| eval: false

___ * ___
```

:::
:::

```{webr}
#| exercise: binom-assignments-ex6
#| solution: true
#| input:
#|   - n_students
#|   - p_assign

n_students * p_assign
```

```{webr}
#| exercise: binom-assignments-ex6
#| check: true

exercise_results$binom_assignments_ex6 <<- "fail"
gradethis::grade_this({
  if (abs(.result - .solution) < 0.01) {
    exercise_results$binom_assignments_ex6 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

7. The standard deviation in the number of successes for a binomial experiment can also be computed. The quantity $\displaystyle{s_X = \sqrt{n\cdot p\left(1 - p\right)}}$, where $n$ denotes the number of trials run and $p$ denotes the probability of success on a single trial, is the standard deviation in number of successes. Use the code block below to compute the standard deviation in number of students who have failed to hand in at least one assignment from random samples of `{ojs} n_students` students:

```{webr}
#| exercise: binom-assignments-ex7
#| input:
#|   - n_students
#|   - p_assign
______
```

:::{.hint exercise="binom-assignments-ex7"}
:::{.callout-note collapse="false"}
## Hint 1

Use the formula from the statement of the question to compute the answer.

:::
:::

:::{.hint exercise="binom-assignments-ex7"}
:::{.callout-note collapse="false"}
## Hint 2

Recall that you can compute the square root in R using the `sqrt()` function.

:::
:::

:::{.hint exercise="binom-assignments-ex7"}
:::{.callout-note collapse="false"}
## Hint 3

For this problem, $n$ = `{ojs} n_students` and $p$ = `{ojs} p_assign`.

:::
:::

:::{.hint exercise="binom-assignments-ex7"}
:::{.callout-note collapse="false"}
## Hint 4 (Solved)

Fill in the blanks with `{ojs} n_students`, `{ojs} p_assign`, and `{ojs} p_assign` again:

```{r}
#| echo: true
#| eval: false

sqrt(___ * ___ * (1 - ___))
```

:::
:::

```{webr}
#| exercise: binom-assignments-ex7
#| solution: true
#| input:
#|   - n_students
#|   - p_assign

sqrt(n_students * p_assign * (1 - p_assign))
```

```{webr}
#| exercise: binom-assignments-ex7
#| check: true

exercise_results$binom_assignments_ex7 <<- "fail"
gradethis::grade_this({
  if (abs(.result - .solution) < 0.001) {
    exercise_results$binom_assignments_ex7 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

Be sure to write down what questions you had as you worked through these problems and to have a teacher, colleague, or tutor help clarify things for you.

## Submit

If you are part of a course with an instructor who is grading your work on these activities, please copy and submit ***both*** of the hashes below using the method your instructor has requested.

:::{.callout-warning}
## Question Hash

The hash below encodes your responses to the multiple choice questions in this activity.

```{ojs}
//| echo: false

function buildQuestionResults() {
  return {
    notebook: "Topic 5: Discrete Probability Distributions",
    type: "questions",
    timestamp: new Date().toISOString(),
    questions: {
      q1_basic_prob_1: {
        selected: q1,
        correct_answer: "2",
        result: q1 === null ? "unattempted" : (q1 == 2 ? "correct" : "incorrect")
      },
      q2_basic_prob_2: {
        selected: q2,
        correct_answer: "4",
        result: q2 === null ? "unattempted" : (q2 == 4 ? "correct" : "incorrect")
      },
      q3_basic_prob_3: {
        selected: q3,
        correct_answer: "2",
        result: q3 === null ? "unattempted" : (q3 == 2 ? "correct" : "incorrect")
      },
      q4_independent_events_1: {
        selected: q4,
        correct_answer: "6",
        result: q4 === null ? "unattempted" : (q4 == 6 ? "correct" : "incorrect")
      },
      q5_independent_events_2: {
        selected: q5,
        correct_answer: "2",
        result: q5 === null ? "unattempted" : (q5 == 2 ? "correct" : "incorrect")
      },
      q6_quiz_1: {
        selected: q6,
        correct_answer: "3",
        result: q6 === null ? "unattempted" : (q6 == 3 ? "correct" : "incorrect")
      },
      q7_quiz_2: {
        selected: q7,
        correct_answer: "2",
        result: q7 === null ? "unattempted" : (q7 == 2 ? "correct" : "incorrect")
      },
      q8_quiz_3: {
        selected: q8,
        correct_answer: "5",
        result: q8 === null ? "unattempted" : (q8 == 5 ? "correct" : "incorrect")
      },
      q9_quiz_4: {
        selected: q9,
        correct_answer: "3",
        result: q9 === null ? "unattempted" : (q9 == 3 ? "correct" : "incorrect")
      },
      q10_quiz_5: {
        selected: q10,
        correct_answer: "8",
        result: q10 === null ? "unattempted" : (q10 == 8 ? "correct" : "incorrect")
      }
    }
  };
}

function toBase64(str) {
  return btoa(unescape(encodeURIComponent(str)));
}

question_hash = {
  q1; q2; q3; q4; q5; q6; q7; q8; q9; q10;
  return toBase64(JSON.stringify(buildQuestionResults()));
}
```

```{ojs}
//| echo: false

html`<div style="font-family: monospace; font-size: 0.85em; background: #f5f5f5; padding: 12px; border-radius: 6px; word-break: break-all; border: 1px solid #ddd; user-select: all; cursor: pointer;" onclick="navigator.clipboard.writeText(this.innerText)">
  ${question_hash}
</div>
<p style="margin-top: 8px; font-size: 0.9em; color: #555;">
  Click the box to copy to clipboard.
</p>`
```

:::

:::{.callout-warning}
## Exercise Hash

Click the button below to generate your exercise submission code. This hash encodes your work on the graded code exercises in this activity.

You must have attempted the graded exercises before clicking — clicking generates a snapshot of your current results. If you have completed the activity over multiple sessions, please go back through and hit the *Run Code* button on each graded exercise before generating the hash below, to ensure your most recent results are recorded.

```{webr}
#| exercise: generate-exercise-hash
#| caption: "Generate Exercise Code"

exercise_results$timestamp <- format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
exercise_results$type <- "exercises"

hash <- base64encode(charToRaw(toJSON(exercise_results, auto_unbox = TRUE)))
cat(hash)
```

:::

## Summary

:::{.callout-tip}
## Main Takeaways

+ The **probability** of an event $A$ is a measure of its likelihood and is denoted $\mathbb{P}[A]$. Every probability must be between 0 and 1.
+ If $A$ and $B$ are **independent events**, then $\mathbb{P}[A \text{ and } B] = \mathbb{P}[A] \cdot \mathbb{P}[B]$.
+ A **binomial experiment** satisfies: (1) $n$ repeated trials, (2) each trial has two possible outcomes, and (3) trials are independent with constant probability of success $p$.
+ If $X$ counts successes in a binomial experiment with $n$ trials and success probability $p$:
  + $\mathbb{P}[X = k] \approx$ `dbinom(k, n, p)` — for **exactly** $k$ successes
  + $\mathbb{P}[X \leq k] \approx$ `pbinom(k, n, p)` — for **at most** $k$ successes
  + Draw a picture to help you see how to use `pbinom()` and/or `dbinom()` to calculate probabilities. These two functions above are sufficient to handle any binomial probability scenario — the challenge is identifying how to combine them.
+ The **expected number** of successes is $\mathbb{E}[X] = n \cdot p$.
+ The **standard deviation** of number of successes is $s_X = \sqrt{n \cdot p \cdot (1 - p)}$.
:::

:::{.callout-tip}
## Looking Ahead

The next activity introduces the **normal distribution** — a continuous probability distribution that underpins much of classical statistical inference. Our focus will be on learning to compute probabilities and percentiles from this important distribution.
:::
