---
title: "Topic 6: The Normal Distribution (Spreadsheets, Static)"
format: 
  live-html:
    include-in-header:
      text: |
        <style>
        .btn-exercise-solution { display: none !important; }
        </style>
    webr:
      packages:
        - gradethis
        - jsonlite
        - base64enc
      cell-options:
        persist: true
    live:
      show-solutions: true
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}
{{< include ./_extensions/r-wasm/live/_gradethis.qmd >}}

```{webr}
#| edit: false
#| output: false
#| define:
#|   - ok_response

library(htmltools)
ok_response <- function(response, n) {
  if (is.na(response)) div(HTML("You haven't answered yet."), style = "color: purple")
  else if (response == n) div(HTML("Correct ✓"), style = "color: green")
  else div(HTML("Incorrect ✗"), style = "color: red")
}
```

```{ojs}
//| echo: false

function ok_checkbox(response, n) {
  if (!response || response.length === 0)
    return html`<span style="color:purple">You haven't answered yet.</span>`;
  if (response.toString() === n)
    return html`<span style="color:green">Correct ✓</span>`;
  return html`<span style="color:red">Not Yet! ✗</span>`;
}
```

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(gradethis)
library(patchwork)

theme_set(theme_bw())

set.seed(241)

boundary1 <- round(runif(1, 0.75, 2), 2)
boundary2 <- round(runif(1, -2, -1.5), 2)
boundary3 <- round(runif(1, -2.5, -0.75), 2)
boundary4 <- round(runif(1, 0.75, 2.5), 2)
boundary5a <- round(runif(1, -1.5, -0.75), 2)
boundary5b <- round(runif(1, 1, 2.5), 2)
boundary6a <- round(runif(1, -1.5, -0.75), 2)
boundary6b <- round(runif(1, 1.5, 2.5), 2)
boundary7 <- round(runif(1, 1, 2.5), 2)
chickPoxSampSize <- sample(50:150, 1)

chickPox_surprised   <- floor(chickPoxSampSize * 0.9 - 3.5 * sqrt(chickPoxSampSize * 0.9 * 0.1))
chickPox_exact1      <- floor(0.9 * chickPoxSampSize)
chickPox_exact2      <- floor(0.1 * chickPoxSampSize)
chickPox_atmost      <- floor(0.75 * chickPoxSampSize)
chickPox_atleast     <- floor(0.95 * chickPoxSampSize)
chickPox_between_lo  <- floor(0.8 * chickPoxSampSize)
chickPox_between_hi  <- floor(0.9 * chickPoxSampSize)
```

```{webr}
#| context: setup
#| echo: false
#| output: false
#| define:
#|   - boundary1
#|   - boundary2
#|   - boundary3
#|   - boundary4
#|   - boundary5a
#|   - boundary5b
#|   - boundary6a
#|   - boundary6b
#|   - boundary7
#|   - chickPoxSampSize
#|   - chickPox_surprised
#|   - chickPox_exact1
#|   - chickPox_exact2
#|   - chickPox_atmost
#|   - chickPox_atleast
#|   - chickPox_between_lo
#|   - chickPox_between_hi

library(jsonlite)
library(base64enc)

set.seed(241)

boundary1 <- round(runif(1, 0.75, 2), 2)
boundary2 <- round(runif(1, -2, -1.5), 2)
boundary3 <- round(runif(1, -2.5, -0.75), 2)
boundary4 <- round(runif(1, 0.75, 2.5), 2)
boundary5a <- round(runif(1, -1.5, -0.75), 2)
boundary5b <- round(runif(1, 1, 2.5), 2)
boundary6a <- round(runif(1, -1.5, -0.75), 2)
boundary6b <- round(runif(1, 1.5, 2.5), 2)
boundary7 <- round(runif(1, 1, 2.5), 2)
chickPoxSampSize <- sample(50:150, 1)

chickPox_surprised   <- floor(chickPoxSampSize * 0.9 - 3.5 * sqrt(chickPoxSampSize * 0.9 * 0.1))
chickPox_exact1      <- floor(0.9 * chickPoxSampSize)
chickPox_exact2      <- floor(0.1 * chickPoxSampSize)
chickPox_atmost      <- floor(0.75 * chickPoxSampSize)
chickPox_atleast     <- floor(0.95 * chickPoxSampSize)
chickPox_between_lo  <- floor(0.8 * chickPoxSampSize)
chickPox_between_hi  <- floor(0.9 * chickPoxSampSize)

exercise_results <- list(
  notebook           = "Topic 6: The Normal Distribution (Spreadsheets, Static)",
  pnorm_ex1          = "unattempted",
  pnorm_ex2          = "unattempted",
  pnorm_ex3          = "unattempted",
  pnorm_ex4          = "unattempted",
  pnorm_ex5          = "unattempted",
  pnorm_ex6          = "unattempted",
  pnorm_ex7          = "unattempted",
  find_percentile    = "unattempted",
  chick_pox_exp      = "unattempted",
  chick_pox_sd       = "unattempted",
  chick_pox_exact1   = "unattempted",
  chick_pox_exact2   = "unattempted",
  chick_pox_atmost   = "unattempted",
  chick_pox_atleast  = "unattempted",
  chick_pox_between  = "unattempted",
  z_score_QR         = "unattempted",
  z_score_VR         = "unattempted",
  GRE_QR_percentile  = "unattempted",
  GRE_VR_above       = "unattempted"
)
```

:::{.callout-note}
## About

This activity covers working with normally distributed data, including the calculation of probabilities and percentiles, and introduces the notion of the z-score.

```{ojs}
//| echo: false

{
  const btn = html`<button style="
    background: #436b95;
    color: white;
    border: none;
    padding: 8px 16px;
    border-radius: 4px;
    cursor: pointer;
    font-size: 0.9em;
  ">🔄 Reset My Responses</button>`;

  btn.onclick = () => {
    if (confirm('Reset all responses for this activity? This cannot be undone.')) {
      Object.keys(localStorage)
        .filter(k => k.endsWith('_selected_t6') || k.endsWith('_correct_t6') || k.endsWith('_result_t6'))
        .forEach(k => localStorage.removeItem(k));
      location.reload();
    }
  };

  return btn;
}
```

*Note.* The button above resets multiple choice and checkbox questions. Currently, resetting answer entry cells must be done manually via hitting the *Start Over* button on each individual interactive cell.

:::

:::{.callout-note}
## Using a Spreadsheet Program
There are times throughout this activity where you'll be asked to do work inside of your favorite spreadsheet program.

+ If you are a Google Sheets user, [open a blank Google Sheet](https://sheets.new){target="_blank"}
+ If you are an Excel user, open an empty workbook on your computer or [create a blank workbook in Excel online](https://office.live.com/start/Excel.aspx)
:::

## The Normal Distribution

Throughout this activity we'll investigate the probability distribution that is most central to our study of statistics: the normal distribution. If we are confident that our data are nearly normal, that opens the door for us to leverage many powerful statistical methods. This activity gives you practice working with normally distributed data.

## Objectives

**Workbook Objectives:** After completing this workbook you should be able to:

+ Compute probabilities of events from populations well-modeled by a normal distribution.
+ Given a variable $X$ which follows an assumed normal distribution, compute and interpret various percentile thresholds for $X$.
+ Identify scenarios to which the normal or binomial distributions can be applied, and use the appropriate distribution to answer various probability-related questions.

## The Normal Distribution

**Definition:** If a random variable $X$ is normally distributed with mean $\mu$ and standard deviation $\sigma$, we often write $X\sim N\left(\mu, \sigma\right)$. Three different normal distributions appear below.

+ In **<span style="color:blue">blue</span>** is a normal distribution with $\mu = 0$ and $\sigma = 5$
+ In **<span style="color:red">red</span>** is a normal distribution with $\mu = 0$ and $\sigma = 0.5$
+ In **black** is a normal distribution with $\mu = 0$ and $\sigma = 1$ (the so-called *Standard Normal Distribution*)

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false

ggplot() + 
  geom_line(mapping = aes(x = seq(-15, 15, length.out = 200), 
                          y = dnorm(seq(-15, 15, length.out = 200), 0, 5)), 
            color = "blue", lwd = 1.25) + 
  geom_line(mapping = aes(x = seq(-15, 15, length.out = 200), 
                          y = dnorm(seq(-15, 15, length.out = 200), 0, 0.5)), 
            color = "red", lwd = 1.25) + 
  geom_line(mapping = aes(x = seq(-15, 15, length.out = 200), 
                          y = dnorm(seq(-15, 15, length.out = 200), 0, 1)), 
            color = "black", lwd = 1.25) + 
  labs(title = "Three Normal Distributions", x = "", y = "")
```

Notice that all three distributions are bell-shaped and are centered at their mean ($\mu = 0$). The larger the standard deviation, the shorter and wider the curve, while the smaller the standard deviation, the taller and more narrow the curve.

Given that $X\sim N\left(\mu, \sigma\right)$, we can compute probabilities associated with observed values of $X$ by finding the corresponding area beneath the normal curve having mean $\mu$ and standard deviation $\sigma$.

**Properties of the Normal Distribution:** Consider $X\sim N\left(\mu, \sigma\right)$.

+ The area beneath the entire distribution is 1 (since this is equivalent to the probability that $X$ takes on any of its possible values).

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| fig-width: 3
#| fig-height: 3
#| fig-align: center

ggplot(NULL, aes(c(-3, 3))) +
  geom_area(stat = "function", fun = dnorm, 
            args = list(mean = 0, sd = 1), 
            xlim = c(-3, 3), fill = "steelblue", alpha = 0.5) + 
  geom_line(mapping = aes(x = seq(-3, 3, length.out = 200), 
                          y = dnorm(seq(-3, 3, length.out = 200), 0, 1)), 
            color = "blue", lwd = 1.25) + 
  xlab("") + ylab("") + 
  ggtitle("The shaded area is 1") + 
  scale_x_continuous(breaks = c(0), labels = c(expression(mu))) +
  theme(axis.text.x = element_text(size = 14))
```

+ $\displaystyle{\mathbb{P}\left[X\leq \mu\right] = \mathbb{P}\left[X\geq \mu\right] = 0.5}$ (the area underneath a full half of the distribution is 0.5)

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| fig-width: 6
#| fig-height: 3
#| fig-align: center

p_left <- ggplot(NULL, aes(c(-3, 3))) +
  geom_area(stat = "function", fun = dnorm, 
            args = list(mean = 0, sd = 1), 
            xlim = c(-3, 0), fill = "steelblue", alpha = 0.5) + 
  geom_line(mapping = aes(x = seq(-3, 3, length.out = 200), 
                          y = dnorm(seq(-3, 3, length.out = 200), 0, 1)), 
            color = "blue", lwd = 1.25) + 
  xlab("") + ylab("") + 
  ggtitle("The shaded area is 0.5") + 
  scale_x_continuous(breaks = c(0), labels = c(expression(mu))) +
  theme(axis.text.x = element_text(size = 14))

p_right <- ggplot(NULL, aes(c(-3, 3))) +
  geom_area(stat = "function", fun = dnorm, 
            args = list(mean = 0, sd = 1), 
            xlim = c(0, 3), fill = "steelblue", alpha = 0.5) + 
  geom_line(mapping = aes(x = seq(-3, 3, length.out = 200), 
                          y = dnorm(seq(-3, 3, length.out = 200), 0, 1)), 
            color = "blue", lwd = 1.25) + 
  xlab("") + ylab("") + 
  ggtitle("The shaded area is also 0.5") + 
  scale_x_continuous(breaks = c(0), labels = c(expression(mu))) +
  theme(axis.text.x = element_text(size = 14))

p_left + p_right
```

+ The distribution is symmetric. In symbols, we write $\mathbb{P}\left[X\leq \mu - k\right] = \mathbb{P}\left[X \geq \mu + k\right]$ for any $k$. Geometrically, this means that corresponding lower-tail (left-tail) and upper-tail (right-tail) areas are equal.

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| fig-width: 6
#| fig-height: 3
#| fig-align: center

p_sym_left <- ggplot(NULL, aes(c(-3, 3))) +
  geom_area(stat = "function", fun = dnorm, 
            args = list(mean = 0, sd = 1), 
            xlim = c(-3, -1.25), fill = "steelblue", alpha = 0.5) + 
  geom_line(mapping = aes(x = seq(-3, 3, length.out = 200), 
                          y = dnorm(seq(-3, 3, length.out = 200), 0, 1)), 
            color = "blue", lwd = 1.25) + 
  xlab("") + ylab("") + 
  ggtitle("A shaded area in the left tail") + 
  scale_x_continuous(breaks = c(-1.25, 0), 
                     labels = c(expression(mu - k), expression(mu))) +
  theme(axis.text.x = element_text(size = 14))

p_sym_right <- ggplot(NULL, aes(c(-3, 3))) +
  geom_area(stat = "function", fun = dnorm, 
            args = list(mean = 0, sd = 1), 
            xlim = c(1.25, 3), fill = "steelblue", alpha = 0.5) + 
  geom_line(mapping = aes(x = seq(-3, 3, length.out = 200), 
                          y = dnorm(seq(-3, 3, length.out = 200), 0, 1)), 
            color = "blue", lwd = 1.25) + 
  xlab("") + ylab("") + 
  ggtitle("The same shaded area in the right tail") + 
  scale_x_continuous(breaks = c(0, 1.25), 
                     labels = c(expression(mu), expression(mu + k))) +
  theme(axis.text.x = element_text(size = 14))

p_sym_left + p_sym_right
```

+ $\displaystyle{\mathbb{P}\left[X = k\right] = 0}$ (the probability that $X$ takes on any prescribed value exactly is $0$)

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| fig-width: 3
#| fig-height: 3
#| fig-align: center

ggplot(NULL, aes(c(-3, 3))) +
  geom_area(stat = "function", fun = dnorm, 
            args = list(mean = 0, sd = 1), 
            xlim = c(0.98, 1.02), fill = "steelblue", alpha = 0.5) + 
  geom_line(mapping = aes(x = seq(-3, 3, length.out = 200), 
                          y = dnorm(seq(-3, 3, length.out = 200), 0, 1)), 
            color = "blue", lwd = 1.25) + 
  xlab("") + ylab("") + 
  ggtitle("The area is 0") + 
  ylim(c(0, 0.42)) +
  scale_x_continuous(breaks = c(0, 1), labels = c(expression(mu), "k")) +
  theme(axis.text.x = element_text(size = 14))
```

:::{.callout-important}
Unlike the binomial distribution, where the distinction between *at least* and *more than* required careful adjustment, the probability that a continuous variable $X$ takes on any prescribed value exactly is $0$ (that is, $\mathbb{P}\left[X = k\right] = 0$). This means there is no difference between $\mathbb{P}\left[X \leq k\right]$ and $\mathbb{P}\left[X < k\right]$. Similarly, there is no difference between $\mathbb{P}\left[X \geq k\right]$ and $\mathbb{P}\left[X > k\right]$.
:::

Sometimes it is useful to be able to estimate probabilities or to estimate the proportion of a population that falls into a range, as long as the population is nearly normal. A convenient set of rough estimates come from the *Empirical Rule*.

:::{.callout-tip}
## The Empirical Rule

If $X\sim N\left(\mu, \sigma\right)$, then

+ $\mathbb{P}\left[\mu - \sigma \leq X\leq \mu + \sigma\right] \approx 0.67$ — that is, about 67% of observations lie within one standard deviation of the mean.
+ $\mathbb{P}\left[\mu - 2\sigma \leq X\leq \mu + 2\sigma\right] \approx 0.95$ — that is, about 95% of observations lie within two standard deviations of the mean.
+ $\mathbb{P}\left[\mu - 3\sigma \leq X\leq \mu + 3\sigma\right] \approx 0.997$ — that is, about 99.7% of observations lie within three standard deviations of the mean.
:::

For each of the following, assume that $X\sim N\left(\mu = 85, \sigma = 5\right)$.

:::{.callout-caution}
## Check Your Understanding: The Empirical Rule I

Use the *Empirical Rule* to approximate $\mathbb{P}\left[80\leq X\leq 90\right]$.

```{ojs}
//| echo: false

mutable ok_response = (response, n) => { return html`Loading...` };
viewof q1 = Inputs.radio(
  new Map([
    ["50%", 1],
    ["67%", 2],
    ["95%", 3],
    ["Nearly 100%", 4],
    ["It is impossible to tell", 5]
  ]),
  {value: JSON.parse(localStorage.getItem("q1_selected_t6") ?? "null")}
);

{
  localStorage.setItem("q1_selected_t6", JSON.stringify(q1));
  localStorage.setItem("q1_correct_t6", "2");
  localStorage.setItem("q1_result_t6", q1 === null ? "unattempted" : (q1 == 2 ? "correct" : "incorrect"));
}
ok_response(q1, "2");
```

:::

:::{.callout-caution}
## Check Your Understanding: The Empirical Rule II

According to the *Empirical Rule*, which of the following are boundary values for which we expect about 95% of observed values of $X$ to fall between?

```{ojs}
//| echo: false

viewof q2 = Inputs.checkbox(
  new Map([
    ["0", 1],
    ["50", 2],
    ["75", 3],
    ["80", 4],
    ["90", 5],
    ["95", 6],
    ["100", 7]
  ]),
  {value: JSON.parse(localStorage.getItem("q2_selected_t6") ?? "[]") ?? []}
);

{
  localStorage.setItem("q2_selected_t6", JSON.stringify(q2));
  localStorage.setItem("q2_correct_t6", "3,6");
  localStorage.setItem("q2_result_t6", (!q2 || q2.length === 0) ? "unattempted" : (q2.toString() === "3,6" ? "correct" : "incorrect"));
}
ok_checkbox(q2, "3,6");
```

:::

## Standardization and $z$-Scores

**Scenario:** Two students, Bob and Sally, are trying to compare how well they did on a college entrance exam. The difficulty comes in that Bob took the SAT which is known to follow an approximate normal distribution with a mean score of 1068 and a standard deviation of 210, while Sally took the ACT which also follows an approximately normal distribution but with a mean score of 20.8 and a standard deviation of 5.8. If Bob scored a 1400 on the SAT and Sally scored a 31 on the ACT, who scored relatively higher?

How do we answer this question? We'll see two methods.

**Method 1:** We can standardize the test scores so that they have comparable units.

**Definition (z-score):** If an observation $x$ comes from a nearly normal population with mean $\mu$ and standard deviation $\sigma$, then we compute the ***$z$-score*** associated with $x$ as follows:

$$\displaystyle{z = \frac{x - \mu}{\sigma}}$$

An observation's $z$-score is simply the number of standard deviations it falls above or below the mean.

**Try It!** Use your spreadsheet to compute Bob and Sally's $z$-scores and answer the questions that follow.

:::{.callout-tip}
## Computing a $z$-Score in a Spreadsheet

The $z$-score formula is simply arithmetic — no special spreadsheet function is required. In a blank cell, type `=` followed by the formula $(x - \mu) / \sigma$ using the appropriate values.

For example, to compute Bob's $z$-score, type `=(1400-1068)/210` and press Enter. Compute Sally's $z$-score the same way using her ACT score (31), the ACT mean (20.8), and the ACT standard deviation (5.8).

Be sure to label each result so you remember which $z$-score belongs to which student.
:::

:::{.callout-caution}
## Check Your Understanding: $z$-Scores I

Which of the following is the $z$-score corresponding to Bob? (round to two decimal places)

```{ojs}
//| echo: false

viewof q3 = Inputs.radio(
  new Map([
    ["1.58", 1],
    ["1394.91", 2],
    ["-1.58", 3],
    ["1.76", 4],
    ["-1.76", 5]
  ]),
  {value: JSON.parse(localStorage.getItem("q3_selected_t6") ?? "null")}
);

{
  localStorage.setItem("q3_selected_t6", JSON.stringify(q3));
  localStorage.setItem("q3_correct_t6", "1");
  localStorage.setItem("q3_result_t6", q3 === null ? "unattempted" : (q3 == 1 ? "correct" : "incorrect"));
}
ok_response(q3, "1");
```

:::

:::{.callout-caution}
## Check Your Understanding: $z$-Scores II

Which of the following is the $z$-score corresponding to Sally? (round to two decimal places)

```{ojs}
//| echo: false

viewof q4 = Inputs.radio(
  new Map([
    ["1.58", 1],
    ["27.41", 2],
    ["-1.58", 3],
    ["1.76", 4],
    ["-1.76", 5]
  ]),
  {value: JSON.parse(localStorage.getItem("q4_selected_t6") ?? "null")}
);

{
  localStorage.setItem("q4_selected_t6", JSON.stringify(q4));
  localStorage.setItem("q4_correct_t6", "4");
  localStorage.setItem("q4_result_t6", q4 === null ? "unattempted" : (q4 == 4 ? "correct" : "incorrect"));
}
ok_response(q4, "4");
```

:::

:::{.callout-caution}
## Check Your Understanding: $z$-Scores III

Who did relatively better on their standardized exam, Bob or Sally? Why?

```{ojs}
//| echo: false

viewof q5 = Inputs.radio(
  new Map([
    ["Sally, since her score is 1.76 standard deviations above the mean, while Bob's was only 1.58 standard deviations above the mean.", 1],
    ["Bob, since his z-score was 1.58 which is closer to the mean than Sally's z-score of 1.76.", 2],
    ["Bob, since his score was 1400 and Sally's score was only 31.", 3],
    ["Bob, since the SAT had a larger standard deviation it is a harder exam.", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q5_selected_t6") ?? "null")}
);

{
  localStorage.setItem("q5_selected_t6", JSON.stringify(q5));
  localStorage.setItem("q5_correct_t6", "1");
  localStorage.setItem("q5_result_t6", q5 === null ? "unattempted" : (q5 == 1 ? "correct" : "incorrect"));
}
ok_response(q5, "1");
```

:::

:::{.callout-tip}
## A Recap on $z$-Scores

We can use $z$-scores as a common unit for comparing observations from completely different populations (such as SAT scores and ACT scores). Here's a recap of the most important information so far:

+ If an observation $x$ comes from a nearly normal population with mean $\mu$ and standard deviation $\sigma$, we can compute its $z$-score using the formula: $\displaystyle{z = \frac{x - \mu}{\sigma}}$.
+ A $z$-score measures the number of standard deviations which an observation falls above or below the mean.
  + A positive $z$-score means that an observation was above the mean.
  + A negative $z$-score means that an observation was below the mean.
  + The larger a $z$-score is in absolute value, the further the corresponding observation falls from the mean.
:::

**Method 2:** We can compute the *percentile* corresponding to Bob's SAT score and the *percentile* corresponding to Sally's ACT score.

**Definition (percentile):** Given an observation $x$ from a population, the ***percentile*** corresponding to $x$ is the proportion of the population which falls at or below $x$.

In the images below, Bob's percentile is shown as the shaded area in the distribution on the left, while Sally's is shown in the distribution on the right.

::::{.columns}
:::{.column width="50%"}

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| fig-width: 3
#| fig-height: 3

ggplot(NULL, aes(c(500, 1600))) +
  geom_area(stat = "function", fun = dnorm, 
            args = list(mean = 1068, sd = 210), 
            xlim = c(500, 1400), fill = "steelblue", alpha = 0.5) + 
  geom_line(mapping = aes(x = seq(500, 1600, length.out = 200), 
                          y = dnorm(seq(500, 1600, length.out = 200), 1068, 210)), 
            color = "blue", lwd = 1.25) + 
  xlab("") + ylab("") + 
  ggtitle("SAT Distribution") + 
  xlim(c(500, 1600)) +
  scale_x_continuous(breaks = c(1068, 1400), 
                     labels = c(expression(mu == 1068), "Bob")) +
  theme(axis.text.x = element_text(size = 12))
```
:::

:::{.column width="50%"}

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| fig-width: 3
#| fig-height: 3

ggplot(NULL, aes(c(0, 36))) +
  geom_area(stat = "function", fun = dnorm, 
            args = list(mean = 20.8, sd = 5.8), 
            xlim = c(0, 31), fill = "steelblue", alpha = 0.5) + 
  geom_line(mapping = aes(x = seq(0, 36, length.out = 200), 
                          y = dnorm(seq(0, 36, length.out = 200), 20.8, 5.8)), 
            color = "blue", lwd = 1.25) + 
  xlab("") + ylab("") + 
  ggtitle("ACT Distribution") + 
  xlim(c(0, 36)) +
  scale_x_continuous(breaks = c(20.8, 31), 
                     labels = c(expression(mu == 20.8), "Sally")) +
  theme(axis.text.x = element_text(size = 12))
```
:::
::::

:::{.callout-tip}
## Computing Percentiles in a Spreadsheet

In Excel and Google Sheets, the `NORM.DIST()` function computes the area to the **left** of a given boundary value — that is, the percentile. It takes four arguments:

1. The boundary value ($q$)
2. The mean of the distribution ($\mu$)
3. The standard deviation ($\sigma$)
4. `TRUE` — this tells the function to return the cumulative (area) result rather than the height of the curve (you'll **always** use `TRUE` as the last argument to `NORM.DIST()`)

So if $X\sim N\left(\mu, \sigma\right)$, then $\mathbb{P}\left[X\leq q\right] \approx$ `=NORM.DIST(q, mean, sd, TRUE)`.

For example, Bob's percentile is calculated by typing `=NORM.DIST(1400, 1068, 210, TRUE)` into an empty cell in your spreadsheet.
:::

**Try It!** Use your spreadsheet to compute Bob's and Sally's percentiles. In a blank cell, type `=NORM.DIST(1400, 1068, 210, TRUE)` for Bob. Then adapt the formula to compute Sally's percentile (her ACT score was 31, the mean was 20.8, and the standard deviation was 5.8). Label both results.

:::{.callout-caution}
## Check Your Understanding: Percentiles I

Which of the following is the percentile corresponding to Bob? (round to four decimal places)

```{ojs}
//| echo: false

viewof q6 = Inputs.radio(
  new Map([
    ["0.9431", 1],
    ["0.0569", 2],
    ["0.4431", 3],
    ["0.9608", 4],
    ["0.0392", 5]
  ]),
  {value: JSON.parse(localStorage.getItem("q6_selected_t6") ?? "null")}
);

{
  localStorage.setItem("q6_selected_t6", JSON.stringify(q6));
  localStorage.setItem("q6_correct_t6", "1");
  localStorage.setItem("q6_result_t6", q6 === null ? "unattempted" : (q6 == 1 ? "correct" : "incorrect"));
}
ok_response(q6, "1");
```

:::

:::{.callout-caution}
## Check Your Understanding: Percentiles II

Which of the following is the percentile corresponding to Sally? (round to four decimal places)

```{ojs}
//| echo: false

viewof q7 = Inputs.radio(
  new Map([
    ["0.9431", 1],
    ["0.0569", 2],
    ["0.4431", 3],
    ["0.9607", 4],
    ["0.0392", 5]
  ]),
  {value: JSON.parse(localStorage.getItem("q7_selected_t6") ?? "null")}
);

{
  localStorage.setItem("q7_selected_t6", JSON.stringify(q7));
  localStorage.setItem("q7_correct_t6", "4");
  localStorage.setItem("q7_result_t6", q7 === null ? "unattempted" : (q7 == 4 ? "correct" : "incorrect"));
}
ok_response(q7, "4");
```

:::

:::{.callout-caution}
## Check Your Understanding: Percentiles III

Who did relatively better on their standardized exam, Bob or Sally? Why?

```{ojs}
//| echo: false

viewof q8 = Inputs.radio(
  new Map([
    ["Sally, since she scored in a higher percentile than Bob.", 1],
    ["Bob, since he scored in a lower percentile than Sally.", 2],
    ["Bob, since he scored closer to the mean.", 3],
    ["Bob, since his percentile score was so high even though the standard deviation on the SAT was so large.", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q8_selected_t6") ?? "null")}
);

{
  localStorage.setItem("q8_selected_t6", JSON.stringify(q8));
  localStorage.setItem("q8_correct_t6", "1");
  localStorage.setItem("q8_result_t6", q8 === null ? "unattempted" : (q8 == 1 ? "correct" : "incorrect"));
}
ok_response(q8, "1");
```

:::

We'll make good use of this second method for a while, but don't forget about standardization and $z$-scores — we'll need that strategy quite often later in our course! For now, let's move on to practicing with finding probabilities from a normal distribution using `NORM.DIST()` in your spreadsheet.

## Computing Probability from a Normal Distribution

Through this section you'll get practice finding probabilities by using `NORM.DIST()` to compute areas in your spreadsheet. Remember that `NORM.DIST(q, mean, sd, TRUE)` always returns the area to the **left** of the boundary value `q` in the distribution with the specified mean and standard deviation.

For these first few questions I'll draw pictures for you, but you should be prepared to draw your own shortly.

---

::::{.columns}
:::{.column width="55%"}

**Question 1:** Find $\mathbb{P}\left[Z <\right.$ `{ojs} boundary1` $\left.\right]$ — remember that $Z\sim N\left(\mu = 0, \sigma = 1\right)$.

In your spreadsheet, compute this probability and enter your result below.

```{webr}
#| exercise: pnorm-ex1
#| input:
#|   - boundary1
# Enter your answer / result here

```

:::{.hint exercise="pnorm-ex1"}
:::{.callout-note collapse="false"}
## Hint 1

Be sure you can answer: "*why does the picture to the right correspond to the probability requested?*" The shaded area is everything to the left of the boundary value — that is exactly what `NORM.DIST()` with `TRUE` computes.

:::
:::

:::{.hint exercise="pnorm-ex1"}
:::{.callout-note collapse="false"}
## Hint 2

In a blank spreadsheet cell, type `=NORM.DIST(` followed by four arguments separated by commas: the boundary value, the mean, the standard deviation, and `TRUE`. The standard normal distribution has mean 0 and standard deviation 1.

:::
:::

:::{.hint exercise="pnorm-ex1"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

In your spreadsheet, type `=NORM.DIST(___, 0, 1, TRUE)`, where the blank is filled in by `{ojs} boundary1`, and press Enter. Input the result below.

:::
:::

```{webr}
#| exercise: pnorm-ex1
#| solution: true
#| input:
#|   - boundary1

pnorm(boundary1, 0, 1)
```

```{webr}
#| exercise: pnorm-ex1
#| check: true
#| input:
#|   - boundary1

exercise_results$pnorm_ex1 <<- "fail"
gradethis::grade_this({
  if (abs(.result - pnorm(boundary1, 0, 1)) < 0.001) {
    exercise_results$pnorm_ex1 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

:::
:::{.column width="45%"}

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| fig-width: 3
#| fig-height: 3

ggplot(NULL, aes(c(-3, 3))) +
  geom_area(stat = "function", fun = dnorm, 
            args = list(mean = 0, sd = 1), 
            xlim = c(-3, boundary1), fill = "steelblue", alpha = 0.5) + 
  geom_line(mapping = aes(x = seq(-3, 3, length.out = 200), 
                          y = dnorm(seq(-3, 3, length.out = 200), 0, 1)), 
            color = "blue", lwd = 1.25) + 
  xlab("") + ylab("") +
  xlim(c(-3, 3)) +
  scale_x_continuous(breaks = c(0, boundary1)) +
  theme(axis.text.x = element_text(size = 12, angle = 45))
```

:::
::::

---

::::{.columns}
:::{.column width="55%"}

**Question 2:** Find $\mathbb{P}\left[Z <\right.$ `{ojs} boundary2` $\left.\right]$.

In your spreadsheet, compute this probability and enter your result below.

```{webr}
#| exercise: pnorm-ex2
#| input:
#|   - boundary2
# Enter your answer / result here

```

:::{.hint exercise="pnorm-ex2"}
:::{.callout-note collapse="false"}
## Hint 1

Would you be able to draw the picture if I hadn't provided it? Try practicing this picture-drawing skill — it will serve you well. 
:::
:::

:::{.hint exercise="pnorm-ex2"}
:::{.callout-note collapse="false"}
## Hint 2

How "big" do you expect this probability to be? Should it be more than 0.5 or less than 0.5?

:::
:::

:::{.hint exercise="pnorm-ex2"}
:::{.callout-note collapse="false"}
## Hint 3

Since the boundary value `{ojs} boundary2` is negative and the distribution is symmetric around 0, the shaded area covers less than half of the distribution — so the answer should be less than 0.5. `NORM.DIST()` with `TRUE` will give you the area to the left of your boundary directly.

:::
:::

:::{.hint exercise="pnorm-ex2"}
:::{.callout-note collapse="false"}
## Hint 4 (Solved)

In your spreadsheet, type `=NORM.DIST(___, 0, 1, TRUE)`, with the blank filled in by `{ojs} boundary2` and press Enter. Input the result below.

:::
:::

```{webr}
#| exercise: pnorm-ex2
#| solution: true
#| input:
#|   - boundary2

pnorm(boundary2, 0, 1)
```

```{webr}
#| exercise: pnorm-ex2
#| check: true
#| input:
#|   - boundary2

exercise_results$pnorm_ex2 <<- "fail"
gradethis::grade_this({
  if (abs(.result - pnorm(boundary2, 0, 1)) < 0.001) {
    exercise_results$pnorm_ex2 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

:::
:::{.column width="45%"}

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| fig-width: 3
#| fig-height: 3

ggplot(NULL, aes(c(-3, 3))) +
  geom_area(stat = "function", fun = dnorm, 
            args = list(mean = 0, sd = 1), 
            xlim = c(-3, boundary2), fill = "steelblue", alpha = 0.5) + 
  geom_line(mapping = aes(x = seq(-3, 3, length.out = 200), 
                          y = dnorm(seq(-3, 3, length.out = 200), 0, 1)), 
            color = "blue", lwd = 1.25) + 
  xlab("") + ylab("") +
  xlim(c(-3, 3)) +
  scale_x_continuous(breaks = c(0, boundary2)) +
  theme(axis.text.x = element_text(size = 12, angle = 45))
```

:::
::::

---

::::{.columns}
:::{.column width="55%"}

**Question 3:** Find $\mathbb{P}\left[Z >\right.$ `{ojs} boundary3` $\left.\right]$.

In your spreadsheet, compute this probability and enter your result below.

```{webr}
#| exercise: pnorm-ex3
#| input:
#|   - boundary3
# Enter your answer / result here

```

:::{.hint exercise="pnorm-ex3"}
:::{.callout-note collapse="false"}
## Hint 1

This one is different from the previous two — the shaded area is to the *right* of the boundary value. `NORM.DIST()` with `TRUE` always gives the area to the *left*, so you'll need to think about how to use it here.

:::
:::

:::{.hint exercise="pnorm-ex3"}
:::{.callout-note collapse="false"}
## Hint 2

The total area under the normal curve is 1. If you compute the area to the left of `{ojs} boundary3` and subtract it from 1, you'll be left with the area to the right — which is what you want.

:::
:::

:::{.hint exercise="pnorm-ex3"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

In your spreadsheet, type `=1-NORM.DIST(___, 0, 1, TRUE)`, where the blank is filled in by `{ojs} boundary3`, and press Enter. Input the result below.

:::
:::

```{webr}
#| exercise: pnorm-ex3
#| solution: true
#| input:
#|   - boundary3

1 - pnorm(boundary3, 0, 1)
```

```{webr}
#| exercise: pnorm-ex3
#| check: true
#| input:
#|   - boundary3

exercise_results$pnorm_ex3 <<- "fail"
gradethis::grade_this({
  if (abs(.result - (1 - pnorm(boundary3, 0, 1))) < 0.001) {
    exercise_results$pnorm_ex3 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

:::
:::{.column width="45%"}

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| fig-width: 3
#| fig-height: 3

ggplot(NULL, aes(c(-3, 3))) +
  geom_area(stat = "function", fun = dnorm, 
            args = list(mean = 0, sd = 1), 
            xlim = c(boundary3, 3), fill = "steelblue", alpha = 0.5) + 
  geom_line(mapping = aes(x = seq(-3, 3, length.out = 200), 
                          y = dnorm(seq(-3, 3, length.out = 200), 0, 1)), 
            color = "blue", lwd = 1.25) + 
  xlab("") + ylab("") +
  xlim(c(-3, 3)) +
  scale_x_continuous(breaks = c(0, boundary3)) +
  theme(axis.text.x = element_text(size = 12, angle = 45))
```

:::
::::

---

::::{.columns}
:::{.column width="55%"}

**Question 4:** Find $\mathbb{P}\left[Z >\right.$ `{ojs} boundary4` $\left.\right]$.

In your spreadsheet, compute this probability and enter your result below.

```{webr}
#| exercise: pnorm-ex4
#| input:
#|   - boundary4
# Enter your answer / result here

```

:::{.hint exercise="pnorm-ex4"}
:::{.callout-note collapse="false"}
## Hint 1

Use the same approach as Question 3 — subtract the left-tail area from 1.

:::
:::

:::{.hint exercise="pnorm-ex4"}
:::{.callout-note collapse="false"}
## Hint 2

How large do you expect the answer to be?

:::
:::

:::{.hint exercise="pnorm-ex4"}
:::{.callout-note collapse="false"}
## Hint 3

The boundary `{ojs} boundary4` is positive and to the right of the mean, so the right-tail area should be less than 0.5.

:::
:::

:::{.hint exercise="pnorm-ex4"}
:::{.callout-note collapse="false"}
## Hint 4 (Solved)

In your spreadsheet, type `=1-NORM.DIST(___, 0, 1, TRUE)`, with the blank filled in by `{ojs} boundary4` and press Enter. Input the result below.

:::
:::

```{webr}
#| exercise: pnorm-ex4
#| solution: true
#| input:
#|   - boundary4

1 - pnorm(boundary4, 0, 1)
```

```{webr}
#| exercise: pnorm-ex4
#| check: true
#| input:
#|   - boundary4

exercise_results$pnorm_ex4 <<- "fail"
gradethis::grade_this({
  if (abs(.result - (1 - pnorm(boundary4, 0, 1))) < 0.001) {
    exercise_results$pnorm_ex4 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

:::
:::{.column width="45%"}

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| fig-width: 3
#| fig-height: 3

ggplot(NULL, aes(c(-3, 3))) +
  geom_area(stat = "function", fun = dnorm, 
            args = list(mean = 0, sd = 1), 
            xlim = c(boundary4, 3), fill = "steelblue", alpha = 0.5) + 
  geom_line(mapping = aes(x = seq(-3, 3, length.out = 200), 
                          y = dnorm(seq(-3, 3, length.out = 200), 0, 1)), 
            color = "blue", lwd = 1.25) + 
  xlab("") + ylab("") +
  xlim(c(-3, 3)) +
  scale_x_continuous(breaks = c(0, boundary4)) +
  theme(axis.text.x = element_text(size = 12, angle = 45))
```

:::
::::

---

::::{.columns}
:::{.column width="55%"}

**Question 5:** Find $\mathbb{P}\left[\right.$ `{ojs} boundary5a` $< Z <$ `{ojs} boundary5b` $\left.\right]$.

In your spreadsheet, compute this probability and enter your result below.

```{webr}
#| exercise: pnorm-ex5
#| input:
#|   - boundary5a
#|   - boundary5b
# Enter your answer / result here

```

:::{.hint exercise="pnorm-ex5"}
:::{.callout-note collapse="false"}
## Hint 1

The shaded area sits between two boundary values. A single `NORM.DIST()` call won't get you there directly — you'll need two calls and combine them.

:::
:::

:::{.hint exercise="pnorm-ex5"}
:::{.callout-note collapse="false"}
## Hint 2

Start with the cumulative area up to the upper boundary ( `{ojs} boundary5b` ) — that's larger than what you want because it includes area to the left of `{ojs} boundary5a` as well. Subtract the area to the left of the lower boundary to remove the unwanted part.

:::
:::

:::{.hint exercise="pnorm-ex5"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

In your spreadsheet, type `=NORM.DIST(___, 0, 1, TRUE) - NORM.DIST(___, 0, 1, TRUE)`, where the first blank is filled in by `{ojs} boundary5b` and the second by `{ojs} boundary5a`, and press Enter. Input the result below.

:::
:::

```{webr}
#| exercise: pnorm-ex5
#| solution: true
#| input:
#|   - boundary5a
#|   - boundary5b

pnorm(boundary5b, 0, 1) - pnorm(boundary5a, 0, 1)
```

```{webr}
#| exercise: pnorm-ex5
#| check: true
#| input:
#|   - boundary5a
#|   - boundary5b

exercise_results$pnorm_ex5 <<- "fail"
gradethis::grade_this({
  if (abs(.result - (pnorm(boundary5b, 0, 1) - pnorm(boundary5a, 0, 1))) < 0.001) {
    exercise_results$pnorm_ex5 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

:::
:::{.column width="45%"}

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| fig-width: 3
#| fig-height: 3

ggplot(NULL, aes(c(-3, 3))) +
  geom_area(stat = "function", fun = dnorm, 
            args = list(mean = 0, sd = 1), 
            xlim = c(boundary5a, boundary5b), fill = "steelblue", alpha = 0.5) + 
  geom_line(mapping = aes(x = seq(-3, 3, length.out = 200), 
                          y = dnorm(seq(-3, 3, length.out = 200), 0, 1)), 
            color = "blue", lwd = 1.25) + 
  xlab("") + ylab("") +
  xlim(c(-3, 3)) +
  scale_x_continuous(breaks = c(boundary5a, 0, boundary5b)) +
  theme(axis.text.x = element_text(size = 12, angle = 45))
```

:::
::::

---

::::{.columns}
:::{.column width="55%"}

**Question 6:** Find $\mathbb{P}\left[Z <\right.$ `{ojs} boundary6a` $\text{ or } Z >$ `{ojs} boundary6b` $\left.\right]$.

In your spreadsheet, compute this probability and enter your result below.

```{webr}
#| exercise: pnorm-ex6
#| input:
#|   - boundary6a
#|   - boundary6b
# Enter your answer / result here

```

:::{.hint exercise="pnorm-ex6"}
:::{.callout-note collapse="false"}
## Hint 1

There are two separate shaded regions — a left tail and a right tail. Compute each area separately and add them together.

:::
:::

:::{.hint exercise="pnorm-ex6"}
:::{.callout-note collapse="false"}
## Hint 2

The left-tail area (below `{ojs} boundary6a`) comes directly from `NORM.DIST()` with `TRUE`. The right-tail area (above `{ojs} boundary6b`) requires the subtract-from-one approach.

:::
:::

:::{.hint exercise="pnorm-ex6"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

In your spreadsheet, type `=NORM.DIST(___, 0, 1, TRUE) + (1-NORM.DIST(___, 0, 1, TRUE))`, with the first blank filled in by `{ojs} boundary6a` and the second filled in by `{ojs} boundary6b`, and press Enter. Input the result below.

:::
:::

```{webr}
#| exercise: pnorm-ex6
#| solution: true
#| input:
#|   - boundary6a
#|   - boundary6b

pnorm(boundary6a, 0, 1) + (1 - pnorm(boundary6b, 0, 1))
```

```{webr}
#| exercise: pnorm-ex6
#| check: true
#| input:
#|   - boundary6a
#|   - boundary6b

exercise_results$pnorm_ex6 <<- "fail"
gradethis::grade_this({
  if (abs(.result - (pnorm(boundary6a, 0, 1) + (1 - pnorm(boundary6b, 0, 1)))) < 0.001) {
    exercise_results$pnorm_ex6 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

:::
:::{.column width="45%"}

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| fig-width: 3
#| fig-height: 3

ggplot(NULL, aes(c(-3, 3))) +
  geom_area(stat = "function", fun = dnorm, 
            args = list(mean = 0, sd = 1), 
            xlim = c(-3, boundary6a), fill = "steelblue", alpha = 0.5) + 
  geom_area(stat = "function", fun = dnorm, 
            args = list(mean = 0, sd = 1), 
            xlim = c(boundary6b, 3), fill = "steelblue", alpha = 0.5) + 
  geom_line(mapping = aes(x = seq(-3, 3, length.out = 200), 
                          y = dnorm(seq(-3, 3, length.out = 200), 0, 1)), 
            color = "blue", lwd = 1.25) + 
  xlab("") + ylab("") +
  xlim(c(-3, 3)) +
  scale_x_continuous(breaks = c(boundary6a, 0, boundary6b)) +
  theme(axis.text.x = element_text(size = 12, angle = 45))
```

:::
::::

---

::::{.columns}
:::{.column width="55%"}

**Question 7:** Find $\mathbb{P}\left[\left|Z\right| >\right.$ `{ojs} boundary7` $\left.\right]$.

In your spreadsheet, compute this probability and enter your result below.

```{webr}
#| exercise: pnorm-ex7
#| input:
#|   - boundary7
# Enter your answer / result here

```

:::{.hint exercise="pnorm-ex7"}
:::{.callout-note collapse="false"}
## Hint 1

Recall that $|Z|$ denotes the absolute value of $Z$, which measures distance from 0. We are interested in values of $Z$ whose distance from 0 exceeds `{ojs} boundary7` — so the shaded region is two tails, symmetric around 0.

:::
:::

:::{.hint exercise="pnorm-ex7"}
:::{.callout-note collapse="false"}
## Hint 2

You could use the same approach as Question 6. But because the distribution is symmetric, the two tail areas are equal — so you can compute just the left-tail area and double it, which is a little simpler.

:::
:::

:::{.hint exercise="pnorm-ex7"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

The left-tail boundary is `{ojs} -Number(boundary7)` (the negative version of `{ojs} boundary7`). In your spreadsheet, type `=2*NORM.DIST(___, 0, 1, TRUE)`, where the blank is filled in by `{ojs} -Number(boundary7)`, and press Enter. Input the result below.

:::
:::

```{webr}
#| exercise: pnorm-ex7
#| solution: true
#| input:
#|   - boundary7

2 * pnorm(-boundary7, 0, 1)
```

```{webr}
#| exercise: pnorm-ex7
#| check: true
#| input:
#|   - boundary7

exercise_results$pnorm_ex7 <<- "fail"
gradethis::grade_this({
  if (abs(.result - (pnorm(-boundary7, 0, 1) + (1 - pnorm(boundary7, 0, 1)))) < 0.001) {
    exercise_results$pnorm_ex7 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

:::
:::{.column width="45%"}

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| fig-width: 3
#| fig-height: 3

ggplot(NULL, aes(c(-3, 3))) +
  geom_area(stat = "function", fun = dnorm, 
            args = list(mean = 0, sd = 1), 
            xlim = c(-3, -boundary7), fill = "steelblue", alpha = 0.5) + 
  geom_area(stat = "function", fun = dnorm, 
            args = list(mean = 0, sd = 1), 
            xlim = c(boundary7, 3), fill = "steelblue", alpha = 0.5) + 
  geom_line(mapping = aes(x = seq(-3, 3, length.out = 200), 
                          y = dnorm(seq(-3, 3, length.out = 200), 0, 1)), 
            color = "blue", lwd = 1.25) + 
  xlab("") + ylab("") +
  xlim(c(-3, 3)) +
  scale_x_continuous(breaks = c(-boundary7, 0, boundary7)) +
  theme(axis.text.x = element_text(size = 12, angle = 45))
```

:::
::::

:::{.callout-tip}
## Probabilities from Other Normal Distributions

Through the last seven problems you only worked with the *standard normal distribution* — $Z\sim N\left(\mu = 0, \sigma = 1\right)$. You can find probabilities from any normal distribution using `NORM.DIST()` — just supply the appropriate mean and standard deviation as the second and third arguments instead of 0 and 1.
:::

## Finding Percentile Cutoffs on a Normal Distribution

::::{.columns}
:::{.column width="55%"}

Recall from earlier that the $p^{th}$ percentile of a random variable $X$ is the value $x^*$ such that $\mathbb{P}\left[X < x^*\right] = p$.

If $X\sim N\left(\mu, \sigma\right)$, we can find the cutoff $x^*$ for which $\mathbb{P}\left[X < x^*\right] = p$ using the `NORM.INV()` function in your spreadsheet. The `NORM.INV()` function takes three arguments: the area to the **left** of the desired cutoff ($p$), the mean, and the standard deviation.

So the $p^{th}$ percentile of $X$ is given by `=NORM.INV(p, mean, sd)`.

:::
:::{.column width="45%"}

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| fig-width: 3
#| fig-height: 3

ggplot(NULL, aes(c(-3, 3))) +
  geom_area(stat = "function", fun = dnorm, 
            args = list(mean = 0, sd = 1), 
            xlim = c(-3, 1.96), fill = "steelblue", alpha = 0.5) + 
  geom_line(mapping = aes(x = seq(-3, 3, length.out = 200), 
                          y = dnorm(seq(-3, 3, length.out = 200), 0, 1)), 
            color = "blue", lwd = 1.25) + 
  xlab("") + ylab("") +
  xlim(c(-3, 3)) +
  scale_x_continuous(breaks = c(0, 1.96), 
                     labels = c(expression(mu), "x*")) +
  theme(axis.text.x = element_text(size = 12))
```

:::
::::

Recall that SAT scores follow $N\left(\mu = 1068, \sigma = 210\right)$ and ACT scores follow $N\left(\mu = 20.8, \sigma = 5.8\right)$.

**Try It!** In your spreadsheet, use `NORM.INV()` to find the minimum SAT score required to fall in the 95th percentile. Type `=NORM.INV(0.95, 1068, 210)` and press Enter. Then adapt the formula to find the minimum ACT score required to fall into the top 10% of all ACT test takers. Enter your result for the ACT cutoff below.

```{webr}
#| exercise: find-percentile
# Enter your ACT top-10% cutoff score here

```

:::{.hint exercise="find-percentile"}
:::{.callout-note collapse="false"}
## Hint 1

The arguments to `NORM.INV()` are, in order: the area to the left of the desired cutoff, the mean of the normal distribution, and the standard deviation.

:::
:::

:::{.hint exercise="find-percentile"}
:::{.callout-note collapse="false"}
## Hint 2

To be in the top 10% of ACT scores, a student must score higher than 90% of all test takers — so the area to the left of the cutoff is 0.90. The ACT mean is 20.8 and the standard deviation is 5.8.

:::
:::

:::{.hint exercise="find-percentile"}
:::{.callout-note collapse="false"}
## Hint 3(Solved)

In your spreadsheet, type `=NORM.INV(0.90, 20.8, 5.8)` and press Enter. Input the resulting value below.

:::
:::

```{webr}
#| exercise: find-percentile
#| solution: true

qnorm(0.9, 20.8, 5.8)
```

```{webr}
#| exercise: find-percentile
#| check: true

exercise_results$find_percentile <<- "fail"
gradethis::grade_this({
  if (abs(.result - qnorm(0.9, 20.8, 5.8)) < 0.1) {
    exercise_results$find_percentile <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

## Practice with the Normal and Binomial Distributions

Through this last section you'll work through a set of problems, some of which use the normal distribution while others use the binomial distribution. It is up to you to determine which distribution should be applied in each problem. Below are a few helpful reminders:

:::{.callout-tip}
## Binomial and Normal Distributions — Spreadsheet Functions Recap

+ The **binomial distribution** can be applied to scenarios of repeated trials where each trial has two possible outcomes and the probability of "success" on each trial remains constant. If $X$ is the number of successes in a binomial experiment with $n$ trials and probability of success $p$:
  + $\mathbb{P}\left[X = k\right] \approx$ `=BINOM.DIST(k, n, p, FALSE)`
  + $\mathbb{P}\left[X \leq k\right] \approx$ `=BINOM.DIST(k, n, p, TRUE)`

+ The **normal distribution** can be applied to scenarios where data follow at least a nearly-normal distribution. If $X\sim N\left(\mu, \sigma\right)$:
  + $\mathbb{P}\left[X \leq k\right] \approx$ `=NORM.DIST(k, mean, sd, TRUE)`
  + The $p^{th}$ percentile of $X$ is given by `=NORM.INV(p, mean, sd)`
:::

---

**Practice Problem 1:** The National Vaccine Information Center estimates that 90% of Americans have had chickenpox by the time they reach adulthood. Suppose we take a random sample of `{ojs} chickPoxSampSize` American adults. Answer each of the following:

1. Compute the expected number of adults in our sample who will have had chickenpox. Enter your result below.

```{webr}
#| exercise: chick-pox-exp
#| input:
#|   - chickPoxSampSize
# Enter your answer / result here

```

:::{.hint exercise="chick-pox-exp"}
:::{.callout-note collapse="false"}
## Hint 1

Is this scenario best-modeled by a binomial distribution or a normal distribution?

:::
:::

:::{.hint exercise="chick-pox-exp"}
:::{.callout-note collapse="false"}
## Hint 2

There are `{ojs} chickPoxSampSize` adults being surveyed, they are from a random sample (independent), and each is being asked a yes/no question — "have you had chickenpox?" This fits the assumptions of a binomial experiment.

:::
:::

:::{.hint exercise="chick-pox-exp"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

The expected value of a binomial random variable is $\mathbb{E}[X] = n \cdot p$, where $n$ is the number of trials and $p$ is the probability of success. In your spreadsheet, type `= 68*0.90` and press Enter. Input the result in the code cell.

:::
:::

```{webr}
#| exercise: chick-pox-exp
#| solution: true
#| input:
#|   - chickPoxSampSize

chickPoxSampSize * 0.90
```

```{webr}
#| exercise: chick-pox-exp
#| check: true
#| input:
#|   - chickPoxSampSize

exercise_results$chick_pox_exp <<- "fail"
gradethis::grade_this({
  if (abs(.result - (chickPoxSampSize * 0.9)) < 0.1) {
    exercise_results$chick_pox_exp <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

2. Compute the standard deviation in number of adults who will have had chickenpox across samples of size `{ojs} chickPoxSampSize`. Enter your result below.

```{webr}
#| exercise: chick-pox-sd
#| input:
#|   - chickPoxSampSize
# Enter your answer / result here

```

:::{.hint exercise="chick-pox-sd"}
:::{.callout-note collapse="false"}
## Hint 1

The standard deviation of a binomial random variable is $s_X = \sqrt{n \cdot p \cdot (1 - p)}$. Here $n$ = `{ojs} chickPoxSampSize` and $p$ = 0.90. Use `SQRT()` in your spreadsheet.

:::
:::

:::{.hint exercise="chick-pox-sd"}
:::{.callout-note collapse="false"}
## Hint 2 (Solved)

In your spreadsheet, type `=SQRT(68*0.90*(1-0.90))` and press Enter. Input the resulting value.

:::
:::

```{webr}
#| exercise: chick-pox-sd
#| solution: true
#| input:
#|   - chickPoxSampSize

sqrt(chickPoxSampSize * 0.9 * 0.1)
```

```{webr}
#| exercise: chick-pox-sd
#| check: true
#| input:
#|   - chickPoxSampSize

exercise_results$chick_pox_sd <<- "fail"
gradethis::grade_this({
  if (abs(.result - sqrt(chickPoxSampSize * 0.9 * 0.1)) < 0.1) {
    exercise_results$chick_pox_sd <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

:::{.callout-caution}
## Check Your Understanding: Surprising Results?

Would you be surprised if your sample of `{ojs} chickPoxSampSize` adults contained at most `{ojs} chickPox_surprised` adults who have had chickenpox? Why? Select all that are appropriate.

```{ojs}
//| echo: false

viewof q9 = Inputs.checkbox(
  new Map([
    ["Yes. This result would fall more than two standard deviations away from the expected value.", 1],
    ["Yes. If our assumptions are correct, the probability of this occurring would be less than 5%.", 2],
    ["Yes. We should be surprised by any result at all — we can't know what will happen ahead of time.", 3],
    ["No. A result like this one is certainly possible.", 4],
    ["No. This result does not fall more than two standard deviations from the expected value.", 5]
  ]),
  {value: JSON.parse(localStorage.getItem("q9_selected_t6") ?? "[]") ?? []}
);

{
  localStorage.setItem("q9_selected_t6", JSON.stringify(q9));
  localStorage.setItem("q9_correct_t6", "1,2");
  localStorage.setItem("q9_result_t6", (!q9 || q9.length === 0) ? "unattempted" : (q9.toString() === "1,2" ? "correct" : "incorrect"));
}
ok_checkbox(q9, "1,2");
```

:::

While you answer the following questions, it might be useful to refer back to [the example and solution document](www/BinomialExample_Excel.pdf){target="_blank"} from the Topic 5 activity.

3. Find the probability that exactly `{ojs} chickPox_exact1` adults in your sample of `{ojs} chickPoxSampSize` will have had chickenpox. Enter your result below.

```{webr}
#| exercise: chick-pox-exact1
#| input:
#|   - chickPoxSampSize
#|   - chickPox_exact1
# Enter your answer / result here

```

:::{.hint exercise="chick-pox-exact1"}
:::{.callout-note collapse="false"}
## Hint 1

We are still working with a binomial scenario. Since we want *exactly* a specific number of successes, use `BINOM.DIST()` with `FALSE` as the last argument.

:::
:::

:::{.hint exercise="chick-pox-exact1"}
:::{.callout-note collapse="false"}
## Hint 2

The four arguments are: the number of successes you are interested in ( `{ojs} chickPox_exact1` ), the sample size ( `{ojs} chickPoxSampSize` ), the probability of success ( 0.90 ), and `FALSE`.

:::
:::

:::{.hint exercise="chick-pox-exact1"}
:::{.callout-note collapse="false"}
## Hint 3

In your spreadsheet, type `=BINOM.DIST(___, ___ , 0.90, FALSE)`, where the first blank is filled in by `{ojs} chickPox_exact1` and the second by `{ojs} chickPoxSampSize`, and press Enter. Input the result.

:::
:::

```{webr}
#| exercise: chick-pox-exact1
#| solution: true
#| input:
#|   - chickPoxSampSize
#|   - chickPox_exact1

dbinom(chickPox_exact1, chickPoxSampSize, 0.9)
```

```{webr}
#| exercise: chick-pox-exact1
#| check: true
#| input:
#|   - chickPoxSampSize
#|   - chickPox_exact1

exercise_results$chick_pox_exact1 <<- "fail"
gradethis::grade_this({
  if (abs(.result - dbinom(chickPox_exact1, chickPoxSampSize, 0.9)) < 0.001) {
    exercise_results$chick_pox_exact1 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

4. Find the probability that exactly `{ojs} chickPox_exact2` adults in your sample of `{ojs} chickPoxSampSize` will *not* have had chickenpox. Enter your result below.

```{webr}
#| exercise: chick-pox-exact2
#| input:
#|   - chickPoxSampSize
#|   - chickPox_exact2
# Enter your answer / result here

```

:::{.hint exercise="chick-pox-exact2"}
:::{.callout-note collapse="false"}
## Hint 1

Read carefully — this question asks about adults who have *not* had chickenpox. That changes both the count of interest and the probability argument.

:::
:::

:::{.hint exercise="chick-pox-exact2"}
:::{.callout-note collapse="false"}
## Hint 2

The probability of a single adult *not* having had chickenpox is 1 - 0.90 = 0.10. The number of adults who have not had chickenpox is `{ojs} chickPox_exact2` and the sample size is `{ojs} chickPoxSampSize`.

:::
:::

:::{.hint exercise="chick-pox-exact2"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

In your spreadsheet, type `=BINOM.DIST(___, ___, 0.10, FALSE)`, where the first blank is filled in by `{ojs} chickPox_exact2` and the second by `{ojs} chickPoxSampSize`, and press Enter. Input the result.

:::
:::

```{webr}
#| exercise: chick-pox-exact2
#| solution: true
#| input:
#|   - chickPoxSampSize
#|   - chickPox_exact2

dbinom(chickPox_exact2, chickPoxSampSize, 0.1)
```

```{webr}
#| exercise: chick-pox-exact2
#| check: true
#| input:
#|   - chickPoxSampSize
#|   - chickPox_exact2

exercise_results$chick_pox_exact2 <<- "fail"
gradethis::grade_this({
  if (abs(.result - dbinom(chickPox_exact2, chickPoxSampSize, 0.1)) < 0.001) {
    exercise_results$chick_pox_exact2 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

5. Find the probability that at most `{ojs} chickPox_atmost` adults in your sample of `{ojs} chickPoxSampSize` will have had chickenpox. Enter your result below.

```{webr}
#| exercise: chick-pox-atmost
#| input:
#|   - chickPoxSampSize
#|   - chickPox_atmost
# Enter your answer / result here

```

:::{.hint exercise="chick-pox-atmost"}
:::{.callout-note collapse="false"}
## Hint 1

"At most" covers a range of outcomes (0, 1, 2, … up to `{ojs} chickPox_atmost`). Use `BINOM.DIST()` with `TRUE` as the last argument.

:::
:::

:::{.hint exercise="chick-pox-atmost"}
:::{.callout-note collapse="false"}
## Hint 2 (Solved)

In your spreadsheet, type `=BINOM.DIST(___, ___, 0.90, TRUE)`, where the first blank is filled in by `{ojs} chickPox_atmost` and the second by `{ojs} chickPoxSampSize`, and press Enter. Input the result.

:::
:::

```{webr}
#| exercise: chick-pox-atmost
#| solution: true
#| input:
#|   - chickPoxSampSize
#|   - chickPox_atmost

pbinom(chickPox_atmost, chickPoxSampSize, 0.9)
```

```{webr}
#| exercise: chick-pox-atmost
#| check: true
#| input:
#|   - chickPoxSampSize
#|   - chickPox_atmost

exercise_results$chick_pox_atmost <<- "fail"
gradethis::grade_this({
  if (abs(.result - pbinom(chickPox_atmost, chickPoxSampSize, 0.9)) < 0.00005) {
    exercise_results$chick_pox_atmost <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

6. Find the probability that at least `{ojs} chickPox_atleast` adults in your sample of `{ojs} chickPoxSampSize` will have had chickenpox. Enter your result below.

```{webr}
#| exercise: chick-pox-atleast
#| input:
#|   - chickPoxSampSize
#|   - chickPox_atleast
# Enter your answer / result here

```

:::{.hint exercise="chick-pox-atleast"}
:::{.callout-note collapse="false"}
## Hint 1

`BINOM.DIST()` with `TRUE` computes the probability of *at most* some number of successes — not *at least*. Draw a picture to start.

:::
:::

:::{.hint exercise="chick-pox-atleast"}
:::{.callout-note collapse="false"}
## Hint 2

`BINOM.DIST()` with `TRUE` computes the probability of *at most* some number of successes — not *at least*. Draw a picture to start. You'll need the "subtract from 1" approach here.

:::
:::

:::{.hint exercise="chick-pox-atleast"}
:::{.callout-note collapse="false"}
## Hint 3

To get "at least `{ojs} chickPox_atleast`", subtract the probability of "at most `{ojs} Number(chickPox_atleast) - 1`" from 1.

:::
:::

:::{.hint exercise="chick-pox-atleast"}
:::{.callout-note collapse="false"}
## Hint 4 (Solved)

To get "at least `{ojs} chickPox_atleast`", subtract the probability of "at most `{ojs} Number(chickPox_atleast) - 1`" from 1. In your spreadsheet, type `=1-BINOM.DIST(___, ___, 0.90, TRUE)`, where the first blank is filled in by `{ojs} Number(chickPox_atleast) - 1` and the second by `{ojs} chickPoxSampSize`, and press Enter. Input the result.

:::
:::

```{webr}
#| exercise: chick-pox-atleast
#| solution: true
#| input:
#|   - chickPoxSampSize
#|   - chickPox_atleast

1 - pbinom(chickPox_atleast - 1, chickPoxSampSize, 0.9)
```

```{webr}
#| exercise: chick-pox-atleast
#| check: true
#| input:
#|   - chickPoxSampSize
#|   - chickPox_atleast

exercise_results$chick_pox_atleast <<- "fail"
gradethis::grade_this({
  if (abs(.result - (1 - pbinom(chickPox_atleast - 1, chickPoxSampSize, 0.9))) < 0.0001) {
    exercise_results$chick_pox_atleast <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

7. Find the probability that more than `{ojs} chickPox_between_lo` but less than `{ojs} chickPox_between_hi` adults in your sample of `{ojs} chickPoxSampSize` will have had chickenpox. Enter your result below.

```{webr}
#| exercise: chick-pox-between
#| input:
#|   - chickPoxSampSize
#|   - chickPox_between_lo
#|   - chickPox_between_hi
# Enter your answer / result here

```

:::{.hint exercise="chick-pox-between"}
:::{.callout-note collapse="false"}
## Hint 1

Multiple outcomes are of interest, so use `BINOM.DIST()` with `TRUE`. Draw a picture — you'll need two calls, one for the upper boundary and one to subtract out the outcomes below the lower boundary.

:::
:::

:::{.hint exercise="chick-pox-between"}
:::{.callout-note collapse="false"}
## Hint 2

The phrase "more than `{ojs} chickPox_between_lo` but less than `{ojs} chickPox_between_hi`" means both boundaries are *excluded*. The upper call covers outcomes up through `{ojs} Number(chickPox_between_hi) - 1`; the lower call removes outcomes up through `{ojs} chickPox_between_lo`.

:::
:::

:::{.hint exercise="chick-pox-between"}
:::{.callout-note collapse="false"}
## Hint 3

In your spreadsheet, type `=BINOM.DIST(___, ___, 0.90, TRUE) - BINOM.DIST(___, ___, 0.90, TRUE)`, where the second blank is fille by `{ojs} chickPoxSampSize` in both cases. What should you use in place of the first blank inside of each call to `BINOM.DIST()`?

:::
:::

:::{.hint exercise="chick-pox-between"}
:::{.callout-note collapse="false"}
## Hint 4 (Solved)

In your spreadsheet, type `=BINOM.DIST(___, ___, 0.90, TRUE) - BINOM.DIST(___, ___, 0.90, TRUE)`, where the second blank is filled in by `{ojs} chickPoxSampSize` in both cases and the first blanks are filled in by `{ojs} Number(chickPox_between_hi) - 1` and `{ojs} chickPox_between_lo`, respectively, and press Enter. Input the result.

:::
:::

```{webr}
#| exercise: chick-pox-between
#| solution: true
#| input:
#|   - chickPoxSampSize
#|   - chickPox_between_lo
#|   - chickPox_between_hi

pbinom(chickPox_between_hi - 1, chickPoxSampSize, 0.9) - pbinom(chickPox_between_lo, chickPoxSampSize, 0.9)
```

```{webr}
#| exercise: chick-pox-between
#| check: true
#| input:
#|   - chickPoxSampSize
#|   - chickPox_between_lo
#|   - chickPox_between_hi

exercise_results$chick_pox_between <<- "fail"
gradethis::grade_this({
  if (abs(.result - (pbinom(chickPox_between_hi - 1, chickPoxSampSize, 0.9) - pbinom(chickPox_between_lo, chickPoxSampSize, 0.9))) < 0.0001) {
    exercise_results$chick_pox_between <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

---

**Practice Problem 2:** Sophia took the Graduate Record Examination (GRE) and scored 160 on the Verbal Reasoning section and 157 on the Quantitative Reasoning section. The mean score on the Verbal Reasoning section for all test takers was 151 with a standard deviation of 7, and the mean score for the Quantitative Reasoning section was 153 with a standard deviation of 7.67. Suppose we can assume that both score distributions are nearly normal.

1. Use your spreadsheet to compute Sophia's $z$-score on the Quantitative Reasoning exam. Enter your result below.

```{webr}
#| exercise: z-score-QR
# Enter your answer / result here

```

:::{.hint exercise="z-score-QR"}
:::{.callout-note collapse="false"}
## Hint 1

The $z$-score formula is $(x - \mu) / \sigma$. No special function is needed — just type the arithmetic directly into a spreadsheet cell.

:::
:::

:::{.hint exercise="z-score-QR"}
:::{.callout-note collapse="false"}
## Hint 2

Sophia scored 157 on the Quantitative Reasoning exam, the mean is 153, and the standard deviation is 7.67. In your spreadsheet, type `=(157-153)/7.67` and press Enter. Enter the result below.

:::
:::

```{webr}
#| exercise: z-score-QR
#| solution: true

(157 - 153) / 7.67
```

```{webr}
#| exercise: z-score-QR
#| check: true

exercise_results$z_score_QR <<- "fail"
gradethis::grade_this({
  if (abs(.result - ((157 - 153) / 7.67)) < 0.01) {
    exercise_results$z_score_QR <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

2. Use your spreadsheet to compute Sophia's $z$-score on the Verbal Reasoning exam. Enter your result below.

```{webr}
#| exercise: z-score-VR
# Enter your answer / result here

```

:::{.hint exercise="z-score-VR"}
:::{.callout-note collapse="false"}
## Hint 1

Use the same approach as above. Sophia scored 160 on the Verbal Reasoning exam, the mean is 151, and the standard deviation is 7. In your spreadsheet, type `=(160-151)/7` and press Enter.

:::
:::

```{webr}
#| exercise: z-score-VR
#| solution: true

(160 - 151) / 7
```

```{webr}
#| exercise: z-score-VR
#| check: true

exercise_results$z_score_VR <<- "fail"
gradethis::grade_this({
  if (abs(.result - ((160 - 151) / 7)) < 0.01) {
    exercise_results$z_score_VR <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

:::{.callout-caution}
## Check Your Understanding: Comparing $z$-Scores

What do these $z$-scores tell you?

```{ojs}
//| echo: false

viewof q10 = Inputs.radio(
  new Map([
    ["Sophia scored relatively higher on the Verbal Reasoning exam since her Verbal Reasoning z-score was larger.", 1],
    ["Sophia scored relatively higher on the Quantitative Reasoning exam since her Quantitative Reasoning z-score was larger.", 2],
    ["Sophia scored relatively higher on the Quantitative Reasoning exam since her Quantitative Reasoning z-score was closer to 0.", 3],
    ["It is impossible to tell which exam Sophia performed better on since the means and standard deviations are different.", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q10_selected_t6") ?? "null")}
);

{
  localStorage.setItem("q10_selected_t6", JSON.stringify(q10));
  localStorage.setItem("q10_correct_t6", "1");
  localStorage.setItem("q10_result_t6", q10 === null ? "unattempted" : (q10 == 1 ? "correct" : "incorrect"));
}
ok_response(q10, "1");
```

:::

3. Find the proportion of test takers Sophia scored higher than on the Quantitative Reasoning exam (that is, find her percentile). Enter your result below.

```{webr}
#| exercise: GRE-QR-percentile
# Enter your answer / result here

```

:::{.hint exercise="GRE-QR-percentile"}
:::{.callout-note collapse="false"}
## Hint 1

Scores are approximately normally distributed, so use `NORM.DIST()`. We want the proportion of test takers *below* Sophia's score — that's the area to the left, which is exactly what `NORM.DIST()` with `TRUE` gives.

:::
:::

:::{.hint exercise="GRE-QR-percentile"}
:::{.callout-note collapse="false"}
## Hint 2

Sophia's Quantitative Reasoning score is 157, the mean is 153, and the standard deviation is 7.67. In your spreadsheet, type `=NORM.DIST(157, 153, 7.67, TRUE)` and press Enter. Enter the result below.

:::
:::

```{webr}
#| exercise: GRE-QR-percentile
#| solution: true

pnorm(157, 153, 7.67)
```

```{webr}
#| exercise: GRE-QR-percentile
#| check: true

exercise_results$GRE_QR_percentile <<- "fail"
gradethis::grade_this({
  if (abs(.result - pnorm(157, 153, 7.67)) < 0.001) {
    exercise_results$GRE_QR_percentile <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

4. Find the proportion of test takers doing *better than* Sophia on the Verbal Reasoning exam. Enter your result below.

```{webr}
#| exercise: GRE-VR-above
# Enter your answer / result here

```

:::{.hint exercise="GRE-VR-above"}
:::{.callout-note collapse="false"}
## Hint 1

We want the area to the *right* of Sophia's score — the proportion scoring higher than her. Use the subtract-from-one approach.

:::
:::

:::{.hint exercise="GRE-VR-above"}
:::{.callout-note collapse="false"}
## Hint 2

Sophia's Verbal Reasoning score is 160, the mean is 151, and the standard deviation is 7. In your spreadsheet, type `=1-NORM.DIST(160, 151, 7, TRUE)` and press Enter. Enter the result below.

:::
:::

```{webr}
#| exercise: GRE-VR-above
#| solution: true

1 - pnorm(160, 151, 7)
```

```{webr}
#| exercise: GRE-VR-above
#| check: true

exercise_results$GRE_VR_above <<- "fail"
gradethis::grade_this({
  if (abs(.result - (1 - pnorm(160, 151, 7))) < 0.001) {
    exercise_results$GRE_VR_above <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

## Submit

If you are part of a course with an instructor who is grading your work on these activities, please copy and submit ***both*** of the hashes below using the method your instructor has requested.

:::{.callout-warning}
## Question Hash

The hash below encodes your responses to the multiple choice and checkbox questions in this activity.

```{ojs}
//| echo: false

function buildQuestionResults() {
  return {
    notebook: "Topic 6: The Normal Distribution (Spreadsheets, Static)",
    type: "questions",
    timestamp: new Date().toISOString(),
    questions: {
      q1_empirical_rule_1: {
        selected: q1,
        correct_answer: "2",
        result: q1 === null ? "unattempted" : (q1 == 2 ? "correct" : "incorrect")
      },
      q2_empirical_rule_2: {
        selected: q2 ?? [],
        correct_answer: "3,6",
        result: (!q2 || q2.length === 0) ? "unattempted" : (q2.toString() === "3,6" ? "correct" : "incorrect")
      },
      q3_z_score_bob: {
        selected: q3,
        correct_answer: "1",
        result: q3 === null ? "unattempted" : (q3 == 1 ? "correct" : "incorrect")
      },
      q4_z_score_sally: {
        selected: q4,
        correct_answer: "4",
        result: q4 === null ? "unattempted" : (q4 == 4 ? "correct" : "incorrect")
      },
      q5_z_score_comparison: {
        selected: q5,
        correct_answer: "1",
        result: q5 === null ? "unattempted" : (q5 == 1 ? "correct" : "incorrect")
      },
      q6_percentile_bob: {
        selected: q6,
        correct_answer: "1",
        result: q6 === null ? "unattempted" : (q6 == 1 ? "correct" : "incorrect")
      },
      q7_percentile_sally: {
        selected: q7,
        correct_answer: "4",
        result: q7 === null ? "unattempted" : (q7 == 4 ? "correct" : "incorrect")
      },
      q8_percentile_comparison: {
        selected: q8,
        correct_answer: "1",
        result: q8 === null ? "unattempted" : (q8 == 1 ? "correct" : "incorrect")
      },
      q9_surprising_results: {
        selected: q9 ?? [],
        correct_answer: "1,2",
        result: (!q9 || q9.length === 0) ? "unattempted" : (q9.toString() === "1,2" ? "correct" : "incorrect")
      },
      q10_z_score_gre: {
        selected: q10,
        correct_answer: "1",
        result: q10 === null ? "unattempted" : (q10 == 1 ? "correct" : "incorrect")
      }
    }
  };
}

function toBase64(str) {
  return btoa(unescape(encodeURIComponent(str)));
}

question_hash = {
  q1; q2; q3; q4; q5; q6; q7; q8; q9; q10;
  return toBase64(JSON.stringify(buildQuestionResults()));
}
```

```{ojs}
//| echo: false

html`<div style="font-family: monospace; font-size: 0.85em; background: #f5f5f5; padding: 12px; border-radius: 6px; word-break: break-all; border: 1px solid #ddd; user-select: all; cursor: pointer;" onclick="navigator.clipboard.writeText(this.innerText)">
  ${question_hash}
</div>
<p style="margin-top: 8px; font-size: 0.9em; color: #555;">
  Click the box to copy to clipboard.
</p>`
```

:::

:::{.callout-warning}
## Exercise Hash

Click the button below to generate your exercise submission code. This hash encodes your work on the graded exercises in this activity.

You must have attempted the graded exercises before clicking — clicking generates a snapshot of your current results. If you have completed the activity over multiple sessions, please go back through and enter your answer in each graded exercise cell before generating the hash below, to ensure your most recent results are recorded.

```{webr}
#| exercise: generate-exercise-hash
#| caption: "Generate Exercise Code"

exercise_results$timestamp <- format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
exercise_results$type <- "exercises"

hash <- base64encode(charToRaw(toJSON(exercise_results, auto_unbox = TRUE)))
cat(hash)
```

:::

## Summary

:::{.callout-tip}
## Main Takeaways

+ A **normal distribution** is bell-shaped and fully described by its mean $\mu$ and standard deviation $\sigma$, written $N(\mu, \sigma)$. Larger $\sigma$ produces a shorter, wider curve; smaller $\sigma$ produces a taller, narrower curve.
+ The **Empirical Rule**: for data following $N(\mu, \sigma)$, approximately 67% of observations fall within one standard deviation of the mean, 95% within two, and 99.7% within three.
+ For a continuous distribution, $\mathbb{P}[X = k] = 0$, so there is no distinction between strict ($<$, $>$) and non-strict ($\leq$, $\geq$) inequalities when computing probabilities.
+ A **$z$-score** measures how many standard deviations an observation falls from the mean: $z = (x - \mu)/\sigma$. In a spreadsheet, compute it as `=(x-mean)/sd`. Converting raw observed values to $z$-scores allow comparison across different normal distributions on a common scale.
+ **`=NORM.DIST(q, mean, sd, TRUE)`** returns the area to the **left** of $q$ — that is, $\mathbb{P}[X \leq q]$. For right-tail and two-tail probabilities, draw a picture to inform how to use and combine calls to `NORM.DIST()` in order to find desired probabilities.
+ **`=NORM.INV(p, mean, sd)`** returns the value $x^*$ such that $\mathbb{P}[X \leq x^*] = p$ — the $p^{th}$ percentile. Remember that $p$ must be the area to the **left** of the desired cutoff.
+ When a scenario involves repeated independent yes/no trials with constant probability, use the **binomial distribution** (`BINOM.DIST()`). When data are approximately normally distributed, use the **normal distribution** (`NORM.DIST()`, `NORM.INV()`).
:::

<!--

:::{.callout-tip}
## Looking Ahead

The next activity is a discrete probability and simulation lab exploring the *hot hand* phenomenon in basketball — a fun application of the ideas we've encountered in our short study of probability. You'll simulate data, compare distributions, and think carefully about what "random" actually looks like in practice.

:::

-->