---
title: "Topic 7: Discrete Distributions Lab"
format: 
  live-html:
    include-in-header:
      text: |
        <style>
        .btn-exercise-solution { display: none !important; }
        </style>
    webr:
      packages:
        - dplyr
        - ggplot2
        - gradethis
        - openintro
        - patchwork
        - jsonlite
        - base64enc
      cell-options:
        persist: true
    live:
      show-solutions: true
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}
{{< include ./_extensions/r-wasm/live/_gradethis.qmd >}}

```{webr}
#| edit: false
#| output: false
#| define:
#|   - ok_response

library(htmltools)
ok_response <- function(response, n) {
  if (is.na(response)) div(HTML("You haven't answered yet."), style = "color: purple")
  else if (response == n) div(HTML("Correct ✓"), style = "color: green")
  else div(HTML("Incorrect ✗"), style = "color: red")
}
```

```{ojs}
//| echo: false

function ok_checkbox(response, n) {
  if (!response || response.length === 0) 
    return html`<span style="color:purple">You haven't answered yet.</span>`;
  if (response.toString() === n) 
    return html`<span style="color:green">Correct ✓</span>`;
  return html`<span style="color:red">Not Yet! ✗</span>`;
}
```

```{webr}
#| context: setup
#| echo: false
#| output: false

library(dplyr)
library(ggplot2)
library(openintro)
library(patchwork)
library(jsonlite)
library(base64enc)

data(kobe_basket)
kobe_streak <- calc_streak(kobe_basket$shot)

exercise_results <- list(
  notebook           = "Topic 7: Discrete Distributions Lab",
  plot_streak_kobe   = "unattempted",
  sim_independent    = "unattempted",
  sim_streak         = "unattempted"
)
```

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(openintro)
library(kableExtra)
data(kobe_basket)
```

:::{.callout-note}
## About

In this lab, we learn how to simulate data and compare simulated data to observed data in order to informally test a hypothesis.

```{ojs}
//| echo: false

{
  const btn = html`<button style="
    background: #436b95;
    color: white;
    border: none;
    padding: 8px 16px;
    border-radius: 4px;
    cursor: pointer;
    font-size: 0.9em;
  ">🔄 Reset My Responses</button>`;

  btn.onclick = () => {
    if (confirm('Reset all responses for this activity? This cannot be undone.')) {
      Object.keys(localStorage)
        .filter(k => k.endsWith('_selected_t7') || k.endsWith('_correct_t7') || k.endsWith('_result_t7'))
        .forEach(k => localStorage.removeItem(k));
      location.reload();
    }
  };

  return btn;
}
```


*Note.* The button above resets multiple choice and checkbox questions. Currently, resetting code cells must be done manually via hitting the *Start Over* button on each individual interactive cell.

:::

:::{.callout-note}
## License

This is a derivative of a product of OpenIntro that is released under a [Creative Commons Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0) license. The original lab was adapted for OpenIntro by Andrew Bray and Mine Çetinkaya-Rundel from a lab written by Mark Hansen of UCLA Statistics.
:::

## Discrete Distributions and Simulation

The previous two activities discussed probability distributions. In this seventh activity you'll engage in computer simulation to test a theory by building and comparing distributions. Computer simulations can be really helpful because they allow us to repeat tasks very quickly (much more quickly than we would be able to do so in real life) and they also allow us to model tasks which may be impossible or unethical to perform in real life. Simulation is a truly powerful tool for helping us understand processes in our world.

## Kobe Bryant and the *Hot Hand*

Basketball players who make several baskets in succession are often described as having a *hot hand*. Fans and players have long believed in the *hot hand* phenomenon, which refutes the assumption that each shot is independent of the next. However, a 1985 paper by Gilovich, Vallone, and Tversky collected evidence that contradicted this belief and showed that successive shots are indeed independent events ([link to paper](https://labs.la.utexas.edu/gilden/files/2016/04/Gilo.Vallone.Tversky.pdff){target="_blank"}). This paper started a great controversy that continues to this day, as you can see by searching for *hot hand basketball* on the web.

We do not expect to resolve this controversy today. However, in this activity we'll apply one approach to answering questions like this. The goals for the remainder of this activity are to: 

1. think about the effects of independent and dependent events, 
2. learn how to simulate data in R, and 
3. compare simulated data to actual observed data in order to determine whether the observed data is fundamentally different from the simulated data.

:::{.callout-caution}
## Check Your Understanding: Binomial Assumptions

Which of the following assumptions must be made if we are to simulate a set of basketball shots as a binomial experiment?

```{ojs}
//| echo: false

viewof q1 = Inputs.checkbox(
  new Map([
    ["Every shot has two possible outcomes — a 'hit' or a 'miss'", 1],
    ["If a player 'hits' a shot, then the likelihood that they make their next shot goes up", 2],
    ["Shots are independent — that is, the hot hand phenomenon does not exist", 3],
    ["We must simulate at least 50 shots", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q1_selected_t7") ?? "[]") ?? []}
);

{
  localStorage.setItem("q1_selected_t7", JSON.stringify(q1));
  localStorage.setItem("q1_correct_t7", "1,3");
  localStorage.setItem("q1_result_t7", (!q1 || q1.length === 0) ? "unattempted" : (q1.toString() === "1,3" ? "correct" : "incorrect"));
}
ok_checkbox(q1.toString(), "1,3");
```

:::

Our investigation will focus on the performance of one player: Kobe Bryant of the Los Angeles Lakers. His performance against the Orlando Magic in the 2009 NBA Finals earned him the title *Most Valuable Player* and many spectators commented on how he appeared to show a hot hand. Let's look at the first several rows of Kobe's gameplay data from that series.

```{r}
#| echo: false
#| eval: true

kobe_basket %>% 
  head() %>%
  kable() %>%
  kable_styling(bootstrap_options = c("hover", "striped"))
```

In this data frame, every row records a shot taken by Kobe Bryant. If he hit the shot (made a basket), a hit (`H`) is recorded in the column named `shot`, otherwise a miss (`M`) is recorded.

Just looking at the string of hits and misses, it can be difficult to gauge whether or not it seems like Kobe was shooting with a hot hand. One way we can approach this is by considering the belief that hot hand shooters tend to go on shooting streaks. For this lab, we define the length of a shooting streak to be the *number of consecutive baskets made until a miss occurs*.

For example, in Game 1 Kobe had the following sequence of hits and misses from his nine shot attempts in the first quarter:

$$\textrm{H M | M | H H M | M | M | M}$$

To verify this, run the command `kobe_basket |> pull(shot) |> head(n = 9)` in the code block below.

```{webr}
#| exercise: first9
# Use this code block to verify the first 9 shots

```

:::{.hint exercise="first9"}
:::{.callout-note collapse="false"}
## Hint 1 (Solved)

The `pull()` function extracts a column from a data frame as a vector, and `head(n = 9)` returns the first 9 elements.

```{r}
#| echo: true
#| eval: false

kobe_basket |> pull(shot) |> head(n = 9)
```

:::
:::

```{webr}
#| exercise: first9
#| solution: true

kobe_basket |> pull(shot) |> head(n = 9)
```

```{webr}
#| exercise: first9
#| check: true

exercise_results$first9 <<- "fail"
gradethis::grade_this({
  feedback <- gradethis::code_feedback()
  if (is.null(feedback)) {
    exercise_results$first9 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

Within the nine shot attempts, there are six streaks, which are separated by a "|" above. Their lengths are one, zero, two, zero, zero, zero (in order of occurrence).

:::{.callout-caution}
## Check Your Understanding: Streak Lengths I

What does a streak length of 1 mean — how many hits and misses are in a streak of 1?

```{ojs}
//| echo: false

mutable ok_response = (response, n) => { return html`Loading...` };
viewof q2 = Inputs.radio(
  new Map([
    ["One hit", 1],
    ["One hit and one miss", 2],
    ["One hit and any number of misses", 3],
    ["Just a miss", 4],
    ["Any number of misses", 5]
  ]),
  {value: JSON.parse(localStorage.getItem("q2_selected_t7") ?? "null")}
);

{
  localStorage.setItem("q2_selected_t7", JSON.stringify(q2));
  localStorage.setItem("q2_correct_t7", "2");
  localStorage.setItem("q2_result_t7", q2 === null ? "unattempted" : (q2 == 2 ? "correct" : "incorrect"));
}
ok_response(q2, "2");
```

:::

:::{.callout-caution}
## Check Your Understanding: Streak Lengths II

What does a streak of length 0 mean?

```{ojs}
//| echo: false

viewof q3 = Inputs.radio(
  new Map([
    ["One hit", 1],
    ["One hit and one miss", 2],
    ["One hit and any number of misses", 3],
    ["Just a single miss", 4],
    ["Any number of misses", 5]
  ]),
  {value: JSON.parse(localStorage.getItem("q3_selected_t7") ?? "null")}
);

{
  localStorage.setItem("q3_selected_t7", JSON.stringify(q3));
  localStorage.setItem("q3_correct_t7", "4");
  localStorage.setItem("q3_result_t7", q3 === null ? "unattempted" : (q3 == 4 ? "correct" : "incorrect"));
}
ok_response(q3, "4");
```

:::

In this activity you have access to the custom function `calc_streak()` from the `{openintro}` package. This function calculates the lengths of all shooting streaks. Store the result of `kobe_basket |> pull(shot) |> calc_streak()` in a new object called `kobe_streak`, and then call `kobe_streak` to see the streak lengths printed out.

```{webr}
#| exercise: calc-streak-kobe
______
```

:::{.hint exercise="calc-streak-kobe"}
:::{.callout-note collapse="false"}
## Hint 1

Use the assignment operator `<-` to store the result in `kobe_streak`.

```{r}
#| echo: true
#| eval: false

kobe_streak <- ___
```

:::
:::

:::{.hint exercise="calc-streak-kobe"}
:::{.callout-note collapse="false"}
## Hint 2

Start with the `kobe_streak` data frame.

```{r}
#| echo: true
#| eval: false

kobe_streak <- kobe_basket 
```

:::
:::

:::{.hint exercise="calc-streak-kobe"}
:::{.callout-note collapse="false"}
## Hint 3

Start with the `kobe_streak` data frame, and then...

```{r}
#| echo: true
#| eval: false

kobe_streak <- kobe_basket |>
```

:::
:::

:::{.hint exercise="calc-streak-kobe"}
:::{.callout-note collapse="false"}
## Hint 4

Start with the `kobe_streak` data frame, and then extract the values from the `shot` column.

```{r}
#| echo: true
#| eval: false

kobe_streak <- kobe_basket |>
  pull(shot)
```

:::
:::

:::{.hint exercise="calc-streak-kobe"}
:::{.callout-note collapse="false"}
## Hint 5

Start with the `kobe_streak` data frame, and then extract the values from the `shot` column, and then...

```{r}
#| echo: true
#| eval: false

kobe_streak <- kobe_basket |>
  pull(shot) |>
```

:::
:::

:::{.hint exercise="calc-streak-kobe"}
:::{.callout-note collapse="false"}
## Hint 6

Start with the `kobe_streak` data frame, and then extract the values from the `shot` column, and then calculate the streak lengths with `calc_streak()`.

```{r}
#| echo: true
#| eval: false

kobe_streak <- kobe_basket |>
  pull(shot) |>
  calc_streak()
```

:::
:::

:::{.hint exercise="calc-streak-kobe"}
:::{.callout-note collapse="false"}
## Hint 7 (Solved)

Now print out the results with a single line, running `kobe_streak`.

```{r}
#| echo: true
#| eval: false

kobe_streak <- kobe_basket |>
  pull(shot) |>
  calc_streak()

kobe_streak
```

:::
:::

```{webr}
#| exercise: calc-streak-kobe
#| solution: true

kobe_streak <- kobe_basket |>
  pull(shot) |>
  calc_streak()

kobe_streak
```

```{webr}
#| exercise: calc-streak-kobe
#| check: true

exercise_results$calc_streak_kobe <<- "fail"
gradethis::grade_this({
  feedback <- gradethis::code_feedback()
  if (is.null(feedback)) {
    exercise_results$calc_streak_kobe <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

The result from `calc_streak()` is a data frame with a single column named `length`. Now use `ggplot` to make a barplot of `kobe_streak`.

```{webr}
#| exercise: plot-streak-kobe
______
```

:::{.hint exercise="plot-streak-kobe"}
:::{.callout-note collapse="false"}
## Hint 1

Before trying to build the plot, simply type and run `kobe_streak` to verify the contents of the `kobe_streak` variable.

:::
:::

:::{.hint exercise="plot-streak-kobe"}
:::{.callout-note collapse="false"}
## Hint 2

Start with the `kobe_streak` data frame.

```{r}
#| echo: true
#| eval: false

kobe_streak
```

:::
:::

:::{.hint exercise="plot-streak-kobe"}
:::{.callout-note collapse="false"}
## Hint 3

Start with the `kobe_streak` data frame, and then...

```{r}
#| echo: true
#| eval: false

kobe_streak |>
```

:::
:::

:::{.hint exercise="plot-streak-kobe"}
:::{.callout-note collapse="false"}
## Hint 4

Start with the `kobe_streak` data frame, and then take out a canvas to plot on.

```{r}
#| echo: true
#| eval: false

kobe_streak |>
  ggplot()
```

:::
:::

:::{.hint exercise="plot-streak-kobe"}
:::{.callout-note collapse="false"}
## Hint 5

Start with the `kobe_streak` data frame, and then take out a canvas to plot on. Next add a bar graph layer...

```{r}
#| echo: true
#| eval: false

kobe_streak |>
  ggplot() +
  geom_bar(aes(x = ___))
```

:::
:::

:::{.hint exercise="plot-streak-kobe"}
:::{.callout-note collapse="false"}
## Hint 6 (Solved)

Start with the `kobe_streak` data frame, and then take out a canvas to plot on. Next add a bar graph layer with `length` as the `x` aesthetic.

```{r}
#| echo: true
#| eval: false

kobe_streak |>
  ggplot() +
  geom_bar(aes(x = length))
```

:::
:::

```{webr}
#| exercise: plot-streak-kobe
#| solution: true

kobe_streak |> 
  ggplot() + 
  geom_bar(aes(x = length))
```

```{webr}
#| exercise: plot-streak-kobe
#| check: true

exercise_results$plot_streak_kobe <<- "fail"
gradethis::grade_this({
  feedback <- gradethis::code_feedback()
  if (is.null(feedback)) {
    exercise_results$plot_streak_kobe <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

Note that instead of making a histogram, we chose to make a bar plot from a table of the streak data. A bar plot is preferable here since our variable is discrete — counts — instead of continuous.

:::{.callout-caution}
## Check Your Understanding: Kobe's Streak Distribution

Describe the distribution of Kobe's streaks from the 2009 NBA Finals.

```{ojs}
//| echo: false

viewof q4 = Inputs.radio(
  new Map([
    ["The distribution is uniform", 1],
    ["The distribution is symmetric", 2],
    ["The distribution is skewed left", 3],
    ["The distribution is skewed right", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q4_selected_t7") ?? "null")}
);

{
  localStorage.setItem("q4_selected_t7", JSON.stringify(q4));
  localStorage.setItem("q4_correct_t7", "4");
  localStorage.setItem("q4_result_t7", q4 === null ? "unattempted" : (q4 == 4 ? "correct" : "incorrect"));
}
ok_response(q4, "4");
```

:::

:::{.callout-caution}
## Check Your Understanding: Typical Streak Length

What was Kobe's typical streak length?

```{ojs}
//| echo: false

viewof q5 = Inputs.radio(
  new Map([
    ["0", 1],
    ["1", 2],
    ["2", 3],
    ["3", 4],
    ["4", 5],
    ["5", 6]
  ]),
  {value: JSON.parse(localStorage.getItem("q5_selected_t7") ?? "null")}
);

{
  localStorage.setItem("q5_selected_t7", JSON.stringify(q5));
  localStorage.setItem("q5_correct_t7", "1");
  localStorage.setItem("q5_result_t7", q5 === null ? "unattempted" : (q5 == 1 ? "correct" : "incorrect"));
}
ok_response(q5, "1");
```

:::

:::{.callout-caution}
## Check Your Understanding: Longest Streak

How long was Kobe's longest streak?

```{ojs}
//| echo: false

viewof q6 = Inputs.radio(
  new Map([
    ["0", 1],
    ["1", 2],
    ["2", 3],
    ["3", 4],
    ["4", 5],
    ["5", 6]
  ]),
  {value: JSON.parse(localStorage.getItem("q6_selected_t7") ?? "null")}
);

{
  localStorage.setItem("q6_selected_t7", JSON.stringify(q6));
  localStorage.setItem("q6_correct_t7", "5");
  localStorage.setItem("q6_result_t7", q6 === null ? "unattempted" : (q6 == 5 ? "correct" : "incorrect"));
}
ok_response(q6, "5");
```

:::

## Compared to What?

We've shown that Kobe had some long shooting streaks, but are they long enough to support the belief that he had hot hands? What can we compare them to? 

It's impossible to run an experiment where we manipulate the *hot hand* condition -- it is posited as a phenomenon, after all. Because of this, we'll return to the idea of *independence*, which is something that we have been able to define. Two processes are independent if the outcome of one process doesn't affect the outcome of the second. If each shot that a player takes is an independent process, having made or missed your first shot will not affect the probability that you will make or miss your second shot.

A shooter with a hot hand will have shots that are *not* independent of one another. Specifically, if the shooter makes their first shot, the hot hand model says they will have a *higher* probability of making their second shot.

Let's suppose for a moment that the hot hand model is valid for Kobe. During his career, the percentage of time Kobe makes a basket (his shooting percentage) is about 45%, or in probability notation,

$$\mathbb{P}\left[\textrm{first shot = H}\right] = 0.45$$

If he makes the first shot and has a hot hand (*not* independent shots), then the probability that he makes his second shot would go up to a higher percentage, say, 60%,

$$\mathbb{P}\left[\textrm{second shot = H} \, | \, \textrm{first shot = H}\right] = 0.60$$

As a result of these increased probabilities, you'd expect Kobe to have longer streaks. Compare this to the skeptical perspective where Kobe does *not* have a hot hand, where each shot is independent of the next. If he hit his first shot, the probability that he makes the second is still 0.45.

$$\mathbb{P}\left[\textrm{second shot = H} \, | \, \textrm{first shot = H}\right] = 0.45$$

In other words, making the first shot did nothing to affect the probability that he'd make his second shot. If Kobe's shots are independent, then he'd have the same probability of hitting every shot regardless of his past shots: 45%.

Now that we've phrased the situation in terms of independent shots, let's return to the question: how do we tell if Kobe's shooting streaks are long enough to indicate that he has hot hands? We can compare his streak lengths to someone without hot hands: an independent shooter.

## Simulations in R

We'll come back to Kobe shortly, but for now let's think about how we might simulate an independent shooter. While we don't have any data from a shooter we know to have independent shots, that sort of data is very easy to simulate in R. In a simulation, you set the ground rules of a random process and then the computer uses random numbers to generate an outcome that adheres to those rules. For example, the following code block is set up to simulate a single flip of a fair coin. Run it a few times and see what happens — then edit the code to simulate 10 flips at a time, or a hundred, or maybe even a hundred thousand! The computer can flip 100,000 coins much faster than you and I could ever hope to.

```{webr}
#| exercise: head-tail
#| caption: Ungraded, Exploratory Only

sample(c("heads", "tails"), size = 1, replace = TRUE)
```

:::{.hint exercise="head-tail"}
:::{.callout-note collapse="false"}
## Hint 1

To increase the number of simulated flips, edit the `size` parameter.

:::
:::

The first argument `c("heads", "tails")` represents the *outcomes* and can be thought of as a hat with two slips of paper in it: one slip says `heads` and the other says `tails`. The function `sample` draws one slip from the hat and tells us if it was a head or a tail.

As you've discovered, if you wanted to simulate flipping a fair coin 100 times, you could either run the `sample()` function 100 times or, more simply, adjust the `size` argument, which governs how many flips to simulate. The `replace = TRUE` argument indicates we put the selected slip of paper back in the hat before drawing again). Use the code block below to simulate 100 flips of a fair coin, storing the results in an object called `sim_fair_coin`. Then use `table(sim_fair_coin)` to compute a frequency table for the outcomes — what do you notice?

```{webr}
#| exercise: sim-fair-coin
______
```

:::{.hint exercise="sim-fair-coin"}
:::{.callout-note collapse="false"}
## Hint 1

Start with the code provided earlier to simulate your coin flips, but adjust the `size` argument.

:::
:::

:::{.hint exercise="sim-fair-coin"}
:::{.callout-note collapse="false"}
## Hint 2

Use the assignment operator (`<-`) to store the results in `sim_fair_coin`.

```{r}
#| echo: true
#| eval: false

sim_fair_coin <- ___
```

:::
:::

:::{.hint exercise="sim-fair-coin"}
:::{.callout-note collapse="false"}
## Hint 3

Copy and paste the code from earlier. How many flips do you need to simulate?

```{r}
#| echo: true
#| eval: false

sim_fair_coin <- sample(c("heads", "tails"), 
                        size = ___, 
                        replace = TRUE)
```

:::
:::

:::{.hint exercise="sim-fair-coin"}
:::{.callout-note collapse="false"}
## Hint 4

You're asked to simulate 100 flips. Now, use `table()` to conveniently display the results.

```{r}
#| echo: true
#| eval: false

sim_fair_coin <- sample(c("heads", "tails"), 
                        size = 100, 
                        replace = TRUE)

table(___)
```

:::
:::

:::{.hint exercise="sim-fair-coin"}
:::{.callout-note collapse="false"}
## Hint 5 (Solved)

Once you've simulated the flips, use `table(sim_fair_coin)` to compute your frequency table.

```{r}
#| echo: true
#| eval: false

sim_fair_coin <- sample(c("heads", "tails"), 
                        size = 100, 
                        replace = TRUE)
table(sim_fair_coin)
```

:::
:::

```{webr}
#| exercise: sim-fair-coin
#| solution: true

sim_fair_coin <- sample(c("heads", "tails"), 
                        size = 100, 
                        replace = TRUE)
table(sim_fair_coin)
```

```{webr}
#| exercise: sim-fair-coin
#| check: true

exercise_results$sim_fair_coin <<- "fail"
gradethis::grade_this({
  feedback <- gradethis::code_feedback()
  if (is.null(feedback)) {
    exercise_results$sim_fair_coin <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

Since there are only two elements in the *outcomes*, the probability that we "flip" a fair coin and it lands heads is 0.5. Say we're trying to simulate an unfair coin that we know only lands heads 20% of the time. We can adjust for this by adding an argument called `prob`, which provides a vector of two probability weights. Execute the following code cell to simulate this unfair coin.

```{webr}
#| exercise: sim-unfair-coin
#| caption: Ungraded, Exploration Only

sim_unfair_coin <- sample(c("heads", "tails"), 
                          size = 100, 
                          replace = TRUE, 
                          prob = c(0.2, 0.8))
table(sim_unfair_coin)
```

The argument `prob = c(0.2, 0.8)` indicates that for the two elements in the *outcomes* vector, we want to select the first one, `heads`, with probability 0.2 and the second one, `tails`, with probability 0.8. Another way of thinking about this is to imagine the outcome space as a bag of 10 chips, where 2 chips are labeled "head" and 8 chips are labeled "tail". Therefore at each draw, the probability of drawing a chip that says "head" is 20%, and "tail" is 80%.

## Simulating the Independent Shooter

Simulating a basketball player who has independent shots uses the same mechanism that we use to simulate a coin flip. To simulate a single shot from an independent shooter with a shooting percentage of 50% we type:

```{r}
#| echo: true
#| eval: false

sim_basket <- sample(c("H", "M"), size = 1, replace = TRUE)
```

To make a valid comparison between Kobe and our simulated independent shooter, we need to align both their shooting percentage and the number of attempted shots. Use the code block below to help you find the number of shots we need to simulate.

```{webr}
#| exercise: find-num-shots
#| caption: Ungraded, Exploration Only
# Explore the kobe_basket data frame to find the number of shots

```

:::{.hint exercise="find-num-shots"}
:::{.callout-note collapse="false"}
## Hint 1

The simulated shooter should take the same number of shots as Kobe took in the series.

:::
:::

:::{.hint exercise="find-num-shots"}
:::{.callout-note collapse="false"}
## Hint 2

Every row of the `kobe_basket` data frame corresponds to a shot taken by Kobe Bryant in the series.

:::
:::

:::{.hint exercise="find-num-shots"}
:::{.callout-note collapse="false"}
## Hint 3

There are several ways to find the number of rows in a data frame — try `nrow()`, `dim()`, or `glimpse()`.

:::
:::

:::{.callout-caution}
## Check Your Understanding: Number of Shots

How many shots did Kobe Bryant take during the 2009 NBA Finals?

```{ojs}
//| echo: false

viewof q7 = Inputs.radio(
  new Map([
    ["0", 1],
    ["10", 2],
    ["123", 3],
    ["133", 4],
    ["76", 5],
    ["6", 6],
    ["45", 7]
  ]),
  {value: JSON.parse(localStorage.getItem("q7_selected_t7") ?? "null")}
);

{
  localStorage.setItem("q7_selected_t7", JSON.stringify(q7));
  localStorage.setItem("q7_correct_t7", "4");
  localStorage.setItem("q7_result_t7", q7 === null ? "unattempted" : (q7 == 4 ? "correct" : "incorrect"));
}
ok_response(q7, "4");
```

:::

:::{.callout-caution}
## Check Your Understanding: Shooting Percentage

What should the shooting percentage be for our simulated random independent shooter?

```{ojs}
//| echo: false

viewof q8 = Inputs.radio(
  new Map([
    ["50%", 1],
    ["45%", 2],
    ["60%", 3],
    ["40%", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q8_selected_t7") ?? "null")}
);

{
  localStorage.setItem("q8_selected_t7", JSON.stringify(q8));
  localStorage.setItem("q8_correct_t7", "2");
  localStorage.setItem("q8_result_t7", q8 === null ? "unattempted" : (q8 == 2 ? "correct" : "incorrect"));
}
ok_response(q8, "2");
```

:::

Use your answers from the previous questions, our discussion about simulated coin flipping, and the code block below to simulate an independent shooter which we can compare to Kobe Bryant. Store the results of your simulated shots in an object called `sim_basket`, and then print them out.

```{webr}
#| exercise: sim-independent
______
```

:::{.hint exercise="sim-independent"}
:::{.callout-note collapse="false"}
## Hint 1

Use the assignment operator (`<-`) to assign the outcome to `sim_basket`.

:::
:::

:::{.hint exercise="sim-independent"}
:::{.callout-note collapse="false"}
## Hint 2

Use the `sample()` function to simulate the shooter.

```{r}
#| echo: true
#| eval: false

sim_basket <- sample(c("H", "M"), 
                     ___, 
                     ___, 
                     ___)
```

:::
:::

:::{.hint exercise="sim-independent"}
:::{.callout-note collapse="false"}
## Hint 3

For Kobe Bryant, the probability of hitting a shot is 0.45 and missing is 0.55. Use this to fill in the `prob` argument.

```{r}
#| echo: true
#| eval: false

sim_basket <- sample(c("H", "M"), 
                     ___, 
                     prob = c(0.45, 0.55),
                     ___)
```

:::
:::

:::{.hint exercise="sim-independent"}
:::{.callout-note collapse="false"}
## Hint 4

How many shots should we simulate? Fill in the `size` argument, and remember to set `replace = TRUE`.

```{r}
#| echo: true
#| eval: false

sim_basket <- sample(c("H", "M"), 
                     size = ___, 
                     prob = c(0.45, 0.55),
                     replace = ___)
```

:::
:::

:::{.hint exercise="sim-independent"}
:::{.callout-note collapse="false"}
## Hint 5

Kobe took 133 shots during the playoff series.

```{r}
#| echo: true
#| eval: false

sim_basket <- sample(c("H", "M"), 
                     size = 133, 
                     prob = c(0.45, 0.55),
                     replace = TRUE)
```

:::
:::

:::{.hint exercise="sim-independent"}
:::{.callout-note collapse="false"}
## Hint 6 (Solved)

Now that you've simulated those independent shots, print out the results.

```{r}
#| echo: true
#| eval: false

sim_basket <- sample(c("H", "M"), 
                     size = 133, 
                     prob = c(0.45, 0.55),
                     replace = TRUE)

sim_basket
```

:::
:::

```{webr}
#| exercise: sim-independent
#| solution: true

sim_basket <- sample(c("H", "M"), 
                     size = 133, 
                     prob = c(0.45, 0.55), 
                     replace = TRUE)

sim_basket
```

```{webr}
#| exercise: sim-independent
#| check: true

exercise_results$sim_independent <<- "fail"
gradethis::grade_this({
  feedback <- gradethis::code_feedback()
  if (is.null(feedback)) {
    exercise_results$sim_independent <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

Note that we've named the new vector `sim_basket`, the same name that we gave to the previous simulation of a single shot, reflecting a shooting percentage of 50%. In this situation, R overwrites the old object with the new one, so always make sure that you don't need the information in an old variable before assigning something new to it.

With the results of the simulation saved as `sim_basket`, we have the data necessary to compare Kobe to our independent shooter. Execute the following code cell to see Kobe's actual shots alongside our simulated shooter's shots. I've created a fixed version of `sim_basket` using a random seed so that we will all see the same simulated data for comparison.

```{webr}
#| exercise: compare-basket
#| caption: Ungraded, Exploratory Only

set.seed(8)
sim_basket <- sample(c("H", "M"), 
                     size = 133, 
                     replace = TRUE)
#Kobe shots
kobe_basket |> pull(shot)

#Simulated shots
sim_basket
```

Both data sets represent the results of 133 shot attempts, each with the same shooting percentage of 45%. We know that our simulated data is from a shooter that has independent shots — that is, we know the simulated shooter does **not** have a hot hand.

Use the `calc_streak()` function in the code block below to compute the streak lengths of `sim_basket` and store the result in an object called `sim_streak`. Print out the streak lengths.

```{webr}
#| exercise: sim-streak
#| setup: true
#| echo: false

set.seed(8)
sim_basket <- sample(c("H", "M"), 
                     size = 133, 
                     replace = TRUE)
```

```{webr}
#| exercise: sim-streak
______
```

:::{.hint exercise="sim-streak"}
:::{.callout-note collapse="false"}
## Hint 1

Use the `calc_streak()` function on `sim_basket` and store the result as `sim_streak`.

```{r}
#| echo: true
#| eval: false

sim_streak <- calc_streak(___)
```

:::
:::

:::{.hint exercise="sim-streak"}
:::{.callout-note collapse="false"}
## Hint 2

Use the `calc_streak()` function on `sim_basket` and store the result as `sim_streak`. Now print out the results.

```{r}
#| echo: true
#| eval: false

sim_streak <- calc_streak(sim_basket)
```

:::
:::

:::{.hint exercise="sim-streak"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

Use the `calc_streak()` function on `sim_basket` and store the result as `sim_streak`. Now print out the results.

```{r}
#| echo: true
#| eval: false

sim_streak <- calc_streak(sim_basket)
sim_streak
```

:::
:::

```{webr}
#| exercise: sim-streak
#| solution: true

sim_streak <- calc_streak(sim_basket)
```

```{webr}
#| exercise: sim-streak
#| check: true

exercise_results$sim_streak <<- "fail"
gradethis::grade_this({
  feedback <- gradethis::code_feedback()
  if (is.null(feedback)) {
    exercise_results$sim_streak <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

Okay, now you've got everything you need in order to help you discover whether there is evidence to suggest that Kobe Bryant had a *hot hand* in the 2009 NBA Finals. Use the code blocks below each question to help you answer.

**1.** Describe the distribution of streak lengths. What is the typical streak length for this simulated independent shooter with a 45% shooting percentage? How long is the simulated player's longest streak of baskets in 133 shots?

```{webr}
#| exercise: discuss-q1
#| setup: true
#| echo: false

set.seed(8)
sim_basket <- sample(c("H", "M"), 
                     size = 133, 
                     replace = TRUE)
sim_streak <- calc_streak(sim_basket)
```

```{webr}
#| exercise: discuss-q1
#| caption: Ungraded, Exploratory Only
# Explore the distribution of sim_streak

```

:::{.hint exercise="discuss-q1"}
:::{.callout-note collapse="false"}
## Hint 1

One approach is to produce a barplot of `sim_streak`, just like you did earlier for `kobe_streak`, and compare the two distributions visually.

:::
:::

:::{.hint exercise="discuss-q1"}
:::{.callout-note collapse="false"}
## Hint 2

Start with the `sim_streak` data frame, and then...

```{r}
#| echo: true
#| eval: false
sim_streak |>
```

:::
:::

:::{.hint exercise="discuss-q1"}
:::{.callout-note collapse="false"}
## Hint 3

Start with the `sim_streak` data frame, and then take out a canvas to draw a plot.

```{r}
#| echo: true
#| eval: false
sim_streak |>
  ggplot()
```

:::
:::

:::{.hint exercise="discuss-q1"}
:::{.callout-note collapse="false"}
## Hint 4

Start with the `sim_streak` data frame, and then take out a canvas to draw a plot. Now add a barplot layer to your plot.

```{r}
#| echo: true
#| eval: false
sim_streak |>
  ggplot() + 
  geom_bar(aes(x = ___))
```

:::
:::

:::{.hint exercise="discuss-q1"}
:::{.callout-note collapse="false"}
## Hint 5 (Solved)

Start with the `sim_streak` data frame, and then take out a canvas to draw a plot. Now add a barplot layer to your plot, mapping the streak `length` to the `x` aesthetic..

```{r}
#| echo: true
#| eval: false
sim_streak |>
  ggplot() + 
  geom_bar(aes(x = length))
```

:::
:::

**2.** If you were to run the simulation of the independent shooter a second time, how would you expect its streak distribution to compare to the distribution from the question above? Exactly the same? Somewhat similar? Totally different? Explain your reasoning.

```{webr}
#| exercise: discuss-q2
#| caption: Ungraded, Exploratory Only
# Try re-running the simulation and observe the results

```

:::{.hint exercise="discuss-q2"}
:::{.callout-note collapse="false"}
## Hint 1

There's no single correct answer here. The code block is here is for you to actually try it out. Copy, paste, and execute the simulation code to create a new independent shooter — what do you observe?

:::
:::

:::{.hint exercise="discuss-q2"}
:::{.callout-note collapse="false"}
## Hint 2

Next, calculate the streak lengths and store them.

```{r}
#| echo: true
#| eval: false 
sim_basket <- sample(c("H", "M"), 
                     size = 133, 
                     prob = c(0.45, 0.55), 
                     replace = TRUE)

sim_streak <- calc_streak(sim_basket)
```

:::
:::

:::{.hint exercise="discuss-q2"}
:::{.callout-note collapse="false"}
## Hint 3

Now, plot the results.

```{r}
#| echo: true
#| eval: false 
sim_basket <- sample(c("H", "M"), 
                     size = 133, 
                     prob = c(0.45, 0.55), 
                     replace = TRUE)

sim_streak <- calc_streak(sim_basket)

sim_streak %>%
  ggplot() + 
  geom_bar(aes(x = length))
```

:::
:::

:::{.hint exercise="discuss-q2"}
:::{.callout-note collapse="false"}
## Hint 4 (Solved)

Run the code several times to see what happens. Discuss what you see and compare it to Kobe's distribution of streaks.

```{r}
#| echo: true
#| eval: false 
sim_basket <- sample(c("H", "M"), 
                     size = 133, 
                     prob = c(0.45, 0.55), 
                     replace = TRUE)

sim_streak <- calc_streak(sim_basket)

sim_streak %>%
  ggplot() + 
  geom_bar(aes(x = length))
```

:::
:::

**Discussion Question 3.** How does Kobe Bryant's distribution of streak lengths compare to the distribution of streak lengths for the simulated shooter? Using this comparison, do you have evidence Kobe's shooting patterns exhibit the *hot hand* phenomenon?

```{webr}
#| exercise: discuss-q3
#| fig.height: 2
#| caption: Ungraded, Exploratory Only
# Compare kobe_streak and sim_streak

```

:::{.hint exercise="discuss-q3"}
:::{.callout-note collapse="false"}
## Hint 1

There's no single correct answer here either, but you might use `ggplot()` to compare Kobe's actual streak length distribution to that of your simulated shooter side by side. As a reminder, your simulated shooter does **not** have a hot hand.

:::
:::

:::{.hint exercise="discuss-q3"}
:::{.callout-note collapse="false"}
## Hint 2

A shooter with a *hot hand* should have a distribution of streaks that is different from the simulated streak distributions you are generating. In particular, a shooter with a *hot hand* would have more long streaks of made shots.

:::
:::

:::{.hint exercise="discuss-q3"}
:::{.callout-note collapse="false"}
## Hint 3

Run the code below several times. Does it look like Kobe Bryant has a streak distribution that differs from our independent shooters'?

```{r}
#| echo: true
#| eval: false
#| fig.height: 3
sim_basket <- sample(c("H", "M"), 
                     size = 133, 
                     prob = c(0.45, 0.55), 
                     replace = TRUE)

sim_streak <- calc_streak(sim_basket)

p1 <- kobe_streak %>%
  ggplot() + 
  geom_bar(aes(x = length)) + 
  labs(
    title = "Kobe"
  )

p2 <- sim_streak %>%
  ggplot() + 
  geom_bar(aes(x = length)) + 
  labs(
    title = "Simulated Independent Shooter"
  )

p1 + p2
```

:::
:::

## Submit

If you are part of a course with an instructor who is grading your work on these activities, please copy and submit ***both*** of the hashes below using the method your instructor has requested.

:::{.callout-warning}
## Question Hash

The hash below encodes your responses to the multiple choice and checkbox questions in this activity.

```{ojs}
//| echo: false

function buildQuestionResults() {
  return {
    notebook: "Topic 7: Discrete Distributions Lab",
    type: "questions",
    timestamp: new Date().toISOString(),
    questions: {
      q1_binomial_assumptions: {
        selected: q1,
        correct_answer: "1,3",
        result: (!q1 || q1.length === 0) ? "unattempted" : (q1.toString() === "1,3" ? "correct" : "incorrect")
      },
      q2_streak_length_1: {
        selected: q2,
        correct_answer: "2",
        result: q2 === null ? "unattempted" : (q2 == 2 ? "correct" : "incorrect")
      },
      q3_streak_length_0: {
        selected: q3,
        correct_answer: "4",
        result: q3 === null ? "unattempted" : (q3 == 4 ? "correct" : "incorrect")
      },
      q4_streak_distribution: {
        selected: q4,
        correct_answer: "4",
        result: q4 === null ? "unattempted" : (q4 == 4 ? "correct" : "incorrect")
      },
      q5_typical_streak: {
        selected: q5,
        correct_answer: "1",
        result: q5 === null ? "unattempted" : (q5 == 1 ? "correct" : "incorrect")
      },
      q6_longest_streak: {
        selected: q6,
        correct_answer: "5",
        result: q6 === null ? "unattempted" : (q6 == 5 ? "correct" : "incorrect")
      },
      q7_num_shots: {
        selected: q7,
        correct_answer: "4",
        result: q7 === null ? "unattempted" : (q7 == 4 ? "correct" : "incorrect")
      },
      q8_shooting_percentage: {
        selected: q8,
        correct_answer: "2",
        result: q8 === null ? "unattempted" : (q8 == 2 ? "correct" : "incorrect")
      }
    }
  };
}

function toBase64(str) {
  return btoa(unescape(encodeURIComponent(str)));
}

question_hash = {
  q1; q2; q3; q4; q5; q6; q7; q8;
  return toBase64(JSON.stringify(buildQuestionResults()));
}
```

```{ojs}
//| echo: false

html`<div style="font-family: monospace; font-size: 0.85em; background: #f5f5f5; padding: 12px; border-radius: 6px; word-break: break-all; border: 1px solid #ddd; user-select: all; cursor: pointer;" onclick="navigator.clipboard.writeText(this.innerText)">
  ${question_hash}
</div>
<p style="margin-top: 8px; font-size: 0.9em; color: #555;">
  Click the box to copy to clipboard.
</p>`
```

:::

:::{.callout-warning}
## Exercise Hash

Click the button below to generate your exercise submission code. This hash encodes your work on the graded code exercises in this activity.

You must have attempted the graded exercises before clicking — clicking generates a snapshot of your current results. If you have completed the activity over multiple sessions, please go back through and hit the *Run Code* button on each graded exercise before generating the hash below, to ensure your most recent results are recorded.

```{webr}
#| exercise: generate-exercise-hash
#| caption: "Generate Exercise Code"

exercise_results$timestamp <- format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
exercise_results$type <- "exercises"

hash <- base64encode(charToRaw(toJSON(exercise_results, auto_unbox = TRUE)))
cat(hash)
```

:::

## Summary

In this lab you explored the power of computer simulation as a tool for informally examining hypotheses about real-world processes. Below are the key takeaways and a heads-up about what to expect moving forward.

:::{.callout-tip}
## Main Takeaways

+ **Simulation allows us to model independent processes.** By using R's `sample()` function with appropriate outcome spaces and probabilities, we can generate data from a process we fully control — one where we *know* shots are independent.
+ **The `prob` argument in `sample()` lets us tune our simulation to match real-world conditions.** Setting `prob = c(0.45, 0.55)` ensures our simulated shooter has the same shooting percentage as Kobe Bryant, making the comparison meaningful.
+ **Streak length is a useful summary of shooting patterns.** The `calc_streak()` function converts a sequence of hits and misses into a distribution of streak lengths, which gives us a way to compare Kobe's actual performance to what we'd expect from an independent shooter.
+ **Comparing observed data to simulated data is a form of informal hypothesis testing.** If Kobe's streak distribution *looks* similar to those of our simulated independent shooters, we have little evidence of a hot hand. If it looks dramatically different, we might reconsider.
+ **Simulation results vary from run to run.** Each simulated shooter produces a slightly different streak distribution, which is a reminder that random processes are inherently variable — we should look for consistent *patterns* across many simulations rather than drawing conclusions from a single run.
:::

:::{.callout-tip}
## Looking Ahead

In this lab you used simulation *informally* — generating data from a known process and comparing it to observed data by eye. While we won't continue using simulation as our primary tool, the underlying logic carries forward throughout the rest of the course. In the coming activities we'll formalize this process of comparing what we observe to what we'd expect if some assumption were true, and we'll develop rigorous statistical methods for deciding when the difference between the two is large enough to matter. That process — *hypothesis testing* — is one of the most important ideas in all of statistics, and everything you did in this lab was quietly building intuition for it.
:::
