---
title: "Topic 15: Hypothesis Tests and Confidence Intervals for Numerical Data (Spreadsheets)"
format: 
  live-html:
    include-in-header:
      text: |
        <style>
        .btn-exercise-solution { display: none !important; }
        .exercise-editor-body { max-height: 300px; overflow-y: auto; }
        .exercise-editor-btn-run-code .btn-label-exercise-editor {
          font-size: 0;
        }
        .exercise-editor-btn-run-code .btn-label-exercise-editor::after {
          content: "Submit Answer";
          font-size: 1rem;
        }
        </style>
    webr:
      packages:
        - dplyr
        - ggplot2
        - gradethis
        - jsonlite
        - base64enc
      cell-options:
        persist: true
    live:
      show-solutions: true
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}
{{< include ./_extensions/r-wasm/live/_gradethis.qmd >}}

```{webr}
#| edit: false
#| output: false
#| define:
#|   - ok_response

library(htmltools)
ok_response <- function(response, n) {
  if (is.na(response)) div(HTML("You haven't answered yet."), style = "color: purple")
  else if (response == n) div(HTML("Correct ✓"), style = "color: green")
  else div(HTML("Incorrect ✗"), style = "color: red")
}
```

```{webr}
#| context: setup
#| echo: false
#| output: false
#| define:
#|   - boundary1
#|   - df1
#|   - boundary2
#|   - df2
#|   - area3
#|   - df3
#|   - clevel4
#|   - df4

library(dplyr)
library(ggplot2)
library(jsonlite)
library(base64enc)

theme_set(theme_bw(base_size = 12))

set.seed(241)
boundary1 <- round(runif(1, 0.75, 2), 2)
df1       <- sample(5:25, size = 1)
boundary2 <- round(runif(1, -2, -1.5), 2)
df2       <- sample(5:25, size = 1)
area3     <- round(runif(1, 0.05, 0.95), 2)
df3       <- sample(5:25, size = 1)
clevel4   <- sample(c(90, 95, 98, 99), size = 1)
df4       <- sample(5:25, size = 1)

SDNY <- read.csv("https://raw.githubusercontent.com/drs22Col/FedSentencing/master/Data/NYSouthernDistrictData.csv")
SDNYdrug <- SDNY[grepl("[D,d]rug", SDNY$Offense) & SDNY$SentenceMonths >= 0, ]
SDNYdrug$Race <- ifelse(SDNYdrug$Race == 1, "White", "Non-White")

whiteSentences    <- SDNYdrug$SentenceMonths[SDNYdrug$Race == "White"]
nonWhiteSentences <- SDNYdrug$SentenceMonths[SDNYdrug$Race == "Non-White"]

exercise_results <- list(
  notebook            = "Topic 15: Hypothesis Tests and Confidence Intervals for Numerical Data (Spreadsheets)",
  pt_ex1_left         = "unattempted",
  pt_ex2_right        = "unattempted",
  qt_ex3              = "unattempted",
  qt_ex4              = "unattempted",
  sentencing_ci_2     = "unattempted",
  sentencing_ci_4     = "unattempted",
  sentencing_ci_7     = "unattempted",
  sentencing_ci_8     = "unattempted",
  sentencing_ci_9     = "unattempted",
  sentencing_ht_9     = "unattempted",
  sentencing_ht_10    = "unattempted",
  sentencing_ht_11    = "unattempted",
  sentencing_ht_12    = "unattempted",
  sentencing_ht_13    = "unattempted",
  sentencing_ht_14    = "unattempted",
  sentencing_ht_16    = "unattempted",
  sentencing_ht_17    = "unattempted",
  sentencing_ht_18    = "unattempted",
  sentencing_ht_18_5  = "unattempted",
  sentencing_ht_19    = "unattempted"
)
```

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)

theme_set(theme_bw(base_size = 12))

set.seed(241)
boundary1 <- round(runif(1, 0.75, 2), 2)
df1       <- sample(5:25, size = 1)
boundary2 <- round(runif(1, -2, -1.5), 2)
df2       <- sample(5:25, size = 1)
area3     <- round(runif(1, 0.05, 0.95), 2)
df3       <- sample(5:25, size = 1)
clevel4   <- sample(c(90, 95, 98, 99), size = 1)
df4       <- sample(5:25, size = 1)

SDNY <- read.csv("https://raw.githubusercontent.com/drs22Col/FedSentencing/master/Data/NYSouthernDistrictData.csv")
SDNYdrug <- SDNY[grepl("[D,d]rug", SDNY$Offense) & SDNY$SentenceMonths >= 0, ]
SDNYdrug$Race <- ifelse(SDNYdrug$Race == 1, "White", "Non-White")

whiteSentences    <- SDNYdrug$SentenceMonths[SDNYdrug$Race == "White"]
nonWhiteSentences <- SDNYdrug$SentenceMonths[SDNYdrug$Race == "Non-White"]
```

:::{.callout-note}
## About

This activity introduces techniques for inference on a single population mean and for comparing two population means. The $t$-distributions are introduced, historical context is provided, and we work through two applications using federal sentencing data on drug-related convictions from the Southern District of New York State.

```{ojs}
//| echo: false

{
  const btn = html`<button style="
    background: #436b95;
    color: white;
    border: none;
    padding: 8px 16px;
    border-radius: 4px;
    cursor: pointer;
    font-size: 0.9em;
  ">🔄 Reset My Responses</button>`;

  btn.onclick = () => {
    if (confirm('Reset all responses for this activity? This cannot be undone.')) {
      Object.keys(localStorage)
        .filter(k => k.endsWith('_selected_t15e') || k.endsWith('_correct_t15e') || k.endsWith('_result_t15e'))
        .forEach(k => localStorage.removeItem(k));
      location.reload();
    }
  };

  return btn;
}
```

*Note.* The button above resets multiple choice and checkbox questions. Currently, resetting answer entry cells must be done manually via hitting the *Start Over* button on each individual interactive cell.

:::

:::{.callout-note}
## Using a Spreadsheet Program

Through the beginning of this activity, you'll be asked to use empty cells on a blank excel sheet for calculations. After this initial section, you'll work with actual criminal sentencing data from the Southern District of New York State. Use the links below to obtain a copy of the spreadsheet.

+ If you are a Google Sheets user, [click here to create a copy of the spreadsheet for this activity](https://docs.google.com/spreadsheets/d/11nyYaGNov9lBk2OFFFdfMsRfWRKOvEf8ajSgD8AVjrI/copy){target="_blank"}
+ If you are an Excel user, [click here to download a copy of the spreadsheet for this activity](https://docs.google.com/spreadsheets/d/11nyYaGNov9lBk2OFFFdfMsRfWRKOvEf8ajSgD8AVjrI/export?format=xlsx)

:::

## Hypothesis Testing and Confidence Intervals for Numerical Data

In this activity, we continue our exploration of statistical inference. Through the past few activities you've become more comfortable with hypothesis testing and confidence intervals for categorical variables. Here we extend those frameworks to numerical variables. We'll first be formally introduced to the family of $t$-distributions, and then work through a pair of applications to real criminal sentencing data.

As a reminder, the [*Standard Error Decision Tree*](https://agmath.github.io/SiteFiles/StdErrorDecisionTree.pdf){target="_blank"}, [*General Strategy for Conducting Hypothesis Tests*](https://agmath.github.io/SiteFiles/HTgeneral.pdf){target="_blank"}, and [*General Strategy for Constructing Confidence Intervals*](https://agmath.github.io/SiteFiles/CIgeneral.pdf){target="_blank"} are all available for reference. The walkthrough video for the decision tree is below, in case you'd like a refresher.

```{=html}
<center>
<iframe width="560" height="315" src="https://www.youtube.com/embed/5bAylfMIArg?start=16" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>
</center>
```

While watching the walkthrough video you probably noticed that the side of the decision tree corresponding to numerical data (inference for the mean, $\mu$) is more complex than the side for proportions. Much of this stems from the fact that using a sample standard deviation ($s$) as an approximation for the population standard deviation ($\sigma$) adds uncertainty to our estimate of the sampling distribution. To account for this added uncertainty, we utilize a class of *penalized* normal distributions called the $t$-distributions. Watch at least one of the introductory videos below.

**A Detailed Introduction:**

```{=html}
<center>
<iframe width="560" height="315" src="https://www.youtube.com/embed/bqfcFCjaE1c" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>
</center>
```

**A Shorter Introduction:**

```{=html}
<center>
<iframe width="560" height="315" src="https://www.youtube.com/embed/32CuxWdOlow" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>
</center>
```

## What Does the $t$-Distribution Look Like?

The videos above have helped us identify scenarios where we should utilize a $t$-distribution instead of the standard normal ($z$) distribution. There is a simple rule to follow: **any time we use a sample standard deviation as a proxy for the population standard deviation in our estimate for the standard error, we should use a $t$-distribution**.

The plot below shows the standard normal distribution in **black**, a $t$-distribution with 3 degrees of freedom in **red**, and a $t$-distribution with 12 degrees of freedom in **blue**.

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false

ggplot() + 
  geom_line(aes(x = seq(-3.5, 3.5, length.out = 200), 
                y = dnorm(seq(-3.5, 3.5, length.out = 200), 0, 1)), 
            color = "black", lwd = 1.25) + 
  geom_line(aes(x = seq(-3.5, 3.5, length.out = 200), 
                y = dt(seq(-3.5, 3.5, length.out = 200), 3)), 
            color = "red", lwd = 1.25) + 
  geom_line(aes(x = seq(-3.5, 3.5, length.out = 200), 
                y = dt(seq(-3.5, 3.5, length.out = 200), 12)), 
            color = "blue", lwd = 1.25) + 
  labs(title = "Normal and t-Distributions", x = "", y = "")
```

Notice that all three distributions are bell-shaped, but the $t$-distributions have **fatter tails** than the normal distribution. This means that the $t$-distributions have more area/probability in their tails than the standard normal distribution. Also notice that the $t$-distribution with 12 degrees of freedom is more similar to the standard normal distribution than the one with 3 degrees of freedom. As degrees of freedom increase, the $t$-distribution approaches (becomes more like) the normal distribution.

## Using the $t$-Distributions

When we introduced the normal distribution, we identified two helper functions that you could use in your spreadsheet formulas: `NORM.DIST()` to find probabilities (areas to the left of a boundary value) and `NORM.INV()` to find percentile cutoffs. We have analogous functions for the $t$-distribution:

+ **`=T.DIST(q, df, TRUE)`** — finds the probability of a randomly chosen observation falling to the *left* of the boundary value `q` in a $t$-distribution with `df` degrees of freedom.
+ **`=T.INV(p, df)`** — finds the cutoff value for which the area to the *left* of that cutoff in a $t$-distribution with `df` degrees of freedom is `p`.

Note that unlike `NORM.DIST()` and `NORM.INV()`, these functions have no mean or standard deviation arguments — we always work with standardized variables (via the test statistic formula) when using the $t$-distributions.

Let's practice. Don't forget to draw pictures — you are much more likely to make mistakes if you skip this step. Use the `BlankSheet` tab of the spreadsheet you copied at the beginning of this activity for your calculations.

---

**Question 1:** Find $\mathbb{P}\left[t <\right.$ `{ojs} boundary1` $\left.\right]$ in a $t$-distribution with `{ojs} df1` degrees of freedom.

In your spreadsheet, compute this probability and enter your result below.

```{webr}
#| exercise: pt-ex1-left
#| input:
#|   - boundary1
#|   - df1
# Enter your answer / result here

```

:::{.hint exercise="pt-ex1-left"}
:::{.callout-note collapse="false"}
## Hint 1

Start by drawing a $t$-distribution centered at 0. Where does `{ojs} boundary1` fall — to the left or right of center? Shade the region to the left of your boundary value. Should your answer be larger or smaller than 0.5?

:::
:::

:::{.hint exercise="pt-ex1-left"}
:::{.callout-note collapse="false"}
## Hint 2

`T.DIST()` with `TRUE` as the third argument gives the area to the left of your boundary value — exactly what we want here. In your spreadsheet, type `=T.DIST(___, ___, TRUE)` with the boundary value as the first argument and the degrees of freedom as the second.

:::
:::

:::{.hint exercise="pt-ex1-left"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

In your spreadsheet, type `=T.DIST(___, ___, TRUE)`, with the first argument filled in by `{ojs} boundary1` (the boundary value) and the second by `{ojs} df1` (the degrees of freedom). Enter the result below.

:::
:::

```{webr}
#| exercise: pt-ex1-left
#| solution: true
#| input:
#|   - boundary1
#|   - df1

pt(boundary1, df = df1)
```

```{webr}
#| exercise: pt-ex1-left
#| check: true
#| input:
#|   - boundary1
#|   - df1

exercise_results$pt_ex1_left <<- "fail"
gradethis::grade_this({
  if (abs(.result - pt(boundary1, df = df1)) < 0.001) {
    exercise_results$pt_ex1_left <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

---

**Question 2:** Find $\mathbb{P}\left[t >\right.$ `{ojs} boundary2` $\left.\right]$ in a $t$-distribution with `{ojs} df2` degrees of freedom.

In your spreadsheet, compute this probability and enter your result below.

```{webr}
#| exercise: pt-ex2-right
#| input:
#|   - boundary2
#|   - df2
# Enter your answer / result here

```

:::{.hint exercise="pt-ex2-right"}
:::{.callout-note collapse="false"}
## Hint 1

Draw the $t$-distribution and shade the region to the *right* of `{ojs} boundary2`. Note that `{ojs} boundary2` is negative — it falls to the left of center. Should your answer be larger or smaller than 0.5?

:::
:::

:::{.hint exercise="pt-ex2-right"}
:::{.callout-note collapse="false"}
## Hint 2

`T.DIST()` with `TRUE` always gives the area to the *left*. For a right-tail probability, subtract that result from 1.

:::
:::

:::{.hint exercise="pt-ex2-right"}
:::{.callout-note collapse="false"}
## Hint 3

In your spreadsheet, type `=1 - T.DIST(___, ___, TRUE)`. Can you fill in the blanks?

:::
:::

:::{.hint exercise="pt-ex2-right"}
:::{.callout-note collapse="false"}
## Hint 4 (Solved)

In your spreadsheet, type `=1 - T.DIST(___, ___, TRUE)`, where the first blank is filled by `{ojs} boundary2` (the boundary value) and the second by `{ojs} df2` (the degrees of freedom). Enter the result in the code block.

:::
:::

```{webr}
#| exercise: pt-ex2-right
#| solution: true
#| input:
#|   - boundary2
#|   - df2

1 - pt(boundary2, df = df2)
```

```{webr}
#| exercise: pt-ex2-right
#| check: true
#| input:
#|   - boundary2
#|   - df2

exercise_results$pt_ex2_right <<- "fail"
gradethis::grade_this({
  if (abs(.result - (1 - pt(boundary2, df = df2))) < 0.001) {
    exercise_results$pt_ex2_right <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

---

**Question 3:** Find the cutoff value in a $t$-distribution with `{ojs} df3` degrees of freedom for which the area to the *left* of the cutoff is `{ojs} area3`.

In your spreadsheet, compute this cutoff and enter your result below.

```{webr}
#| exercise: qt-ex3
#| input:
#|   - area3
#|   - df3
# Enter your answer / result here

```

:::{.hint exercise="qt-ex3"}
:::{.callout-note collapse="false"}
## Hint 1

We are looking for a boundary (cutoff) value rather than a probability. Which function gives us a boundary value — `T.DIST()` or `T.INV()`?

:::
:::

:::{.hint exercise="qt-ex3"}
:::{.callout-note collapse="false"}
## Hint 2

Use `T.INV()` since we want a boundary value.

:::
:::

:::{.hint exercise="qt-ex3"}
:::{.callout-note collapse="false"}
## Hint 3

Draw a picture of the $t$-distribution with `{ojs} area3` of the area shaded to the left of a cutoff value. Should your cutoff value be positive or negative? How do you know?

:::
:::

:::{.hint exercise="qt-ex3"}
:::{.callout-note collapse="false"}
## Hint 4

Since the area to the left ( `{ojs} area3` ) is more than 0.5, the boundary value we are looking for must be to the right of the middle of the distribution. The $t$-distributions are all centered at 0, so our boundary value must be positive.

:::
:::

:::{.hint exercise="qt-ex3"}
:::{.callout-note collapse="false"}
## Hint 5

In a blank cell of your spreadsheet, use `=T.INV(___, ___)` where the first argument is the area to the left of the boundary value we are seeking and the second argument is the degrees of freedom for the $t$-distribution.

:::
:::

:::{.hint exercise="qt-ex3"}
:::{.callout-note collapse="false"}
## Hint 6 (Solved)

In a blank cell of your spreadsheet, use `=T.INV(___, ___)` where the first argument is `{ojs} area3` (the area to the left of the boundary value we are seeking) and the second argument is `{ojs} df3` (the degrees of freedom for the $t$-distribution).

:::
:::

```{webr}
#| exercise: qt-ex3
#| solution: true
#| input:
#|   - area3
#|   - df3

qt(area3, df = df3)
```

```{webr}
#| exercise: qt-ex3
#| check: true
#| input:
#|   - area3
#|   - df3

exercise_results$qt_ex3 <<- "fail"
gradethis::grade_this({
  if (abs(.result - qt(area3, df = df3)) < 0.005) {
    exercise_results$qt_ex3 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

---

**Question 4:** Find the critical value associated with a `{ojs} clevel4`% confidence interval using a $t$-distribution with `{ojs} df4` degrees of freedom.

In your spreadsheet, compute this critical value and enter your result below.

```{webr}
#| exercise: qt-ex4
#| input:
#|   - clevel4
#|   - df4
# Enter your answer / result here

```

:::{.hint exercise="qt-ex4"}
:::{.callout-note collapse="false"}
## Hint 1

Start by drawing a $t$-distribution.

:::
:::

:::{.hint exercise="qt-ex4"}
:::{.callout-note collapse="false"}
## Hint 2

A `{ojs} clevel4`% confidence interval captures the middle `{ojs} clevel4`% of the distribution. How much area is left over for the two tails combined, and how much is in each tail individually?

:::
:::

:::{.hint exercise="qt-ex4"}
:::{.callout-note collapse="false"}
## Hint 3

The remaining area is `{ojs} (1 - clevel4/100).toFixed(2)`, which will be split equally between the two tails. This means that each tail contains an area of `{ojs} ((1 - clevel4/100)/2).toFixed(3)`.

:::
:::

:::{.hint exercise="qt-ex4"}
:::{.callout-note collapse="false"}
## Hint 4

The remaining area is `{ojs} (1 - clevel4/100).toFixed(2)`, which will be split equally between the two tails. This means that each tail contains and area of `{ojs} ((1 - clevel4/100)/2).toFixed(3)`. 

The critical value is the boundary separating the upper tail from the rest of the distribution. What is the total area to the *left* of that boundary?

:::
:::

:::{.hint exercise="qt-ex4"}
:::{.callout-note collapse="false"}
## Hint 5

The area to the left of the critical value is the lower tail area plus the middle area: `{ojs} ((1 - clevel4/100)/2).toFixed(3)` + `{ojs} (clevel4/100).toFixed(2)` = `{ojs} ((1 - clevel4/100)/2 + clevel4/100).toFixed(3)`. 

Use `T.INV()` with this area and the degrees of freedom.

:::
:::

:::{.hint exercise="qt-ex4"}
:::{.callout-note collapse="false"}
## Hint 6 (Solved)

In your spreadsheet, type `=T.INV(___, ___)`, with the first blank filled by `{ojs} ((1 - clevel4/100)/2 + clevel4/100).toFixed(3)` and the second blank filled by `{ojs} df4`. Enter the result in the code cell.

:::
:::

```{webr}
#| exercise: qt-ex4
#| solution: true
#| input:
#|   - clevel4
#|   - df4

qt(1 - (1 - clevel4/100)/2, df = df4)
```

```{webr}
#| exercise: qt-ex4
#| check: true
#| input:
#|   - clevel4
#|   - df4

exercise_results$qt_ex4 <<- "fail"
gradethis::grade_this({
  if (abs(.result - qt(1 - (1 - clevel4/100)/2, df = df4)) < 0.005) {
    exercise_results$qt_ex4 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

:::{.callout-tip}
## Summary: Working with the $t$-Distributions

We've now practiced with two spreadsheet new functions for working with $t$-distributions:

+ **`=T.DIST(q, df, TRUE)`** — returns the area to the **left** of the boundary value `q` in a $t$-distribution with `df` degrees of freedom. This is the $t$-distribution analog of `=NORM.DIST(q, 0, 1, TRUE)`. It works correctly for both positive and negative boundary values.
+ **`=T.INV(p, df)`** — returns the boundary value for which the area to the **left** is `p` in a $t$-distribution with `df` degrees of freedom. This is the $t$-distribution analog of `=NORM.INV(p, 0, 1)`.

As with the normal distribution, you will need to think carefully about how to use these functions to obtain the quantities (areas/probabilities or boundary values) that you are looking for. Drawing a picture will help you.

:::

## Applications to Criminal Sentencing

We'll work with a [dataset on Federal Sentencing from the Southern District of New York State](https://github.com/drs22Col/FedSentencing){target="_blank"}. The full dataset is available to you on the `SDNYfull` tab of the spreadsheet you copied at the beginning of this activity, and the records filtered to only include sentences from drug-related convictions is available on the `SDNYdrug` tab. 

Open you spreadsheet and switch to the `SDNYdrug` tab. You'll complete the work for the next scenarios on that tab.

:::{.callout-tip}
## Spreadsheet Tips to Remember

You'll benefit by recalling and refreshing on the following items. Check back in on the [Topic 2: Introduction to Spreadsheets](https://agmath.github.io/OpenIntroCompanions/02_IntroToSpreadsheets.html) and [Topic 3: Descriptive Statistics](https://agmath.github.io/OpenIntroCompanions/03_DescriptiveStatistics_Spreadsheet.html) activities for more details.

+ **Working with ranges of cells** is done via `top_left_cell:bottom_right_cell`. That is, `C2:F23` selects all the cells in the rectangular grid from cell `C2` (third column, second row) to cell `F23` (sixth colum, twenty-third row). Similarly, `C2:C500` selects all the cells in column `C` between row 2 and row 500.
+ **Spreadsheet software has built-in functions** like `AVERAGE()`, `STDEV()`, `COUNT()`, and more that allow you to quickly run calculations over a range.
+ **Creating grouped summaries** can be done via *Pivot Tables*. You create a pivot table by clicking on a cell inside of you data range, navigating to *Insert*, and then choosing *Pivot Table*.

There will be robust hint sequences to help you along the way when you get stuck.
:::

---

### Task 1: Confidence Interval for Average Sentence Length

Construct a 95% confidence interval for the average sentence length for a drug-related conviction in the Southern District of New York State.

:::{.callout-caution}
## Check Your Understanding: Sentencing CI, Part I

To answer the question as asked, we should:

```{ojs}
//| echo: false

mutable ok_response = (response, n) => { return html`Loading...` };
viewof q1 = Inputs.radio(
  new Map([
    ["Compute a probability.", 1],
    ["Construct a confidence interval.", 2],
    ["Conduct a hypothesis test.", 3],
    ["Calculate a required sample size.", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q1_selected_t15e") ?? "null")}
);

{
  localStorage.setItem("q1_selected_t15e", JSON.stringify(q1));
  localStorage.setItem("q1_correct_t15e", "2");
  localStorage.setItem("q1_result_t15e", q1 === null ? "unattempted" : (q1 == 2 ? "correct" : "incorrect"));
}
ok_response(q1, "2");
```

:::

Use an empty cell on the `SDNYdrug` tab of your spreadsheet to calculate the average sentence length for drug-related convictions in the Southern District of New York State. Enter your result in the code cell below.

```{webr}
#| exercise: sentencing-ci-2

```

:::{.hint exercise="sentencing-ci-2"}
:::{.callout-note collapse="false"}
## Hint 1

Open the `SDNYdrug` tab of the spreadsheet and find a blank cell to type your formula into.

:::
:::

:::{.hint exercise="sentencing-ci-2"}
:::{.callout-note collapse="false"}
## Hint 2

You want to calculate an average, so your formula will take the form `=AVERAGE(___)`. What is the range of cells you are averaging over?

:::
:::

:::{.hint exercise="sentencing-ci-2"}
:::{.callout-note collapse="false"}
## Hint 3

You want to calculate an average, so your formula will take the form `=AVERAGE(___)`. What is the range of cells you are averaging over?

The sentence lengths, measured in months, are in column `F` of the sheet.

:::
:::

:::{.hint exercise="sentencing-ci-2"}
:::{.callout-note collapse="false"}
## Hint 4

You want to calculate an average, so your formula will take the form `=AVERAGE(___)`. What is the range of cells you are averaging over?

The sentence lengths, measured in months, are in column `F` of the sheet and the values are populating rows 2 through 281.

:::
:::

:::{.hint exercise="sentencing-ci-2"}
:::{.callout-note collapse="false"}
## Hint 5

You want to calculate an average, so your formula will take the form `=AVERAGE(___)`. What is the range of cells you are averaging over?

The sentence lengths, measured in months, are in column `F` of the sheet and the values are populating rows 2 through 281. The range is `F2:F281`.

:::
:::

:::{.hint exercise="sentencing-ci-2"}
:::{.callout-note collapse="false"}
## Hint 6 (Solved)

Calculate the average sentence length in a blank cell using `=AVERAGE(F2:F281)`. Enter the result in the code cell.

Don't forget to label the cell so that you don't forget what it contains.

:::
:::

```{webr}
#| exercise: sentencing-ci-2
#| solution: true
mean(SDNYdrug$SentenceMonths)
```

```{webr}
#| exercise: sentencing-ci-2
#| check: true

exercise_results$sentencing_ci_2 <<- "fail"
gradethis::grade_this({
  if (abs(.result - .solution) < 0.005) {
    exercise_results$sentencing_ci_2 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```



:::{.callout-caution}
## Check Your Understanding: Sentencing CI, Part II

The standard error formula is:

```{ojs}
//| echo: false

viewof q2 = Inputs.radio(
  new Map([
    ["SE = sqrt(p(1-p)/n)", 1],
    ["SE = σ/sqrt(n)", 2],
    ["SE = s/sqrt(n)", 3],
    ["SE = sqrt(p1(1-p1)/n1 + p2(1-p2)/n2)", 4],
    ["SE = sqrt(s1²/n1 + s2²/n2)", 5]
  ]),
  {value: JSON.parse(localStorage.getItem("q2_selected_t15e") ?? "null")}
);

{
  localStorage.setItem("q2_selected_t15e", JSON.stringify(q2));
  localStorage.setItem("q2_correct_t15e", "3");
  localStorage.setItem("q2_result_t15e", q2 === null ? "unattempted" : (q2 == 3 ? "correct" : "incorrect"));
}
ok_response(q2, "3");
```

:::{.callout-note collapse="true"}
## Hint 1

Open the [*Standard Error Decision Tree*](www/StdErrorDecisionTree.pdf){target="_blank"} and start from the top of the tree.

:::{.callout-note collapse="true"}
## Hint 2

Are we conducting inference on a mean ($\mu$) or a proportion ($p$)?

:::{.callout-note collapse="true"}
## Hint 3

The question "*How long was the sentence associated with this conviction?*" is answered with a number of months. This is numerical data that is best summarized by a *mean*. We're on the right side of the tree.

:::{.callout-note collapse="true"}
## Hint 4

How many groups/populations are we comparing? Do we know the population standard deviation, $\sigma$?

:::{.callout-note collapse="true"}
## Hint 5

We have just a single population here -- all of the individual sentences belong to one group. We don't know the population standard deviation in sentence lengths.

Note that we can calculate the standard deviation in sentence lengths, but that will only the the standard deviation for the sentences in our sample ($s$).

What is the standard error ($S_E$) from the box you landed in?

:::
:::
:::
:::
:::

:::

Using your spreadsheet, calculate the standard error for the average sentence length.

```{webr}
#| exercise: sentencing-ci-4

```

:::{.hint exercise="sentencing-ci-4"}
:::{.callout-note collapse="false"}
## Hint 1

Find a blank cell and calculate the standard deviation in sentence lengths first.

:::
:::

:::{.hint exercise="sentencing-ci-4"}
:::{.callout-note collapse="false"}
## Hint 2

You can use `=STDEV(___)` to calculate the standard deviation. What is the range of values you're calculating the standard deviation over?

:::
:::

:::{.hint exercise="sentencing-ci-4"}
:::{.callout-note collapse="false"}
## Hint 3

We're still working with the sentence lengths in months, so the range is the same as it was when we calculated the average sentence length. Use `=STDEV(F2:F281)` to find the standard deviation and label this cell so that you don't forget what the number means.

:::
:::

:::{.hint exercise="sentencing-ci-4"}
:::{.callout-note collapse="false"}
## Hint 4

Now find the number of observations. 

:::
:::

:::{.hint exercise="sentencing-ci-4"}
:::{.callout-note collapse="false"}
## Hint 5 

Now find the number of observations. There are a few ways to do this:

+ Recall that your range spanned the cells `F2:F281` -- there are 280 cells in that range.
+ Highlight all the cells in the range `F2:F281`. Your spreasheet software should tell you the number of cells you have selected near the bottom-right of your window, just below the spreadsheet area.
+ $\bigstar$ A formula using `=COUNT(F2:F281)` will count all of the cells in the range that contain numeric values. This is the best approach because the others will ignore missing values and your count may be too high.

:::
:::

:::{.hint exercise="sentencing-ci-4"}
:::{.callout-note collapse="false"}
## Hint 6

Now that you have the standard deviation in sentence lengths (about 53.9147 months) and the total sample size (280 observations), compute the standard error using the formula $S_E = s/\sqrt{n}$

:::
:::

:::{.hint exercise="sentencing-ci-4"}
:::{.callout-note collapse="false"}
## Hint 7 (Solved)

Using a blank cell, type `=53.9147/SQRT(280)` (or even better, do the calculation by referencing the cells containing the standard deviation and sample size). Be sure to label the cell and then type the resulting value into the code block.

:::
:::


```{webr}
#| exercise: sentencing-ci-4
#| solution: true

sd(SDNYdrug$SentenceMonths) / sqrt(nrow(SDNYdrug))
```

```{webr}
#| exercise: sentencing-ci-4
#| check: true

exercise_results$sentencing_ci_4 <<- "fail"
gradethis::grade_this({
  if (abs(.result - .solution) < 0.005) {
    exercise_results$sentencing_ci_4 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

:::{.callout-caution}
## Check Your Understanding: Sentencing CI, Part III

The distribution to be used is:

```{ojs}
//| echo: false

viewof q3 = Inputs.radio(
  new Map([
    ["Normal.", 1],
    ["t-distribution with df = n - 1.", 2],
    ["t-distribution with df = min(n1-1, n2-1).", 3],
    ["t-distribution with df = n_diff - 1.", 4],
    ["We should not use either distribution.", 5]
  ]),
  {value: JSON.parse(localStorage.getItem("q3_selected_t15e") ?? "null")}
);

{
  localStorage.setItem("q3_selected_t15e", JSON.stringify(q3));
  localStorage.setItem("q3_correct_t15e", "2");
  localStorage.setItem("q3_result_t15e", q3 === null ? "unattempted" : (q3 == 2 ? "correct" : "incorrect"));
}
ok_response(q3, "2");
```

:::{.callout-note collapse="true"}
## Hint 1

The [*Standard Error Decision Tree](www/StdErrorDecisionTree.pdf){target="_blank"} tells you which distribution to use.

:::{.callout-note collapse="true"}
## Hint 2

Check the box you landed in when finding the standard error ($S_E$) formula.

:::{.callout-note collapse="true"}
## Hint 3

If that box is not a specialty box (ANOVA or $\chi^2$) and it contains information about degrees of freedom ($df$), then you're using a $t$-distribution with that number of degrees of freedom. Otherwise, you're using the normal distribution.

:::
:::
:::

:::

:::{.callout-caution}
## Check Your Understanding: Sentencing CI, Part IV

The desired level of confidence is:

```{ojs}
//| echo: false

viewof q4 = Inputs.radio(
  new Map([
    ["90%", 1],
    ["0.95%", 2],
    ["95%", 3],
    ["98%", 4],
    ["99%", 5]
  ]),
  {value: JSON.parse(localStorage.getItem("q4_selected_t15e") ?? "null")}
);

{
  localStorage.setItem("q4_selected_t15e", JSON.stringify(q4));
  localStorage.setItem("q4_correct_t15e", "3");
  localStorage.setItem("q4_result_t15e", q4 === null ? "unattempted" : (q4 == 3 ? "correct" : "incorrect"));
}
ok_response(q4, "3");
```

:::

Now use your spreadsheet to compute the critical value for the confidence interval and enter your result below.

```{webr}
#| exercise: sentencing-ci-7
# Enter your critical value here

```

:::{.hint exercise="sentencing-ci-7"}
:::{.callout-note collapse="false"}
## Hint 1

You'll need to use `T.INV()`.

:::
:::

:::{.hint exercise="sentencing-ci-7"}
:::{.callout-note collapse="false"}
## Hint 2

You'll nedd to use `T.INV()`. For a 95% confidence interval, the middle 95% of the distribution is captured, leaving 5% split equally between the two tails. Draw a picture and determine the total area to the *left* of the upper-tail boundary value (the critical value).

:::
:::

:::{.hint exercise="sentencing-ci-7"}
:::{.callout-note collapse="false"}
## Hint 3

Each tail holds 2.5% of the area, so the total area to the left of the critical value is 0.025 + 0.95 = 0.975. The degrees of freedom is $n - 1$ where $n$ is your sample size.

:::
:::

:::{.hint exercise="sentencing-ci-7"}
:::{.callout-note collapse="false"}
## Hint 4 (Solved)

With $n = 280$, the degrees of freedom is 279. In your spreadsheet, type `=T.INV(0.975, 279)`. Label the cell and input the result into the code block.
:::
:::

```{webr}
#| exercise: sentencing-ci-7
#| solution: true

qt(0.975, df = nrow(SDNYdrug) - 1)
```

```{webr}
#| exercise: sentencing-ci-7
#| check: true

exercise_results$sentencing_ci_7 <<- "fail"
gradethis::grade_this({
  if (abs(.result - .solution) < 0.005) {
    exercise_results$sentencing_ci_7 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

:::{.callout-tip}
## A Note on the Critical Value

With so many observations, the correct critical value differs very little from 1.96 (the value obtained from the normal distribution). Remember that the critical values on the Standard Error Decision Tree are for the normal distribution only. Any time we use a sample standard deviation as a stand-in for the population standard deviation, we should use `T.INV()` — this will make a real difference when sample sizes are smaller.
:::

Now use your spreadsheet to compute the lower bound of the 95% confidence interval and enter your result below.

```{webr}
#| exercise: sentencing-ci-8
# Enter your lower bound here

```

:::{.hint exercise="sentencing-ci-8"}
:::{.callout-note collapse="false"}
## Hint 1

The confidence interval formula is $(\text{point estimate}) \pm (\text{critical value}) \times S_E$. For the lower bound, we subtract the margin of error from the point estimate.

:::
:::

:::{.hint exercise="sentencing-ci-8"}
:::{.callout-note collapse="false"}
## Hint 2

The confidence interval formula is $(\text{point estimate}) \pm (\text{critical value}) \times S_E$.

You have all three values from earlier steps: the point estimate (~42.75 months), the critical value (~1.97), and the standard error (~3.22 months). In your spreadsheet, type `=42.75-(1.97*3.22)` and press Enter.

:::
:::

```{webr}
#| exercise: sentencing-ci-8
#| solution: true

mean(SDNYdrug$SentenceMonths) - 
  qt(0.975, df = nrow(SDNYdrug) - 1) * 
  (sd(SDNYdrug$SentenceMonths) / sqrt(nrow(SDNYdrug)))
```

```{webr}
#| exercise: sentencing-ci-8
#| check: true

exercise_results$sentencing_ci_8 <<- "fail"
gradethis::grade_this({
  lower <- mean(SDNYdrug$SentenceMonths) - 
    qt(0.975, df = nrow(SDNYdrug) - 1) * 
    (sd(SDNYdrug$SentenceMonths) / sqrt(nrow(SDNYdrug)))
  if (abs(.result - lower) < 0.1) {
    exercise_results$sentencing_ci_8 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

Now compute the upper bound and enter your result below.

```{webr}
#| exercise: sentencing-ci-9
# Enter your upper bound here

```

:::{.hint exercise="sentencing-ci-9"}
:::{.callout-note collapse="false"}
## Hint 1

Same approach as the lower bound — just add the margin of error instead of subtracting it. In your spreadsheet, type `=42.75+(1.97*3.22)` and press Enter.

:::
:::

```{webr}
#| exercise: sentencing-ci-9
#| solution: true

mean(SDNYdrug$SentenceMonths) + 
  qt(0.975, df = nrow(SDNYdrug) - 1) * 
  (sd(SDNYdrug$SentenceMonths) / sqrt(nrow(SDNYdrug)))
```

```{webr}
#| exercise: sentencing-ci-9
#| check: true

exercise_results$sentencing_ci_9 <<- "fail"
gradethis::grade_this({
  upper <- mean(SDNYdrug$SentenceMonths) + 
    qt(0.975, df = nrow(SDNYdrug) - 1) * 
    (sd(SDNYdrug$SentenceMonths) / sqrt(nrow(SDNYdrug)))
  if (abs(.result - upper) < 0.1) {
    exercise_results$sentencing_ci_9 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

:::{.callout-caution}
## Check Your Understanding: Sentencing CI, Part V

Which of the following is the correct interpretation of this confidence interval?

```{ojs}
//| echo: false

viewof q5 = Inputs.radio(
  new Map([
    ["95% of sentence lengths for drug-related charges in the Southern District of New York State will fall between the lower and upper bounds.", 1],
    ["We are 95% confident that the average sentence length for drug-related crimes in this sample falls between the lower and upper bounds.", 2],
    ["There is a 95% chance a criminal charged with a drug-related offense will receive a sentence between the lower and upper bounds.", 3],
    ["We are 95% confident that the true average sentence length for drug-related offenders in the Southern District of New York State is between the lower and upper bounds calculated.", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q5_selected_t15e") ?? "null")}
);

{
  localStorage.setItem("q5_selected_t15e", JSON.stringify(q5));
  localStorage.setItem("q5_correct_t15e", "4");
  localStorage.setItem("q5_result_t15e", q5 === null ? "unattempted" : (q5 == 4 ? "correct" : "incorrect"));
}
ok_response(q5, "4");
```

:::

:::{.callout-caution}
## Check Your Understanding: Sentencing CI, Part VI

Does this sample provide evidence to suggest that the average sentence length for drug-related charges in the Southern District of New York State exceeds three years (36 months)?

```{ojs}
//| echo: false

viewof q6 = Inputs.radio(
  new Map([
    ["No. The confidence interval shows that an average sentence below 36 months is plausible.", 1],
    ["Yes. The confidence interval for the average sentence length includes only values exceeding 36 months.", 2],
    ["No. Some sentence lengths are below 36 months while others exceed 36 months.", 3],
    ["It is impossible to say, since a different sample would result in a different confidence interval.", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q6_selected_t15e") ?? "null")}
);

{
  localStorage.setItem("q6_selected_t15e", JSON.stringify(q6));
  localStorage.setItem("q6_correct_t15e", "2");
  localStorage.setItem("q6_result_t15e", q6 === null ? "unattempted" : (q6 == 2 ? "correct" : "incorrect"));
}
ok_response(q6, "2");
```

:::

---

### Question 2: Hypothesis Test for Difference in Sentence Lengths

Conduct a hypothesis test at the $\alpha = 0.10$ level of significance to determine whether the sample data provides significant evidence that the average sentence length for white offenders and the average sentence length for non-white offenders differs for drug-related cases in the Southern District of New York State.

:::{.callout-caution}
## Check Your Understanding: Sentencing HT, Part I

To answer the question as asked, we should:

```{ojs}
//| echo: false

viewof q7 = Inputs.radio(
  new Map([
    ["Compute a probability.", 1],
    ["Construct a confidence interval.", 2],
    ["Conduct a hypothesis test.", 3],
    ["Calculate a required sample size.", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q7_selected_t15e") ?? "null")}
);

{
  localStorage.setItem("q7_selected_t15e", JSON.stringify(q7));
  localStorage.setItem("q7_correct_t15e", "3");
  localStorage.setItem("q7_result_t15e", q7 === null ? "unattempted" : (q7 == 3 ? "correct" : "incorrect"));
}
ok_response(q7, "3");
```

:::

:::{.callout-caution}
## Check Your Understanding: Sentencing HT, Part II

What is the level of significance associated with this test?

```{ojs}
//| echo: false

viewof q8 = Inputs.radio(
  new Map([
    ["α = 0.01", 1],
    ["α = 0.05", 2],
    ["α = 0.10", 3],
    ["The p-value.", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q8_selected_t15e") ?? "null")}
);

{
  localStorage.setItem("q8_selected_t15e", JSON.stringify(q8));
  localStorage.setItem("q8_correct_t15e", "3");
  localStorage.setItem("q8_result_t15e", q8 === null ? "unattempted" : (q8 == 3 ? "correct" : "incorrect"));
}
ok_response(q8, "3");
```

:::

:::{.callout-caution}
## Check Your Understanding: Sentencing HT, Part III

Does this hypothesis test involve testing a statement about a mean ($\mu$), a proportion ($p$), or something else?

```{ojs}
//| echo: false

viewof q9 = Inputs.radio(
  new Map([
    ["One or more means.", 1],
    ["One or more proportions.", 2],
    ["Something else altogether.", 3]
  ]),
  {value: JSON.parse(localStorage.getItem("q9_selected_t15e") ?? "null")}
);

{
  localStorage.setItem("q9_selected_t15e", JSON.stringify(q9));
  localStorage.setItem("q9_correct_t15e", "1");
  localStorage.setItem("q9_result_t15e", q9 === null ? "unattempted" : (q9 == 1 ? "correct" : "incorrect"));
}
ok_response(q9, "1");
```

:::

:::{.callout-caution}
## Check Your Understanding: Sentencing HT, Part IV

How many groups are being compared in this test?

```{ojs}
//| echo: false

viewof q10 = Inputs.radio(
  new Map([
    ["The test involves only a single group.", 1],
    ["The test compares two groups.", 2],
    ["The test compares more than two groups.", 3]
  ]),
  {value: JSON.parse(localStorage.getItem("q10_selected_t15e") ?? "null")}
);

{
  localStorage.setItem("q10_selected_t15e", JSON.stringify(q10));
  localStorage.setItem("q10_correct_t15e", "2");
  localStorage.setItem("q10_result_t15e", q10 === null ? "unattempted" : (q10 == 2 ? "correct" : "incorrect"));
}
ok_response(q10, "2");
```

:::

:::{.callout-caution}
## Check Your Understanding: Sentencing HT, Part V

Which of the following are the hypotheses associated with this test?

```{ojs}
//| echo: false

viewof q11 = Inputs.radio(
  new Map([
    ["H₀: μ_white − μ_nonwhite = 0; Hₐ: μ_white − μ_nonwhite ≠ 0", 1],
    ["H₀: μ_white − μ_nonwhite = 0; Hₐ: μ_white − μ_nonwhite > 0", 2],
    ["H₀: μ_white − μ_nonwhite = 0; Hₐ: μ_white − μ_nonwhite < 0", 3],
    ["H₀: μ_white ≠ μ_nonwhite; Hₐ: μ_white = μ_nonwhite", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q11_selected_t15e") ?? "null")}
);

{
  localStorage.setItem("q11_selected_t15e", JSON.stringify(q11));
  localStorage.setItem("q11_correct_t15e", "1");
  localStorage.setItem("q11_result_t15e", q11 === null ? "unattempted" : (q11 == 1 ? "correct" : "incorrect"));
}
ok_response(q11, "1");
```

:::{.callout-note collapse="true"}
## Hint 1

The null hypothesis ($H_0$) is a skeptical claim -- "there's nothing to see here", while the alternative hypothesis ($H_a$) is the claim to be tested.

:::{.callout-note collapse="true"}
## Hint 2

The claim we're investigating asks whether or not there is a *difference* in the average sentence lengths.

:::
:::

:::

:::{.callout-caution}
## Check Your Understanding: Sentencing HT VI

Do we know the population standard deviations ($\sigma$) for sentence lengths in each group?

```{ojs}
//| echo: false

viewof q12 = Inputs.radio(
  new Map([
    ["Yes, we can compute the population standard deviation from the data using STDEV().", 1],
    ["Yes, we are provided the population standard deviation.", 2],
    ["No — STDEV() gives us sample standard deviations, not population standard deviations.", 3]
  ]),
  {value: JSON.parse(localStorage.getItem("q12_selected_t15e") ?? "null")}
);

{
  localStorage.setItem("q12_selected_t15e", JSON.stringify(q12));
  localStorage.setItem("q12_correct_t15e", "3");
  localStorage.setItem("q12_result_t15e", q12 === null ? "unattempted" : (q12 == 3 ? "correct" : "incorrect"));
}
ok_response(q12, "3");
```

:::

:::{.callout-caution}
## Check Your Understanding: Sentencing HT, Part VII

Are the observations in the two groups (sentences handed to white offenders and sentences handed to non-white offenders) paired?

```{ojs}
//| echo: false

viewof q13 = Inputs.radio(
  new Map([
    ["No. There is no reason to suggest that sentences are paired.", 1],
    ["Yes. Every sentence handed to a white offender can be naturally paired with a sentence handed to a non-white offender.", 2]
  ]),
  {value: JSON.parse(localStorage.getItem("q13_selected_t15e") ?? "null")}
);

{
  localStorage.setItem("q13_selected_t15e", JSON.stringify(q13));
  localStorage.setItem("q13_correct_t15e", "1");
  localStorage.setItem("q13_result_t15e", q13 === null ? "unattempted" : (q13 == 1 ? "correct" : "incorrect"));
}
ok_response(q13, "1");
```

:::

:::{.callout-caution}
## Check Your Understanding: Sentencing HT, Part VIII

Which standard error formula should be used?

```{ojs}
//| echo: false

viewof q14 = Inputs.radio(
  new Map([
    ["SE = s/sqrt(n)", 1],
    ["SE = sqrt(s1²/n1 + s2²/n2)", 2],
    ["SE = sqrt(σ1²/n1 + σ2²/n2)", 3],
    ["SE = sqrt(p(1-p)/n)", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q14_selected_t15e") ?? "null")}
);

{
  localStorage.setItem("q14_selected_t15e", JSON.stringify(q14));
  localStorage.setItem("q14_correct_t15e", "2");
  localStorage.setItem("q14_result_t15e", q14 === null ? "unattempted" : (q14 == 2 ? "correct" : "incorrect"));
}
ok_response(q14, "2");
```

:::{.callout-note collapse="true"}
## Hint 1

Walk through the standard error decision tree using the answers to the questions above.

:::

:::

:::{.callout-caution}
## Check Your Understanding: Sentencing HT IX

Which distribution does the test statistic follow?

```{ojs}
//| echo: false

viewof q15 = Inputs.radio(
  new Map([
    ["The normal distribution.", 1],
    ["The t-distribution with df = n - 1.", 2],
    ["The t-distribution with df = min(n1 - 1, n2 - 1).", 3],
    ["The t-distribution with df = n_diff - 1.", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q15_selected_t15e") ?? "null")}
);

{
  localStorage.setItem("q15_selected_t15e", JSON.stringify(q15));
  localStorage.setItem("q15_correct_t15e", "3");
  localStorage.setItem("q15_result_t15e", q15 === null ? "unattempted" : (q15 == 3 ? "correct" : "incorrect"));
}
ok_response(q15, "3");
```

:::{.callout-note collapse="true"}
## Hint 1

The box containing your standard error ($S_E$) tells you what distribution to use.

:::{.callout-note collapse="true"}
## Hint 2

If your box contains information about degrees of freedom ($df$), then you're using a $t$ distribution with that number of degrees of freedom.

:::
:::

:::

Walk through the following steps to calculate the values you'll need to construct your *test statistic*.

1. Open the `SDNYdrug` tab of your spreadsheet.
2. Click into the data range -- somewhere in columns `A` through `F`.
3. Use *Insert* and choose *Pivot Table*.
4. Verify that the range contains cells from `A1:F281`, and approve creating the pivot table in a new sheet.
5. Drag the `Race` column to the *Rows* field for the pivot table.
6. Drag the `SentenceMonths` column to the *Values* field for the pivot table.
7. Change the default summary value to `COUNT`. In Google Sheets, you can use the dropdown menu for this. In Excel, click the small "info" icon to the right of the summary value name in the *Values* field and then choose to summarize by *Count*.
8. Repeat steps 6 and 7 twice. You'll drag the `SentenceMonths` column to the *Values* field and update it to compute the `AVERAGE` and then again to compute the standard deviation (`STDEV`).


**Number of white offenders:** Input the number of white offenders whose sentences we have in our sample below.

```{webr}
#| exercise: sentencing-ht-9

```

```{webr}
#| exercise: sentencing-ht-9
#| solution: true

length(whiteSentences)
```

```{webr}
#| exercise: sentencing-ht-9
#| check: true

exercise_results$sentencing_ht_9 <<- "fail"
gradethis::grade_this({
  if (.result == length(whiteSentences)) {
    exercise_results$sentencing_ht_9 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

**Number of non-white offenders:** Input the number of non-white offenders whose sentences we have in our sample below.

```{webr}
#| exercise: sentencing-ht-10

```

```{webr}
#| exercise: sentencing-ht-10
#| solution: true

length(nonWhiteSentences)
```

```{webr}
#| exercise: sentencing-ht-10
#| check: true

exercise_results$sentencing_ht_10 <<- "fail"
gradethis::grade_this({
  if (.result == length(nonWhiteSentences)) {
    exercise_results$sentencing_ht_10 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

**Average sentence length — white offenders:** Input the average sentence length for white offenders below.

```{webr}
#| exercise: sentencing-ht-11

```

```{webr}
#| exercise: sentencing-ht-11
#| solution: true

mean(whiteSentences)
```

```{webr}
#| exercise: sentencing-ht-11
#| check: true

exercise_results$sentencing_ht_11 <<- "fail"
gradethis::grade_this({
  if (abs(.result - mean(whiteSentences)) < 0.01) {
    exercise_results$sentencing_ht_11 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

**Average sentence length — non-white offenders:** Input the average sentence length for non-white offenders below.

```{webr}
#| exercise: sentencing-ht-12


```

```{webr}
#| exercise: sentencing-ht-12
#| solution: true

mean(nonWhiteSentences)
```

```{webr}
#| exercise: sentencing-ht-12
#| check: true

exercise_results$sentencing_ht_12 <<- "fail"
gradethis::grade_this({
  if (abs(.result - mean(nonWhiteSentences)) < 0.01) {
    exercise_results$sentencing_ht_12 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

**Standard deviation in sentence lengths — white offenders:** Input the standard deviation in sentence lengths for white offenders below.

```{webr}
#| exercise: sentencing-ht-13


```

```{webr}
#| exercise: sentencing-ht-13
#| solution: true

sd(whiteSentences)
```

```{webr}
#| exercise: sentencing-ht-13
#| check: true

exercise_results$sentencing_ht_13 <<- "fail"
gradethis::grade_this({
  if (abs(.result - sd(whiteSentences)) < 0.01) {
    exercise_results$sentencing_ht_13 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

**Standard deviation in sentence lengths — non-white offenders:** Input the standard deviation in sentence lengths for white offenders below.

```{webr}
#| exercise: sentencing-ht-14


```

```{webr}
#| exercise: sentencing-ht-14
#| solution: true

sd(nonWhiteSentences)
```

```{webr}
#| exercise: sentencing-ht-14
#| check: true

exercise_results$sentencing_ht_14 <<- "fail"
gradethis::grade_this({
  if (abs(.result - sd(nonWhiteSentences)) < 0.01) {
    exercise_results$sentencing_ht_14 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

Now let's put the pieces together. Use your spreadsheet for all of the calculations below.

:::{.callout-caution}
## Check Your Understanding: Sentencing HT, Part X

The population parameter in question for this hypothesis test is:

```{ojs}
//| echo: false

viewof q16 = Inputs.radio(
  new Map([
    ["μ_white", 1],
    ["μ_nonwhite", 2],
    ["μ_white − μ_nonwhite", 3],
    ["x̄_white − x̄_nonwhite", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q16_selected_t15e") ?? "null")}
);

{
  localStorage.setItem("q16_selected_t15e", JSON.stringify(q16));
  localStorage.setItem("q16_correct_t15e", "3");
  localStorage.setItem("q16_result_t15e", q16 === null ? "unattempted" : (q16 == 3 ? "correct" : "incorrect"));
}
ok_response(q16, "3");
```

:::{.callout-note collapse="true"}
## Hint 1

The *population parameter* here is the difference in population means. 

:::

:::

:::{.callout-caution}
## Check Your Understanding: Sentencing HT, Part XI

The null value is:

```{ojs}
//| echo: false

viewof q17 = Inputs.radio(
  new Map([
    ["0", 1],
    ["0.5", 2],
    ["14", 3],
    ["44.79", 4],
    ["39.25", 5]
  ]),
  {value: JSON.parse(localStorage.getItem("q17_selected_t15e") ?? "null")}
);

{
  localStorage.setItem("q17_selected_t15e", JSON.stringify(q17));
  localStorage.setItem("q17_correct_t15e", "1");
  localStorage.setItem("q17_result_t15e", q17 === null ? "unattempted" : (q17 == 1 ? "correct" : "incorrect"));
}
ok_response(q17, "1");
```

:::{.callout-note collapse="true"}
## Hint 1

The null value is the *expected* value of the point estimate under the assumption that the null hypothesis ($H_0$) is true.

:::{.callout-note collapse="true"}
## Hint 2

Check the number to the right of the equal sign in the null hypothesis.

:::
:::

:::

**Point estimate** — compute in your spreadsheet and enter below:

```{webr}
#| exercise: sentencing-ht-16
# Enter your point estimate here

```

:::{.hint exercise="sentencing-ht-16"}
:::{.callout-note collapse="false"}
## Hint 1

The parameter is $\mu_{\text{white}} - \mu_{\text{non-white}}$, so the point estimate is the difference in sample means.

:::
:::

:::{.hint exercise="sentencing-ht-16"}
:::{.callout-note collapse="false"}
## Hint 2

The parameter is $\mu_{\text{white}} - \mu_{\text{non-white}}$, so the point estimate is the difference in sample means. Subtract the mean sentence for non-white offenders from the mean sentence for white offenders (or vice versa — just be consistent with the way you've written your hypotheses).

:::
:::

```{webr}
#| exercise: sentencing-ht-16
#| solution: true

mean(whiteSentences) - mean(nonWhiteSentences)
```

```{webr}
#| exercise: sentencing-ht-16
#| check: true

exercise_results$sentencing_ht_16 <<- "fail"
gradethis::grade_this({
  if (abs(abs(.result) - abs(mean(whiteSentences) - mean(nonWhiteSentences))) < 0.01) {
    exercise_results$sentencing_ht_16 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

**Standard error** — compute in your spreadsheet and enter below:

```{webr}
#| exercise: sentencing-ht-17
# Enter your standard error here

```

:::{.hint exercise="sentencing-ht-17"}
:::{.callout-note collapse="false"}
## Hint 1

The standard error formula is $S_E = \sqrt{s_1^2/n_1 + s_2^2/n_2}$. You have all four quantities in your *pivot table*. Use a blank cell in your spreadsheet for your calculation.

:::
:::

:::{.hint exercise="sentencing-ht-17"}
:::{.callout-note collapse="false"}
## Hint 2

In your spreadsheet, type `=SQRT((___^2/___)+(___^2/___))` substituting the appropriate values for the blanks. You can either type the values in (just be sure to keep at least four decimal places) or use cell references in the calculation.

:::
:::

:::{.hint exercise="sentencing-ht-17"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

In your spreadsheet, type `=SQRT((___^2/___)+(___^2/___))` substituting the appropriate values for the blanks. You can either type the values in (just be sure to keep at least four decimal places) or use cell references in the calculation.

+ Fill the first blank with the standard deviation for the sentences passed down to white offenders.
+ Fill the second blank with the number of sentences to white offenders in our sample data.
+ Fill the third blank with the standard deviation for the sentences passed down to non-white offenders.
+ Fill the final blank with the number of sentences to non-white offenders in our sample data.

:::
:::

```{webr}
#| exercise: sentencing-ht-17
#| solution: true

sqrt((sd(whiteSentences)^2 / length(whiteSentences)) + 
     (sd(nonWhiteSentences)^2 / length(nonWhiteSentences)))
```

```{webr}
#| exercise: sentencing-ht-17
#| check: true

exercise_results$sentencing_ht_17 <<- "fail"
gradethis::grade_this({
  se <- sqrt((sd(whiteSentences)^2 / length(whiteSentences)) + 
             (sd(nonWhiteSentences)^2 / length(nonWhiteSentences)))
  if (abs(.result - se) < 0.01) {
    exercise_results$sentencing_ht_17 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

**Test statistic** — compute in your spreadsheet and enter below:

```{webr}
#| exercise: sentencing-ht-18
# Enter your test statistic here

```

:::{.hint exercise="sentencing-ht-18"}
:::{.callout-note collapse="false"}
## Hint 1

The test statistic formula is $\text{test statistic} = \frac{(\text{point estimate}) - (\text{null value})}{S_E}$. You have all three components from the steps above. In your spreadsheet, type `=(___ - ___)/___` substituting your computed values.

:::
:::

:::{.hint exercise="sentencing-ht-18"}
:::{.callout-note collapse="false"}
## Hint 2

The test statistic formula is $\text{test statistic} = \frac{(\text{point estimate}) - (\text{null value})}{S_E}$. You have all three components from the steps above. In your spreadsheet, type `=(___ - 0)/___` substituting your computed values.

+ The second blank is the *null value*.

:::
:::

:::{.hint exercise="sentencing-ht-18"}
:::{.callout-note collapse="false"}
## Hint 3

The test statistic formula is $\text{test statistic} = \frac{(\text{point estimate}) - (\text{null value})}{S_E}$. You have all three components from the steps above. In your spreadsheet, type `=(-14.92 - 0)/___` substituting your computed values.

+ The second blank is the *null value*.
+ The first blank is the *point estimate*.

:::
:::

:::{.hint exercise="sentencing-ht-18"}
:::{.callout-note collapse="false"}
## Hint 4 (Solved)

The test statistic formula is $\text{test statistic} = \frac{(\text{point estimate}) - (\text{null value})}{S_E}$. You have all three components from the steps above. In your spreadsheet, type `=(-14.92 - 0)/6.99` substituting your computed values.

+ The second blank is the *null value*.
+ The first blank is the *point estimate*.
+ The final blank is the *standard error*.

:::
:::

```{webr}
#| exercise: sentencing-ht-18
#| solution: true

se <- sqrt((sd(whiteSentences)^2 / length(whiteSentences)) + 
           (sd(nonWhiteSentences)^2 / length(nonWhiteSentences)))

(mean(whiteSentences) - mean(nonWhiteSentences)) / se
```

```{webr}
#| exercise: sentencing-ht-18
#| check: true

exercise_results$sentencing_ht_18 <<- "fail"
gradethis::grade_this({
  if (abs(abs(.result) - abs(.solution)) < 0.01) {
    exercise_results$sentencing_ht_18 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

**Degrees of freedom** — compute in your spreadsheet and enter below:

```{webr}
#| exercise: sentencing-ht-18_5
# Enter your degrees of freedom here

```

:::{.hint exercise="sentencing-ht-18_5"}
:::{.callout-note collapse="false"}
## Hint 1

For a two-sample $t$-test, the degrees of freedom is $\text{df} = \min(n_1 - 1, n_2 - 1)$ — one less than the smaller of the two sample sizes. Which group had fewer observations?

:::
:::

:::{.hint exercise="sentencing-ht-18_5"}
:::{.callout-note collapse="false"}
## Hint 2

The white offenders group had only 37 observations while there were 243 observations for non-white offenders.

:::
:::

:::{.hint exercise="sentencing-ht-18_5"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

The white offenders group had only 37 observations while there were 243 observations for non-white offenders.

The *degrees of freedom* is one less than the smaller of the two sample sizes. That's 36 here.

:::
:::

```{webr}
#| exercise: sentencing-ht-18_5
#| solution: true

min(length(whiteSentences), length(nonWhiteSentences)) - 1
```

```{webr}
#| exercise: sentencing-ht-18_5
#| check: true

exercise_results$sentencing_ht_18_5 <<- "fail"
gradethis::grade_this({
  df <- min(length(whiteSentences), length(nonWhiteSentences)) - 1
  if (.result == df) {
    exercise_results$sentencing_ht_18_5 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

**$p$-value** — compute in your spreadsheet and enter below:

```{webr}
#| exercise: sentencing-ht-19
# Enter your p-value here

```

:::{.hint exercise="sentencing-ht-19"}
:::{.callout-note collapse="false"}
## Hint 1

Draw a picture of the $t$-distribution with your test statistic marked. The alternative hypothesis is two-sided ($\neq$), so both tails contribute to the $p$-value. Since both tail areas are equal, compute one tail area and double it.

:::
:::

:::{.hint exercise="sentencing-ht-19"}
:::{.callout-note collapse="false"}
## Hint 2

The alternative hypothesis is two-sided ($\neq$), so both tails contribute to the $p$-value. Since both tail areas are equal, compute one tail area and double it.

:::
:::

:::{.hint exercise="sentencing-ht-19"}
:::{.callout-note collapse="false"}
## Hint 3

You can use the `T.DIST()` function with the appropriate boundary value and degrees of freedom to calculate the tail area.

:::
:::

:::{.hint exercise="sentencing-ht-19"}
:::{.callout-note collapse="false"}
## Hint 4

Your tail area is the area away from the center of the distribution. Depending on whether your test statistic is positive or negative, you may need to utilize `1 - T.DIST()`.

:::
:::

:::{.hint exercise="sentencing-ht-19"}
:::{.callout-note collapse="false"}
## Hint 4 (Solved)

+ If your test statistic is negative, then use `=2*T.DIST(___, ___, TRUE)` with the first blank filled in by your test statistic and the second by the degrees of freedom to calculate your $p$-value.
+ If your test statistic is positive, then use `=2*(1 - T.DIST(___, ___, TRUE))` with the first blank filled in by your test statistic and the second by the degrees of freedom to calculate your $p$-value.

:::
:::

```{webr}
#| exercise: sentencing-ht-19
#| solution: true

se <- sqrt((sd(whiteSentences)^2 / length(whiteSentences)) + 
           (sd(nonWhiteSentences)^2 / length(nonWhiteSentences)))
ts <- (mean(whiteSentences) - mean(nonWhiteSentences)) / se
df <- min(length(whiteSentences), length(nonWhiteSentences)) - 1

2 * (1 - pt(abs(ts), df = df))
```

```{webr}
#| exercise: sentencing-ht-19
#| check: true

exercise_results$sentencing_ht_19 <<- "fail"
gradethis::grade_this({
  se <- sqrt((sd(whiteSentences)^2 / length(whiteSentences)) + 
             (sd(nonWhiteSentences)^2 / length(nonWhiteSentences)))
  ts <- (mean(whiteSentences) - mean(nonWhiteSentences)) / se
  df <- min(length(whiteSentences), length(nonWhiteSentences)) - 1
  pval <- 2 * (1 - pt(abs(ts), df = df))
  if (abs(.result - pval) < 0.01) {
    exercise_results$sentencing_ht_19 <<- "pass"
    pass(random_praise())
  }
  fail(random_encouragement())
})
```

:::{.callout-caution}
## Check Your Understanding: Sentencing HT XII

What is the result of the test?

```{ojs}
//| echo: false

viewof q18 = Inputs.radio(
  new Map([
    ["Since p ≥ α, we do not have enough evidence to reject the null hypothesis.", 1],
    ["Since p ≥ α, we accept the null hypothesis.", 2],
    ["Since p < α, we reject the null hypothesis and accept the alternative hypothesis.", 3],
    ["Since p < α, we fail to reject the null hypothesis.", 4],
    ["It is impossible to determine.", 5]
  ]),
  {value: JSON.parse(localStorage.getItem("q18_selected_t15e") ?? "null")}
);

{
  localStorage.setItem("q18_selected_t15e", JSON.stringify(q18));
  localStorage.setItem("q18_correct_t15e", "3");
  localStorage.setItem("q18_result_t15e", q18 === null ? "unattempted" : (q18 == 3 ? "correct" : "incorrect"));
}
ok_response(q18, "3");
```

:::{.callout-note collapse="true"}
## Hint 1

Compare your computed $p$-value to $\alpha = 0.10$. Is the $p$-value smaller than $\alpha$, or larger?

:::{.callout-note collapse="true"}
## Hint 2

If the $p$-value is at least as large as $\alpha$, our observed data would not be very surprising under the null hypothesis. If the $p$-value is less than $\alpha$ then the observed data ***is*** surprising. 

:::{.callout-note collapse="true"}
## Hint 3

If our data is surprising under the null hypothesis, then there are two options. First, maybe we just happened to observe an extreme sample by chance. Second, our data wasn't generated within a reality in which the null hypothesis ($H_0$) is true.

:::{.callout-note collapse="true"}
## Hint 4 (Solved)

Because we observed a sample which was unlikely in the "null reality", we reject the null hypothesis in favor of the alternative.

:::
:::
:::
:::

:::

:::{.callout-caution}
## Check Your Understanding: Sentencing HT XIII

The result of the test means that:

```{ojs}
//| echo: false

viewof q19 = Inputs.radio(
  new Map([
    ["The sample data did not provide significant evidence to suggest a difference in average sentence length for white versus non-white defendants.", 1],
    ["The sample data proved there is no difference in average sentence length for white versus non-white defendants.", 2],
    ["The sample data provided significant evidence to suggest a difference in average sentence length for white versus non-white defendants for drug-related cases in the Southern District of New York State.", 3],
    ["Sample data cannot be used to test claims about a population.", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q19_selected_t15e") ?? "null")}
);

{
  localStorage.setItem("q19_selected_t15e", JSON.stringify(q19));
  localStorage.setItem("q19_correct_t15e", "3");
  localStorage.setItem("q19_result_t15e", q19 === null ? "unattempted" : (q19 == 3 ? "correct" : "incorrect"));
}
ok_response(q19, "3");
```

:::

:::{.callout-caution}
## Check Your Understanding: Sentencing HT XIV

Which of the following is an appropriate implication of our result?

```{ojs}
//| echo: false

viewof q20 = Inputs.radio(
  new Map([
    ["We have proved that racial bias exists in sentencing for drug-related cases in the Southern District of New York State.", 1],
    ["We have found evidence to support the hypothesis that racial bias exists in sentencing for drug-related cases in the SDNY. Our result suggests that a more formal audit of sentencing recommendations should be undertaken.", 2],
    ["We have found evidence to support the hypothesis that racial bias exists in the courts across all case types.", 3],
    ["We have found evidence to suggest that racial minorities receive more lengthy sentences in drug-related cases in the Southern District of New York State.", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q20_selected_t15e") ?? "null")}
);

{
  localStorage.setItem("q20_selected_t15e", JSON.stringify(q20));
  localStorage.setItem("q20_correct_t15e", "2");
  localStorage.setItem("q20_result_t15e", q20 === null ? "unattempted" : (q20 == 2 ? "correct" : "incorrect"));
}
ok_response(q20, "2");
```

:::

## Submit

If you are part of a course with an instructor who is grading your work on these activities, please copy and submit ***both*** of the hashes below using the method your instructor has requested.

:::{.callout-warning}
## Question Hash

The hash below encodes your responses to the multiple choice questions in this activity.

```{ojs}
//| echo: false

function buildQuestionResults() {
  return {
    notebook: "Topic 15: Hypothesis Tests and Confidence Intervals for Numerical Data (Spreadsheets)",
    type: "questions",
    timestamp: new Date().toISOString(),
    questions: {
      q1_sentencing_ci_inference_type: {
        selected: q1,
        correct_answer: "2",
        result: q1 === null ? "unattempted" : (q1 == 2 ? "correct" : "incorrect")
      },
      q2_sentencing_ci_se_formula: {
        selected: q2,
        correct_answer: "3",
        result: q2 === null ? "unattempted" : (q2 == 3 ? "correct" : "incorrect")
      },
      q3_sentencing_ci_distribution: {
        selected: q3,
        correct_answer: "2",
        result: q3 === null ? "unattempted" : (q3 == 2 ? "correct" : "incorrect")
      },
      q4_sentencing_ci_confidence_level: {
        selected: q4,
        correct_answer: "3",
        result: q4 === null ? "unattempted" : (q4 == 3 ? "correct" : "incorrect")
      },
      q5_sentencing_ci_interpretation: {
        selected: q5,
        correct_answer: "4",
        result: q5 === null ? "unattempted" : (q5 == 4 ? "correct" : "incorrect")
      },
      q6_sentencing_ci_exceeds_36_months: {
        selected: q6,
        correct_answer: "2",
        result: q6 === null ? "unattempted" : (q6 == 2 ? "correct" : "incorrect")
      },
      q7_sentencing_ht_inference_type: {
        selected: q7,
        correct_answer: "3",
        result: q7 === null ? "unattempted" : (q7 == 3 ? "correct" : "incorrect")
      },
      q8_sentencing_ht_significance_level: {
        selected: q8,
        correct_answer: "3",
        result: q8 === null ? "unattempted" : (q8 == 3 ? "correct" : "incorrect")
      },
      q9_sentencing_ht_parameter_type: {
        selected: q9,
        correct_answer: "1",
        result: q9 === null ? "unattempted" : (q9 == 1 ? "correct" : "incorrect")
      },
      q10_sentencing_ht_num_groups: {
        selected: q10,
        correct_answer: "2",
        result: q10 === null ? "unattempted" : (q10 == 2 ? "correct" : "incorrect")
      },
      q11_sentencing_ht_hypotheses: {
        selected: q11,
        correct_answer: "1",
        result: q11 === null ? "unattempted" : (q11 == 1 ? "correct" : "incorrect")
      },
      q12_sentencing_ht_pop_sd_known: {
        selected: q12,
        correct_answer: "3",
        result: q12 === null ? "unattempted" : (q12 == 3 ? "correct" : "incorrect")
      },
      q13_sentencing_ht_paired: {
        selected: q13,
        correct_answer: "1",
        result: q13 === null ? "unattempted" : (q13 == 1 ? "correct" : "incorrect")
      },
      q14_sentencing_ht_se_formula: {
        selected: q14,
        correct_answer: "2",
        result: q14 === null ? "unattempted" : (q14 == 2 ? "correct" : "incorrect")
      },
      q15_sentencing_ht_distribution: {
        selected: q15,
        correct_answer: "3",
        result: q15 === null ? "unattempted" : (q15 == 3 ? "correct" : "incorrect")
      },
      q16_sentencing_ht_parameter: {
        selected: q16,
        correct_answer: "3",
        result: q16 === null ? "unattempted" : (q16 == 3 ? "correct" : "incorrect")
      },
      q17_sentencing_ht_null_value: {
        selected: q17,
        correct_answer: "1",
        result: q17 === null ? "unattempted" : (q17 == 1 ? "correct" : "incorrect")
      },
      q18_sentencing_ht_test_result: {
        selected: q18,
        correct_answer: "3",
        result: q18 === null ? "unattempted" : (q18 == 3 ? "correct" : "incorrect")
      },
      q19_sentencing_ht_conclusion_context: {
        selected: q19,
        correct_answer: "3",
        result: q19 === null ? "unattempted" : (q19 == 3 ? "correct" : "incorrect")
      },
      q20_sentencing_ht_implication: {
        selected: q20,
        correct_answer: "2",
        result: q20 === null ? "unattempted" : (q20 == 2 ? "correct" : "incorrect")
      }
    }
  };
}

function toBase64(str) {
  return btoa(unescape(encodeURIComponent(str)));
}

question_hash = {
  q1; q2; q3; q4; q5; q6; q7; q8; q9; q10;
  q11; q12; q13; q14; q15; q16; q17; q18; q19; q20;
  return toBase64(JSON.stringify(buildQuestionResults()));
}
```

```{ojs}
//| echo: false

html`<div style="font-family: monospace; font-size: 0.85em; background: #f5f5f5; padding: 12px; border-radius: 6px; word-break: break-all; border: 1px solid #ddd; user-select: all; cursor: pointer;" onclick="navigator.clipboard.writeText(this.innerText)">
  ${question_hash}
</div>
<p style="margin-top: 8px; font-size: 0.9em; color: #555;">
  Click the box to copy to clipboard.
</p>`
```

:::

:::{.callout-warning}
## Exercise Hash

Click the button below to generate your exercise submission code. This hash encodes your work on the graded exercises in this activity.

You must have attempted the graded exercises before clicking — clicking generates a snapshot of your current results. If you have completed the activity over multiple sessions, please go back through and enter your answer in each graded exercise cell before generating the hash below, to ensure your most recent results are recorded.

```{webr}
#| exercise: generate-exercise-hash
#| caption: "Generate Exercise Code"

exercise_results$timestamp <- format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
exercise_results$type <- "exercises"

hash <- base64encode(charToRaw(toJSON(exercise_results, auto_unbox = TRUE)))
cat(hash)
```

:::

## Summary

Great work through all of that — and I hope you found the application to sentencing data in the SDNY both interesting and thought-provoking. The problems were scaffolded step-by-step deliberately, to help you see how each individual piece connects to the larger process of conducting inference with numerical data.

:::{.callout-tip}
## Main Takeaways

+ **The $t$-distributions are a family of penalized normal distributions** parameterized by *degrees of freedom*. As degrees of freedom increase, the $t$-distribution becomes closer to the normal distribution.
+ **Use the $t$-distribution any time you use a sample standard deviation** ($s$) as a substitute for the population standard deviation ($\sigma$) in the standard error formula.
+ We have two functions for $t$-distribution work that you can use in your spreadsheets:
  + **`=T.DIST(q, df, TRUE)`** will return the area to the **left** of boundary value `q`. Continue drawing a picture and having the picture guide you through how to use `T.DIST()` to calculate the area you want.
  + **`=T.INV(p, df)`** provides the boundary value for which the area to the left is `p`. Use `=T.INV(0.975, df)` for the critical value of a 95% CI, for example.

+ **For a single mean**, the standard error is $S_E = s/\sqrt{n}$ and the degrees of freedom is $n - 1$.
+ **For comparing two independent means**, the standard error is $S_E = \sqrt{s_1^2/n_1 + s_2^2/n_2}$ and the degrees of freedom is $\min\{n_1 - 1, n_2 - 1\}$.
+ **The confidence interval and hypothesis test formulas are the same** as before — what changes is the standard error formula and the function used for the critical value or $p$-value.
+ **Statistical significance does not imply proof.** Our finding of a significant difference in sentence lengths is evidence that warrants further investigation — it does not prove racial bias, nor does it tell us the direction of the difference without examining the point estimate.
:::

:::{.callout-tip}
## Looking Ahead

You've now completed the core statistical inference toolkit — confidence intervals and hypothesis tests for proportions, differences in proportions, means, and differences in means. In the coming activities, you'll have the opportunity to practice applying these tools across a variety of mixed scenarios, building fluency in selecting the right method for each situation. The decision tree and general strategy documents will continue to be your primary guides.
:::
