---
title: "Topic 3: Descriptive Statistics for Numerical and Categorical Data"
format: 
  live-html:
    include-in-header:
      text: |
        <style>
        .btn-exercise-solution { display: none !important; }
        </style>
    webr:
      packages:
        - dplyr
        - gradethis
      cell-options: 
        persist: true
    live:
      show-solutions: true
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}
{{< include ./_extensions/r-wasm/live/_gradethis.qmd >}}

```{webr}
#| edit: false
#| output: false
#| define:
#|   - ok_response

library(htmltools)
ok_response <- function(response, n) {
  if (is.na(response)) div(HTML("You haven't answered yet."), style = "color: purple")
  else if (response == n) div(HTML("Correct ✓"), style = "color: green")
  else div(HTML("Incorrect ✗"), style = "color: red")
}
```

```{ojs}
//| echo: false
function ok_checkbox(response, n) {
  if (!response || response.length === 0) 
    return html`<span style="color:purple">You haven't answered yet.</span>`;
  if (response.toString() === n) 
    return html`<span style="color:green">Correct ✓</span>`;
  return html`<span style="color:red">Not Yet! ✗</span>`;
}
```

```{webr}
#| edit: false
#| output: false
#| define:
#|   - ex_mean_sample_one

ex_mean_sample_one <- "unattempted"
```

```{webr}
#| edit: false
#| output: false
#| define:
#|   - ex_mean_sample_three

ex_mean_sample_three <- "unattempted"
```

```{webr}
#| edit: false
#| output: false
#| define:
#|   - ex_my_data

ex_my_data <- "unattempted"
```

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(gradethis)
library(kableExtra)
library(patchwork)

set.seed(241)

diamonds <- diamonds
diamonds <- data.frame(diamonds)

diamonds$region <- sample(c(1,2,3,4,5), prob = c(0.11, .02, 0.65, 0.20, .02), nrow(diamonds), replace = TRUE)
diamonds$region <- as.factor(diamonds$region)

samples <- data.frame(
  Sample_One   = diamonds$carat[sample(1:nrow(diamonds), 8, replace = FALSE)],
  Sample_Two   = diamonds$carat[sample(1:nrow(diamonds), 8, replace = FALSE)],
  Sample_Three = c(diamonds$carat[sample(1:nrow(diamonds), 7, replace = FALSE)], 10)
)

cutSamples <- data.frame(
  Sample_One   = diamonds$cut[sample(1:nrow(diamonds), 8, replace = FALSE)],
  Sample_Two   = diamonds$cut[sample(1:nrow(diamonds), 8, replace = FALSE)],
  Sample_Three = diamonds$cut[sample(1:nrow(diamonds), 8, replace = FALSE)]
)
```

```{webr}
#| context: setup
#| echo: false
#| eval: true
#| message: false
#| warning: false

library(dplyr)
library(ggplot2)

set.seed(241)

diamonds <- diamonds
diamonds <- data.frame(diamonds)

diamonds$region <- sample(c(1,2,3,4,5), prob = c(0.11, .02, 0.65, 0.20, .02), nrow(diamonds), replace = TRUE)
diamonds$region <- as.factor(diamonds$region)

samples <- data.frame(
  Sample_One   = diamonds$carat[sample(1:nrow(diamonds), 8, replace = FALSE)],
  Sample_Two   = diamonds$carat[sample(1:nrow(diamonds), 8, replace = FALSE)],
  Sample_Three = c(diamonds$carat[sample(1:nrow(diamonds), 7, replace = FALSE)], 10)
)

cutSamples <- data.frame(
  Sample_One   = diamonds$cut[sample(1:nrow(diamonds), 8, replace = FALSE)],
  Sample_Two   = diamonds$cut[sample(1:nrow(diamonds), 8, replace = FALSE)],
  Sample_Three = diamonds$cut[sample(1:nrow(diamonds), 8, replace = FALSE)]
)
```

:::{.callout-note}
## About

This activity builds on what we learned in the Topic 1 and Topic 2 activities. We'll use R to analyze numerical and categorical data — computing and interpreting measures of central tendency and spread for numerical data, and frequency tables for categorical data.
:::

## Descriptive Statistics

**Workbook Objectives:**

+ Compute and discuss appropriate summaries for both numerical and categorical data.
+ Discuss the difference between mean and median, and between standard deviation and interquartile range. Identify when each is appropriate.
+ Compute summary statistics both by hand and with R.

**Important Reminders:**

+ A variable is *numerical* if a summary statistic such as the mean has a meaningful interpretation.
+ A variable is *categorical* if it serves to sort observations into different groups.
+ The unique values a variable takes are called the *levels* of that variable.

## Summarizing Numerical Data

### Measuring Central Tendency

**Measures of Central Tendency:** The mean and median both attempt to measure the *center* of a dataset.

+ The **mean** of a set of observations is the traditional average, denoted $\bar{x}$ (or $\mu$ for population data):
$$\bar{x} = \frac{\displaystyle{\sum_{i=1}^{n}{x_i}}}{n} = \frac{x_1 + x_2 + \cdots + x_n}{n}$$

+ The **median** is the middle value when observations are listed in ascending order. If there are two middle values, the median is their mean.

**Example:** The following are three random samples of carat values from the `diamonds` dataset.

```{r}
#| echo: false
#| eval: true

samples |>
  kable() |>
  kable_styling(bootstrap_options = c("hover", "striped"))
```

1. Use the code block below to compute the mean of `Sample_One`.

```{webr}
#| exercise: mean-sample-one
______
```

:::{.hint exercise="mean-sample-one"}
:::{.callout-note collapse="false"}
## Hint 1

Manually add up the carat values in `Sample_One` and divide by the number of diamonds in the sample.

```{r}
#| echo: true
#| eval: false

(___ + ___ + ___ + ___ + ___ + ___ + ___ + ___) / ___
```

:::
:::

:::{.hint exercise="mean-sample-one"}
:::{.callout-note collapse="false"}
## Hint 2

```{r}
#| echo: true
#| eval: false

(0.41 + 1.01 + 1.22 + 1.02 + 0.30 + 0.30 + 1.00 + 1.17) / 8
```

:::
:::

```{webr}
#| exercise: mean-sample-one
#| solution: true

(0.41 + 1.01 + 1.22 + 1.02 + 0.30 + 0.30 + 1.00 + 1.17) / 8
```

```{webr}
#| exercise: mean-sample-one
#| check: true
#| define:
#|   - ex_mean_sample_one

gradethis::grade_this({
  if (.result == .solution) {
    ex_mean_sample_one <<- "pass"
    pass(random_praise())
  }
  ex_mean_sample_one <<- "fail"
  fail(random_encouragement())
})
```

In R we can easily compute means and medians for our samples or for the entire dataset. The `$` operator accesses a column of a data frame, and R includes `mean()` and `median()` functions. Both methods below compute the mean of `Sample_Two`:

```{r}
#| echo: true
#| eval: true

mean(samples$Sample_Two)

samples |>
  summarize(avg_carat_wt = mean(Sample_Two))
```

While the second method involves more typing, it is much more readable. Getting comfortable with the pipe operator (`|>`) will allow you to chain data manipulation steps together.

2. Use the code block below to compute the mean of `Sample_Three`. You may use either method.

```{webr}
#| exercise: mean-sample-three
______
```

:::{.hint exercise="mean-sample-three"}
:::{.callout-note collapse="false"}
## Hint 1

Start with the code below and update it to compute the mean of `Sample_Three`.

```{r}
#| echo: true
#| eval: false

samples |>
  summarize(avg_carat_wt = mean(Sample_Two))
```

:::
:::

```{webr}
#| exercise: mean-sample-three
#| solution: true

samples |>
  summarize(avg_carat_wt = mean(Sample_Three))
```

```{webr}
#| exercise: mean-sample-three
#| check: true
#| define:
#|   - ex_mean_sample_three

gradethis::grade_this({
  if (.result == .solution) {
    ex_mean_sample_three <<- "pass"
    pass(random_praise())
  }
  ex_mean_sample_three <<- "fail"
  fail(random_encouragement())
})
```

3. Use the `median()` function and the code block below to compute the median of each sample, then answer the question that follows. *(This block is for exploration — it is not graded.)*

```{webr}
#| exercise: find-medians
______
```

:::{.hint exercise="find-medians"}
:::{.callout-note collapse="false"}
## Hint 1

Start with the copied code and update it to compute the median instead of the mean.

```{r}
#| echo: true
#| eval: false

samples |>
  summarize(avg_carat_wt = mean(Sample_Two))
```

:::
:::

:::{.hint exercise="find-medians"}
:::{.callout-note collapse="false"}
## Hint 2

The `summarize()` function can compute multiple summary statistics at once.

```{r}
#| echo: true
#| eval: false

samples |>
  summarize(
    avg_carat_wt_samp1 = mean(Sample_One),
    avg_carat_wt_samp2 = mean(Sample_Two),
    avg_carat_wt_samp3 = mean(Sample_Three)
  )
```

:::
:::

:::{.callout-caution}
## Check Your Understanding: Measures of Center

What does your work above tell you about the mean and median as measures of central tendency?

```{ojs}
//| echo: false

mutable ok_response = (response, n) => { return html`Loading...` };
viewof q1 = Inputs.radio(
  new Map([
    ["The mean is generally larger than the median.", 1],
    ["The mean is generally smaller than the median.", 2],
    ["The mean is usually close to the median.", 3],
    ["The mean is more strongly distorted by outliers than the median is.", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q1") ?? "null")}
);

localStorage.setItem("q1", JSON.stringify(q1));
ok_response(q1, "4");
```

:::

4. In our first activity we saw that sample data can be used to make generalizations about populations. Answer the following questions with this in mind.

:::{.callout-caution}
## Check Your Understanding: Sampling and Validity I

Assuming the samples of diamonds were taken randomly, which of the following claims is justified?

```{ojs}
//| echo: false

viewof q2 = Inputs.radio(
  new Map([
    ["The mean from Sample_Three is too high.", 1],
    ["The mean from Sample_One is too low.", 2],
    ["Each of the sample means provides an estimate for the population mean.", 3],
    ["The population mean is equal to the sample means.", 4]
  ]),
  {value: JSON.parse(localStorage.getItem("q2") ?? "null")}
);

localStorage.setItem("q2", JSON.stringify(q2));
ok_response(q2, "3");
```

:::

:::{.callout-caution}
## Check Your Understanding: Sampling and Validity II

Which of the following would result in improved estimates for the true mean carat weight of diamonds in our population? Select all that apply.

```{ojs}
//| echo: false

//mutable ok_checkbox = (response, n) => { return html`Loading...` };
viewof q3 = Inputs.checkbox(
  new Map([
    ["Take the mean of the sample means.", 1],
    ["Combine the three samples into a single sample of 24 diamonds and take its mean.", 2],
    ["Take samples using smaller sample sizes.", 3],
    ["Stratify the diamonds according to price and take a random sample of each price category.", 4],
    ["Take samples including more diamonds.", 5]
  ]),
  {value: JSON.parse(localStorage.getItem("q3") ?? "[]")}
);

localStorage.setItem("q3", JSON.stringify(q3));
ok_checkbox(q3.toString(), "1,2,5");
```

:::

### Aside: Defining Your Own Data

For data not already stored in a data frame, we can use the `c()` function to create a list of values. Consider the distributions of doors knocked on by two political campaign workers last week (Monday–Friday):

$$\text{Worker A}: 23,~24,~25,~26,~27 \qquad \text{Worker B}: 0,~15,~25,~35,~50$$

The following code block finds the mean and median for Worker A. Run it, then add two lines to also find the mean and median for Worker B.

```{webr}
#| exercise: my-data

mean(c(23, 24, 25, 26, 27))
median(c(23, 24, 25, 26, 27))
```

:::{.hint exercise="my-data"}
:::{.callout-note collapse="false"}
## Hint 1

Copy and paste the two existing lines and replace the data inside `c()` to reflect Worker B.

:::
:::

:::{.hint exercise="my-data"}
:::{.callout-note collapse="false"}
## Hint 2

```{r}
#| echo: true
#| eval: false

mean(c(23, 24, 25, 26, 27))
median(c(23, 24, 25, 26, 27))
mean(c(___, ___, ___, ___, ___))
median(c(___, ___, ___, ___, ___))
```

:::
:::

```{webr}
#| exercise: my-data
#| solution: true

mean(c(23, 24, 25, 26, 27))
median(c(23, 24, 25, 26, 27))
mean(c(0, 15, 25, 35, 50))
median(c(0, 15, 25, 35, 50))
```

```{webr}
#| exercise: my-data
#| check: true
#| define:
#|   - ex_my_data

gradethis::grade_this_code({
  if (is_pass()) {
    ex_my_data <<- "pass"
  } else {
    ex_my_data <<- "fail"
  }
})
```

5. Use your explorations of means and medians for the campaign workers to answer the following.

:::{.callout-caution}
## Check Your Understanding: Measures of Center

Which of the following are correct? Select all that apply.

```{ojs}
//| echo: false

viewof q4 = Inputs.checkbox(
  new Map([
    ["Worker A and Worker B had the same mean and median number of doors knocked.", 1],
    ["Worker A and Worker B had identical door-knocking strategies.", 2],
    ["Measures of center alone are not enough to describe numerical data.", 3]
  ]),
  {value: JSON.parse(localStorage.getItem("q4") ?? "[]")}
);

localStorage.setItem("q4", JSON.stringify(q4));
ok_checkbox(q4.toString(), "1,3");
```

:::

## Measuring Spread

The *standard deviation* of a set of observations is denoted $s$ (or $\sigma$ for population data):
$$s = \sqrt{\frac{\displaystyle{\sum_{i=1}^{n}{\left(x_i - \bar{x}\right)^2}}}{n-1}}$$

The *interquartile range* (IQR) measures the spread of the middle 50% of observations:
$$IQR = Q_3 - Q_1$$

where $Q_1$ is the 25th percentile and $Q_3$ is the 75th percentile.

6. Check your intuition about these measures.

:::{.callout-caution}
## Check Your Understanding: Measures of Spread I

Without carrying out the computations, which sample of diamonds has the largest standard deviation?

```{ojs}
//| echo: false

viewof q5 = Inputs.radio(
  new Map([
    ["Sample One", 1],
    ["Sample Two", 2],
    ["Sample Three", 3]
  ]),
  {value: JSON.parse(localStorage.getItem("q5") ?? "null")}
);

localStorage.setItem("q5", JSON.stringify(q5));
ok_response(q5, "3");
```

:::

:::{.callout-caution}
## Check Your Understanding: Measures of Spread II

Which measure of spread is more greatly impacted by the presence of outliers?

```{ojs}
//| echo: false

viewof q6 = Inputs.radio(
  new Map([
    ["Standard Deviation", 1],
    ["Interquartile Range", 2]
  ]),
  {value: JSON.parse(localStorage.getItem("q6") ?? "null")}
);

localStorage.setItem("q6", JSON.stringify(q6));
ok_response(q6, "1");
```

:::

The two plots below show the distribution of `carat` weights for the diamonds in our population — a histogram (left) and a boxplot (right). The purple line marks the mean.

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false

library(patchwork)

p1 <- ggplot(data = diamonds) +
  geom_histogram(aes(x = carat)) +
  geom_vline(xintercept = mean(diamonds$carat), color = "purple", lwd = 1.75) +
  theme_bw()

p2 <- ggplot(data = diamonds) +
  geom_boxplot(aes(y = carat)) +
  geom_hline(yintercept = mean(diamonds$carat), color = "purple", lwd = 1.75, alpha = 0.5) +
  coord_flip() +
  theme_bw()

p1 + p2
```

The code block below computes standard deviations and IQRs for each sample. Run it and explore the output. *(This block is for exploration — it is not graded.)*

```{webr}
#| exercise: compute-sd-iqr

print("Standard Deviations:")
samples |>
  summarize_all(sd)

print("IQRs:")
samples |>
  summarize_all(IQR)

print("Quartile I (25th Percentile):")
samples |>
  summarize_all(quantile, probs = 0.25)

print("Quartile III (75th Percentile):")
samples |>
  summarize_all(quantile, probs = 0.75)
```

Use the code block below to find the standard deviation and IQR for each campaign worker. *(This block is for exploration — it is not graded.)*

```{webr}
#| exercise: sd-iqr-campaign
______
```

:::{.hint exercise="sd-iqr-campaign"}
:::{.callout-note collapse="false"}
## Hint 1

Use the `c()` function to define your door-knocking data, then apply `sd()` and `IQR()`.

:::
:::

:::{.hint exercise="sd-iqr-campaign"}
:::{.callout-note collapse="false"}
## Hint 2

Storing your data in variables first will save some typing.

```{r}
#| echo: true
#| eval: false

worker_a <- c(___)
worker_b <- c(___)
```

:::
:::

:::{.hint exercise="sd-iqr-campaign"}
:::{.callout-note collapse="false"}
## Hint 3 (Solved)

```{r}
#| echo: true
#| eval: false

worker_a <- c(23, 24, 25, 26, 27)
worker_b <- c(0, 15, 25, 35, 50)

sd(worker_a)
IQR(worker_a)
sd(worker_b)
IQR(worker_b)
```

:::
:::

## Summarizing Categorical Data

Categorical data is best summarized using a frequency table. Consider the samples of diamond cuts below:

```{r}
#| echo: false
#| eval: true

cutSamples |>
  kable() |>
  kable_styling(bootstrap_options = c("hover", "striped"))
```

We can pipe our data frame into `count()` to construct a frequency table, and use `mutate()` to add a relative frequency column. The code block below is preset for `Sample_One` — adapt it to summarize `Sample_Two` and `Sample_Three`. *(This block is for exploration — it is not graded.)*

```{webr}
#| exercise: summarize-cut

cutSamples |>
  count(Sample_One)

cutSamples |>
  count(Sample_One) |>
  mutate(relative_frequency = n / sum(n))
```

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false

library(patchwork)

p1 <- ggplot(data = cutSamples) +
  geom_bar(aes(x = Sample_Three)) +
  labs(title = "Cut type in Diamonds (sample)", x = "Cut Type") +
  theme_bw()

p2 <- ggplot(data = diamonds) +
  geom_bar(aes(x = cut)) +
  labs(title = "Cut type in Diamonds (population)", x = "Cut Type") +
  theme_bw()

p1 + p2
```

## Submit

```{ojs}
//| echo: false

// ── collect OJS question responses from localStorage ──────────────────────
function getQ(key) {
  const raw = localStorage.getItem(key);
  if (raw === null) return null;
  try { return JSON.parse(raw); } catch { return null; }
}

// ── collect exercise results from webR→OJS bridge ─────────────────────────
// These are defined by the check chunks above via define:
// ex_mean_sample_one, ex_mean_sample_three, ex_my_data

// ── assemble the results object ───────────────────────────────────────────
function buildResults() {
  return {
    notebook: "Topic 3: Descriptive Statistics",
    timestamp: new Date().toISOString(),
    questions: {
      q1_measures_of_center:        getQ("q1"),
      q2_sampling_validity_1:       getQ("q2"),
      q3_sampling_validity_2:       getQ("q3"),
      q4_measures_of_center_2:      getQ("q4"),
      q5_measures_of_spread_1:      getQ("q5"),
      q6_measures_of_spread_2:      getQ("q6")
    },
    exercises: {
      mean_sample_one:   ex_mean_sample_one,
      mean_sample_three: ex_mean_sample_three,
      my_data:           ex_my_data
    }
  };
}

// ── base64 encode ─────────────────────────────────────────────────────────
function toBase64(str) {
  return btoa(unescape(encodeURIComponent(str)));
}

// ── reactive: regenerate hash whenever any input changes ──────────────────
submission_hash = {
  // touch all reactive dependencies so this reruns when they change
  q1; q2; q3; q4; q5; q6;
  ex_mean_sample_one; ex_mean_sample_three; ex_my_data;
  return toBase64(JSON.stringify(buildResults()));
}
```

:::{.callout-warning}
## Submit Your Work

When you have completed the activity, copy the submission code below and paste it into the course submission form provided by your instructor.

```{ojs}
//| echo: false

html`<div style="font-family: monospace; background: #f5f5f5; padding: 12px; border-radius: 6px; word-break: break-all; border: 1px solid #ddd; user-select: all;">
  ${submission_hash}
</div>
<p style="margin-top: 8px; font-size: 0.9em; color: #555;">
  Click anywhere in the box above to select all, then copy.
</p>`
```

:::

## Summary

+ We can summarize numerical data using measures of central tendency and measures of spread.
+ The **mean** ($\bar{x}$ for samples, $\mu$ for populations) and **median** measure the center of numerical data.
+ The **standard deviation** ($s$ for samples, $\sigma$ for populations) and **IQR** measure the spread of numerical data.
+ The median and IQR are **robust** measures in the presence of outliers.
+ Categorical data is best summarized in a **frequency table** or **relative frequency table**.
